### 2022-09-27 manuscript.tex
initial commit

\documentclass[11pt]{article}

\usepackage[style=prsb,natbib=true,uniquename=false,uniquelist=false,sortcites=false]{biblatex}
\addbibresource{refs.bib}

\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{array}
\usepackage[letterpaper]{geometry}
\geometry{hmargin={1in,1in},vmargin={0.7in,0.7in}}

\usepackage[hidelinks]{hyperref}
\usepackage[inline]{enumitem}
\newlist{ilnum}{enumerate*}{1}
\setlist[ilnum]{ label=(\Roman*) }

\usepackage{amsmath}
\usepackage{amsthm,amssymb}
\usepackage{mleftright}
\mleftright

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% XeLaTeX font settings
\usepackage{fontspec}
\defaultfontfeatures{Ligatures=TeX}
\usepackage{unicode-math}

\setmainfont{Minion Pro}
\setsansfont[Scale=MatchLowercase]{Myriad Pro}
\setmonofont[Scale=MatchLowercase]{Fira Code}
\setmathfont[Scale=MatchLowercase]{TeX Gyre Pagella Math}
\setmathfont[range=up/{num,latin,Latin,greek,Greek}]{Minion Pro}
\setmathfont[range=it/{latin,Latin,greek,Greek}]{Minion Pro It}
\setmathfont[range=bfup/{latin,Latin,greek,Greek}]{Minion Pro Bold}
\setmathfont[range=bfit/{latin,Latin,greek,Greek}]{Minion Pro Bold It}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[explicit]{titlesec}
\titleformat{\section}        {\large \bfseries \scshape}{\thesection.}{1ex}{#1}
\titleformat{\subsection}     {\bfseries \itshape}{\thesubsection.}{1ex}{#1}
\titleformat{\subsubsection}  [runin]{\bfseries}{\thesubsubsection.}{1ex}{#1}

\titlespacing{\section}       {0pt}{1.5ex plus 0.5ex minus 0.5ex}{0.5ex plus 0.5ex minus 0.5ex}
\titlespacing{\subsection}    {0pt}{1.5ex plus 0.5ex minus 0.5ex}{0.0ex plus 0.25ex minus 0.25ex}
\titlespacing{\subsubsection} {0pt}{1.5ex plus 0.5ex minus 0.5ex}{1em}

\usepackage{setspace}
\usepackage{titling}
\usepackage{xparse}
\usepackage{lineno}

%%% new commands and macros
\newcommand{\der}{\mathop{}\!\mathrm{d}}
\newcommand{\Oh}[1]{O\left(#1\right)}
\newcommand{\ExpectationOperator}{\mathrm{E}}
\NewDocumentCommand \E {d[] d()} {%
  \IfNoValueTF {#1} {%
    \IfNoValueTF {#2} {%
      \mathop{\kern0pt\ExpectationOperator}\nolimits%
    }{%
      \mathop{\kern0pt\ExpectationOperator}\nolimits\left(#2\right)%
    }}{%
    \IfNoValueTF {#2} {%
      \mathop{\kern0pt\ExpectationOperator#1}\nolimits%
    }{%
      \mathop{\kern0pt\ExpectationOperator#1}\nolimits\left(#2\right)%
    }%
  }
}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}
\newcommand{\Fst}{F_{\mathrm{ST}}}
\newcommand{\mean}[1]{\overline{#1}}

\begin{document}

\begin{titlingpage}
\setlength{\droptitle}{2em}
\pretitle{\begin{center}\LARGE}
\posttitle{\par\end{center}\vskip 2.5em}

\title{\scshape Evolutionary dynamics, equilibrium selection, and what population and quantitative genetics have taught evolutionary game theory}
\author{Jeremy Van Cleve}
\date{}
\maketitle

\vfill

\noindent
Department of Biology\\
University of Kentucky\\
Lexington, KY 40506 USA\\[1em]
phone: 859-218-3020\\
fax: 859-257-1717\\[1em]
e-mail: \href{mailto:jvancleve@uky.edu}{jvancleve@uky.edu}

\vspace{2em}

\begin{flushright} \textit{Date modified: \today} \end{flushright}
\end{titlingpage}

\linenumbers
\onehalfspacing
\begin{abstract}

The application of game theory to animal behavior by Maynard Smith and Price invigorated evolutionary biology by suggesting that the complex dynamics among natural selection, genes, and behavior could be analyzed using an evolutionary refinement of the Nash equilibrium, namely the evolutionarily stable strategy (ESS)\@. As implied by ``equilibrium'' and ``stable'', an ESS is a predicted long-term outcome that an evolutionary process may reach via natural selection and mutation. ESS analysis has been particularly important in highlighting the importance of social dilemmas in biology where the evolution of cooperative behavior can be undermined by the individual fitness costs of that behavior. However, even simple games like the stag-hunt or coordination game can have multiple ESSs, which begs the question, which ESS is more likely to evolve? Unfortunately, ESS analysis does not reveal which ESS is more likely to evolve nor does it even show how genetic or demographic factors affect the speed of approach to an ESS. Progress on determining which ESSs are more ``realistic'' has been made using tools from stochastic processes and population and quantitative genetics including stochastic stability, absorption times, and G-matrix theory. In this paper, we will review the use of these tools and provide novel insight into how they can predict social evolution as a function of genetic and demographic processes.

\end{abstract}

\vspace{3em}
\noindent {\bfseries Key words}\\

\newpage

\section{Outline/Brainstorm}

\begin{enumerate}
  \item Intro
        \begin{enumerate}
          \item Little intro about Maynard Smith and Price
          \item Game theory as general optimization and search for fixed points
          \item Connect back to Hamilton 1964 and goal of optimization with relatedness
          \item Evolutionary process is dynamic but still has ``fixed points''
          \item Evolutionary change can be decomposed and each piece affects location and stability of fixed points
          \item Evol Game theory captures selection but not demography and genetic forces, which can have strong effects.
        \end{enumerate}
  \item Ok, so thinking more about this, the two main pieces in certain ways are the fitness or success measure and whether maximizing it leads a correct predictions. pop mean fitness as measure fails due to genetics and due to frequency dependence. hamilton suggested inclusive fitness. later authors showed that this is equivalent to an invasion fitness perspective. maximization of this leads to ESS. Stochasticity though due to mutation/drift/etc leads to stationary distribution...
  \item Extra things
        \begin{enumerate}
          \item emphasize the incompleteness of the review. Maybe cite a few ``relevant'' Perc papers?
        \end{enumerate}
\end{enumerate}

%%
How to introduce maximization here? Hamilton wanted a maximization function. Lewontin spelled out clearly some of the important issues involved in maximization. JMS started his papers with the idea that game theory could provide solutions that didn't need group selection (or even initially kin selection) that involved individual-level maximization.
%%
Yes, start with the JMS justification for need of ESS and his dismal of Lewontin: this brings up the conflict between group/individual/kin and the issue of maximization and what to maximize.
%%
maybe I need to think more about my outline to get moving more quickly. I have a good start with mentioning ESS, early refs, and two primary ``confusions'': namely that ESS and the levels and/or kin/group issues are related and that ESS can't handle other evolutionary forces. I use the way that maynard smith frames the ESS papers to motivate Q1; for Q2, I can use feldman/nowak papers to motivate maybe?

the flow for the intro should be something like: maynard smith setup this false dichotomy: ESS is about ``individual'' level selection when really there are two issues, ESS as tool of optimality and then the application of that tool to kin/group models. this requires explaining how ESS is a tool of optimality and how a particular optimality measure, namely lineage growth rate, accounts for kin and group selection since it can captures selection at all levels from the gene on up. accounting for other evolutionary forces is more difficult but one way forward uses the fixation probability and allows mutation/demography etc to affect fixation probabilities and hence transition rates.

\clearpage
\section{Introduction}

Although Richard Lewontin was the first to introduce game theory into biology in 1961 \cite{Lewontin:1961}, very few papers on the topic of game theory and biology were published in the following decade. It wasn't until 1973 when John Maynard Smith and George Price introduced the evolutionarily stable strategy (ESS) and applied it to the study of animal behavior \cite{Maynard-Smith:Price:1973} that biologists more widely came to appreciate the relevance and utility of game theoretic concepts and tools for questions in evolution biology and ecology. Specifically, Maynard Smith and Price posited that individual fitness could be viewed as an analog of the game-theoretic notion of ``utility'', which is the quantity that measures what agents optimize in pursuit of their objectives \cite{Myerson:1991}. Viewed this way, the ESS is a refinement of the famous Nash equilibrium \cite{Nash:1950}. Maynard Smith explained the ESS concept in more detail in an important paper in 1974 \cite{Maynard-Smith:1974}, introduced the famous ``Hawk-Dove'' game in his study of asymmetric games in 1976 \cite{Maynard-Smith:Parker:1976}, and summarized the nascent field of evolutionary game theory in his now classic 1982 book \cite{MaynardSmith:1982}. In the 50 years since the 1973 paper, evolutionary game theory has become an essential tool in evolutionary and behavioral ecology with rich theoretical work that delves into foundational evolutionary and game-theoretic concepts and mathematical and simulation models that predict ESSs for almost innumerable biological systems. Evolutionary game theory has also strongly influenced the social sciences including, naturally, economics, political science, psychology, and anthropology and has drawn applied mathematicians, computer scientists and physicists into mathematical biology.

The role of evolutionary game theory and the ESS method specifically in evolutionary biology has been at the crux of some of the most important conceptual debates in the field including the role of natural selection and adaptation vis-a-vis other forces \cite[e.g.,][]{MaynardSmith:1978,Gould:Lewontin:1979,Lewontin:1979,Orzack:Sober:1994,Gardner:2017,Kern:Hahn:2018,Jensen:Payseur:2019} and the importance of kin and group selection relative to individual selection
\cite[e.g.,][]{Maynard-Smith:1964,Hamilton:1963,Price:1972:cov,Wilson:Wilson:2007,Leigh:2010,Akcay:Van-Cleve:2012,West:Griffin:2007,Gardner:Grafen:2009,Nowak:Tarnita:2010,Abbot:Abe:2011,Allen:Nowak:2013,Birch:2014,Birch:2017,Nowak:McAvoy:2017}. These debates might be captured in part by the following two questions: \begin{ilnum} \item \label{q:I} How do ESS models that focus on individual fitness capture the effects of kin selection or group selection? \item \label{q:II} How does an ESS account for mutation, genetic drift, and other evolutionary forces? \end{ilnum} In this work, I will describe how conceptual advances in evolutionary theory since Maynard Smith and Price \cite{Maynard-Smith:Price:1973} have shed light on both of these questions and have led to a more integrative framework for understanding how multiple evolutionary forces in complex populations can lead to diverse phenotypes.

Question \ref{q:I} derives in part from how Maynard Smith introduced his work on ESSs; he argued that an ESS could provide an explanation for the evolution of behaviors where ``selection acts entirely at the individual level, but in which the success of any particular strategy depends on what strategies are adopted by other members of the population'' \cite[][p. 210]{Maynard-Smith:1974} and is not ``due to group or species selection'' \cite[p. 15]{Maynard-Smith:Price:1973} or selection on ``close relatives'' \cite[p. 210]{Maynard-Smith:1974} as might be the case for kin selection \cite{Hamilton:1964}. In setting up this dichotomy, Maynard Smith implies that the ESS method may not apply when kin or group selection are involved and that kin or group selection models might lead to different results when applied to the same biological scenario. As I will describe below, evolutionary theorists have since realized (including Maynard Smith himself \cite[p. 33]{MaynardSmith:1978}) that the ESS method is really orthogonal to issues of individual vs. group vs. kin selection and instead captures in what sense evolution via natural selection optimizes fitness given a specific measure of fitness. Issues of individual vs. group vs. kin selection are about the ``units'' or ``levels'' at which selection acts and how fitness should be measured to account for selection at those levels. Lewontin also hints at the levels of selection question in his 1961 game theory paper when he discusses the relative merits of individual-level measures of fitness like intrinsic or Malthusian growth rate $r$ versus population-level measures like mean fitness $\mean{w}$ \cite[p. 400-401]{Lewontin:1961} as analogs of utility. The levels of selection question became a major topic of study in evolutionary theory \cite[e.g.,][]{Lewontin:1970,Dawkins:1982,Wilson:Sober:1989,Maynard-Smith:Szathmary:1995,Wilson:1997,Michod:1999,Michod:2006,Szathmary:2015} and philosophy of biology \cite{Hull:1980,Brandon:1982,Damuth:Heisler:1988,Lloyd:1992,Lloyd:1994,Sober:Wilson:1994,Okasha:2006,Okasha:2016} and continues to generate substantial research \cite[e.g.,][]{Black:Bourrat:2020,Cooney:Mori:2022,Veit:2022}. For the purpose of explaining how an ESS can capture selection at multiple levels and among relatives or kin, I will argue below that the right measure of utility is the ``lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:Van-Cleve:2016,Lehmann:Mullon:2016} of a mutant allele at a single genetic locus, which can be shown to be functionally equivalent to a measure of inclusive fitness \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020}.

The origin of question \ref{q:II} rests in a different set of debates in evolutionary biology regarding the relative role of natural selection vis-a-vis other evolutionary forces, such as mutation, recombination, and gene flow, in explaining organismal phenotypes. Early in the 20th century, R. A. Fisher's ``fundamental theorem of natural selection'' (FTNS) \cite{Fisher:1930} established a mathematical expression of the importance of natural selection. The FTNS states that the increase in the mean fitness of a population is equal to the genetic variance in fitness and thus seems to imply that populations always become better adapted to their environments and even that fitness is maximized over the long term. It's hard to overstate the importance of the FTNS in shaping the direction of evolutionary theory. The FTNS came to typify the idea that evolutionary change is dominated by natural selection as a fitness optimizing force. Bill Hamilton appealed to this idea in his original paper on kin selection \cite{Hamilton:1964} and theorized that ``inclusive fitness'' is the fitness quantity that is maximized.

Subsequent work by population geneticists revealed however that mean fitness can decrease due to frequency-dependent selection \cite{Wright:1955,Lewontin:1970} or recombination among multiple genetic loci \cite{Kojima:Kelleher:1961,Moran:1964,Karlin:1975}. Further, studies in the 1960s of the rates of molecular evolution \cite{Zuckerkandl:Pauling:1965,King:Jukes:1969} and levels of polymorphism \cite{Harris:1966,Lewontin:Hubby:1966} in a number of species spurred the development of the neutral \cite{Kimura:1968,Kimura:1983:book} and nearly neutral \cite{Ohta:1974,Ohta:1992} theories \cite{Ohta:Gillespie:1996} of molecular evolution. These theories posited that many mutations are weakly affected by natural selection, and thus their fate is governed mostly by genetic drift \cite{Kimura:1983:book}. Consequently, by the 1970s, evolutionary biologists were heavily debating both the relative role of natural selection versus other forces in shaping evolutionary patterns \cite{Gillespie:Langley:1974,Gillespie:1978}, and some biologists including Lewontin specifically questioned the importance of fitness maximization \cite{Karlin:1975,Gould:Lewontin:1979}. Maynard Smith was a clear proponent of fitness maximization and viewed it natural to ``[assume] that evolution has occurred by natural selection'' \cite[p. 31]{MaynardSmith:1978}. Thus, he proposed the ESS method developed by Price and himself as the appropriate tool for predicting phenotypic evolution and particularly for social traits. While Maynard Smith acknowledged that ESS models make a number of biological assumptions including that traits have a simple genetic basis (i.e., no dominance, epistasis, etc) and that appropriate genetic variation exits \cite{MaynardSmith:1978}, he believed that the limitations imposed by these assumptions are reasonable. This view that simplifying away genetic complexities or constraints is generally reasonable was termed the ``phenotypic gambit'' by Alan Grafen \cite{Grafen:1984} and came to dominate theory development in behavioral and evolutionary ecology \cite{Houston:McNamara:1999}. As genomic data have become available for an increasing diverse set of species, some have argued that further advances in the evolution of behavior and other complex traits requires evolutionary theory that incorporates these data \cite[e.g.,][]{Springer:Crespi:2011,Rittschof:Robinson:2014,Akcay:Linksvayer:2015,Cunningham:2020}. Question \ref{q:II} arises here and asks whether the ESS method can be modified to accommodate other evolutionary forces such as mutation and recombination. I will argue below that the ESS method is more general than originally imagined by Maynard Smith and Price \cite{Maynard-Smith:Price:1973} in that in can in fact be incorporated into an evolutionary framework where natural selection and other evolutionary forces combine to shape the short and long-term evolution of traits.

\section{Acknowledgements}

\section{Funding}

\clearpage
%\setlength{\bibitemsep}{1pt}
\printbibliography

\end{document}

later asserted that since ``[i]t is assumed that evolution has occurred by natural selection'' \cite[p. 31]{MaynardSmith:1978},

the ESS method only identifies how natural selection shapes phenotypic traits to optimize fitness \cite[p. 42]{MaynardSmith:1978} and whereas population genetics can be applied to study ``selectively neutral'' molecular traits that are subject to mutation and genetic drift \cite[p. 37]{MaynardSmith:1978}.

This lead to / is similar to phenotypic gambit and idea of long-term evolution where genetics are inconsequential. Ah right but really maybe my argument is then (see below) that population genetic model that include many forces can be converted into long-term evolution models.

So the idea should be something like ESS is really a mathematical formulation of the long-term evolution process which can include additional evolutionary forces other than natural selection.

The argument here is that invasion fitness / lineage fitness and its finite population analogue, fixation probability, are really just measures of evolutionary success that can include any potential evolutionary force. The maxima of these measures then yield the stable fixed equilibria of a long-term evolutionary process in a deterministic model or the modes of probability density in a stochastic model. Can appeal to physical notions of mean-field (?), potential wells, large deviation theory. Also appeal to stochastic stability and equilibrium selection from game theory (which borrows from physics). ESS then represents attracting points in phenotypic space and is informative even in the non-zero mutation and finite population size cases.

In 1972, Price \cite{Price:1972:ftns} clarified the meaning the of the FTNS using his eponymous equation:

### 2022-10-02 manuscript.tex
added file and edits

@@
 %%
-How to introduce maximization here?
-Hamilton wanted a maximization function.
-Lewontin spelled out clearly some of the important issues involved in maximization.
-JMS started his papers with the idea that game theory could provide solutions that didn't need group selection (or even initially kin selection) that involved individual-level maximization.
-%%
-Yes, start with the JMS justification for need of ESS and his dismal of Lewontin: this brings up the conflict between group/individual/kin and the issue of maximization and what to maximize.
-%%
-maybe I need to think more about my outline to get moving more quickly.
-I have a good start with mentioning ESS, early refs, and two primary ``confusions'': namely that ESS and the levels and/or kin/group issues are related and that ESS can't handle other evolutionary forces.
-I use the way that maynard smith frames the ESS papers to motivate Q1; for Q2, I can use feldman/nowak papers to motivate maybe?
-
 the flow for the intro should be something like: maynard smith setup this false dichotomy: ESS is about ``individual'' level selection when really there are two issues, ESS as tool of optimality and then the application of that tool to kin/group models.
@@
 
-Subsequent work by population geneticists revealed however that mean fitness can decrease due to frequency-dependent selection \cite{Wright:1955,Lewontin:1970} or recombination among multiple genetic loci \cite{Kojima:Kelleher:1961,Moran:1964,Karlin:1975}.
+Subsequent work by population geneticists revealed however that mean fitness can decrease due to frequency-dependent selection \cite{Wright:1955,Lewontin:1970} or recombination among multiple genetic loci \cite{Kojima:Kelleher:1961,Moran:1964,Karlin:1975,Akin:1979}.
 Further, studies in the 1960s of the rates of molecular evolution \cite{Zuckerkandl:Pauling:1965,King:Jukes:1969} and levels of polymorphism \cite{Harris:1966,Lewontin:Hubby:1966} in a number of species spurred the development of the neutral \cite{Kimura:1968,Kimura:1983:book} and nearly neutral \cite{Ohta:1974,Ohta:1992} theories \cite{Ohta:Gillespie:1996} of molecular evolution.
@@
 While Maynard Smith acknowledged that ESS models make a number of biological assumptions including that traits have a simple genetic basis (i.e., no dominance, epistasis, etc) and that appropriate genetic variation exits \cite{MaynardSmith:1978}, he believed that the limitations imposed by these assumptions are reasonable.
-This view that simplifying away genetic complexities or constraints is generally reasonable was termed the ``phenotypic gambit'' by Alan Grafen \cite{Grafen:1984} and came to dominate theory development in behavioral and evolutionary ecology \cite{Houston:McNamara:1999}.
-As genomic data have become available for an increasing diverse set of species, some have argued that further advances in the evolution of behavior and other complex traits requires evolutionary theory that incorporates these data \cite[e.g.,][]{Springer:Crespi:2011,Rittschof:Robinson:2014,Akcay:Linksvayer:2015,Cunningham:2020}.
+This view that simplifying away genetic complexities or constraints is generally reasonable was termed the ``phenotypic gambit'' by Alan Grafen \cite{Grafen:1984} and came to dominate evolutionary theory in behavioral and evolutionary ecology \cite{Houston:McNamara:1999}.
+The near-singular focus of evolutionary game theory and the ESS method on natural selection has become less defensible given the recent and increasing abundance of genomic and transcriptomic data for a diverse set of species \cite[e.g.,][]{Kapheim:Pan:2015,Mikheyev:Linksvayer:2015,Warner:Mikheyev:2017,Kocher:Mallarino:2018,Warner:Mikheyev:2019}; specifically, some have argued that understanding these data requires moving beyond the gambit with more explicit consideration of complex genetic and demographic mechanisms \cite[e.g.,][]{Springer:Crespi:2011,Rittschof:Robinson:2014,Akcay:Linksvayer:2015,Cunningham:2020}.
 Question \ref{q:II} arises here and asks whether the ESS method can be modified to accommodate other evolutionary forces such as mutation and recombination.
 I will argue below that the ESS method is more general than originally imagined by Maynard Smith and Price \cite{Maynard-Smith:Price:1973} in that in can in fact be incorporated into an evolutionary framework where natural selection and other evolutionary forces combine to shape the short and long-term evolution of traits.
+
+\section{Long-term evolution}
+
+Following the development of the ESS approach by Maynard Smith and Price \cite{Maynard-Smith:Price:1973}, mathematical biologists evaluated the approach in light of the older and established theory population genetics.
+Population genetic models can explicitly address how both natural selection and genetic mechanisms including mutation, segregation, and recombination affect allele and genotype frequency dynamics over time.
+These models typically assume that evolution changes the frequencies of a fixed set of genotypes and the emergence of new genotypes through the large-effect mutations (i.e., the creation of new alleles or loci) occurs on a longer timescale than that of the model.
+For this reason, these population genetic models are often said to represent a ``short-term'' evolutionary process.
+Population genetic models showed that \cite{Cavalli-Sforza:Feldman:1978,Eshel:1982}
+
+Long-term evolution cites \cite{Eshel:1996,Hammerstein:1996,Weissing:1996}
+
+Though the idea of fitness optimization by itself has some intuitive appeal, high fitness individuals produce more surviving offspring who themselves are likely to be more fit than average, it needs an explicit evolutionary model in order to be evaluated mathematically and quantitatively.
 
@@
 
-later asserted that since ``[i]t is assumed that evolution has occurred by natural selection'' \cite[p. 31]{MaynardSmith:1978},
-
-the ESS method only identifies how natural selection shapes phenotypic traits to optimize fitness \cite[p.
-42]{MaynardSmith:1978} and whereas population genetics can be applied to study ``selectively neutral'' molecular traits that are subject to mutation and genetic drift \cite[p.
-37]{MaynardSmith:1978}.
-
-This lead to / is similar to phenotypic gambit and idea of long-term evolution where genetics are inconsequential.
-Ah right but really maybe my argument is then (see below) that population genetic model that include many forces can be converted into long-term evolution models.
+maybe my argument is then (see below) that population genetic model that include many forces can be converted into long-term evolution models.
 
@@
 ESS then represents attracting points in phenotypic space and is informative even in the non-zero mutation and finite population size cases.
-
-In 1972, Price \cite{Price:1972:ftns} clarified the meaning the of the FTNS using his eponymous equation:

### 2022-10-10 manuscript.tex
moar writing

@@
 \newcommand{\mean}[1]{\overline{#1}}
+\newcommand{\w}{w}
+\newcommand{\ess}[1]{#1^*}
 
@@
 \item \label{q:II} How does an ESS account for mutation, genetic drift, and other evolutionary forces?
-\end{ilnum} In this work, I will describe how conceptual advances in evolutionary theory since Maynard Smith and Price \cite{Maynard-Smith:Price:1973} have shed light on both of these questions and have led to a more integrative framework for understanding how multiple evolutionary forces in complex populations can lead to diverse phenotypes.
+\end{ilnum} In this work, I will describe how conceptual advances in evolutionary theory since Maynard Smith and Price \cite{Maynard-Smith:Price:1973} have shed light on both of these questions and have led to a broader understanding of the ESS method and a more integrative framework for understanding how multiple evolutionary forces in complex populations can lead to diverse phenotypes.
 
@@
 
+%%%
+%%% Detailed outline for post intro
+%%%
+
+What is ESS? ESS is just phenotypes and fitness and an equilibrium and stability condition. This matches pop gen for simple genetics \cite{Eshel:1982}.
+
+ESS really is a model of a ``long-term evolution'' process of invasion of mutant by resident where fitness determines invasion. This is the ``unbeatable strategy'' of Hamilton.
+
+Long-term evolution depends on a fitness ``measure'' so ``fitness'' isn't just individual fitness.
+Really, we track the frequency or density of individual alleles (i.e., gene's eye view).
+This is lineage growth rate.
+
+Lineage growth rate allows us to measure ESS in complex structure populations with kin and group selection.
+We measure the growth rate of the lineage in all the genetic and demographic environments it finds itself in.
+
+Genetic environments can include other loci linked by recombination. This is the case of the modifier theory for recombination, mutation, migration, etc. Example from Liberman 2011.
+
+Recognizing the importance of genetic parameters in ESS is the opposite of early work to justify ESS which tried to find conditions when ESS predicted same as pop gen but genetic parameters didn't appear.
+
+One way forward to understanding the complex phenotypic evolution including social evolution should involve using ESS theory that incorporates more genetics including multiple loci.
+This is notably complementary to methods that study trait coevolution, e.g., Mullon and Lehmann.
+
+%%%
+%%% Fleshed out outline material
+%%%
+
+\section{What is an ESS?}
+
+The definition of an ESS is relatively simple and only involves a measure of fitness, a set of phenotypes, and a stability condition.
+Let $\w(x,y)$ be the fitness that an individual with phenotype or strategy $x$ obtains when interacting with an individual with phenotype or strategy $y$.
+The phenotypes can be drawn from a set of discrete or continous set of values $X$ that represent a single trait like body size or propensity to cooperate.
+Fitness $\w$ for the moment is simply the survival rate (i.e., viability selection) but as we'll see we can define much more general fitness measures.
+The stability condition \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} says that a phenotype $\ess{x}$ is an ESS when no alternative strategy $y$ can improve upon the fitness that $\ess{x}$ receives when interacting with itself:
+\begin{subequations}
+  \label{eq:ess:def}
+\begin{equation}
+  \label{eq:ess:def:a}
+  \w(\ess{x}, \ess{x}) \ge \w(y, \ess{x})
+\end{equation}
+for all phenotypes $y$.
+In the case that the strategy $y$ gets the same fitness interacting with $\ess{x}$ as $\ess{x}$ does with itself (equation \eqref{eq:ess:def:a} holds as an equality) then $\ess{x}$ must do better interacting with $y$ than $y$ does with itself:
+\begin{equation}
+  \label{eq:ess:def:b}
+  \w(\ess{x}, y) > \w(y, y)
+\end{equation}
+\end{subequations}
+The two conditions in \eqref{eq:ess:def} are equivalent to requiring that phenotype $\ess{x}$ receives higher average fitness than any alternative phenotype $y$ when $y$ is rare in the population \cite{Maynard-Smith:1974,Bishop:Cannings:1976}.
+
+Notably, the conditions in \eqref{eq:ess:def} do not reference the genetic basis of the trait $X$ nor do they describe how the frequency or mean value of the trait changes over time.
+Population genetics models, on the other hand, are built to measure the dynamical process of evolution via the combined effects of natural selection and segregation, recombination, mutation, genetic drift, and other evolutionary forces \cite{Crow:Kimura:1970,Ewens:2004}.
+A natural question then is how the long-run behavior of a population genetic model (e.g., its equilibrium genotypes and induced phenotypes) compares to the equilibrium phenotypes obtained from an analogous ESS model.
+This question was tackled by mathematical population biologists beginning in the late 1970s and through 1980s and 1990s \cite[e..g,][]{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Eshel:1982,Hofbauer:Schuster:1982,Lessard:1984,Cressman:1988,Cressman:Hines:1984,Cressman:Hofbauer:1996,Hammerstein:1996,Weissing:1996,Eshel:1996,Eshel:Feldman:1984}.
+Early work in the simplest haploid asexual model with discrete strategies showed that the ESSs of obtained from condition \eqref{eq:ess:def} must be stable equilibria of haploid population genetic model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980}.
+This result isn't particularly surprising since natural selection is the only evolutionary force acting in the haploid asexual model.
+However, this idea was extended further by Ilan Eshel and Sabin Lessard
+
+%%%
+%%% old scraps and ramblings
+%%%
+
+\section{Stability in dynamical systems and the ESS}
+
+For the behavior or evolutionary ecologist interested in a mathematical model of natural selection, an ESS model offers one main advantage over population genetic models: the ESS model can be formulated entirely in terms of phenotypes and ftiness and can avoid relying explicitly on genetic parameters (e.g., number of loci affecting the trait, recombination rates, etc) because the implicit genetic assumptions (i.e., additive interactions between alleles within a locus and across loci) imply a population genetic model with the same assumptions would generate the same ESSs \cite{Eshel:1982}.
+
 \section{Long-term evolution}
@@
 Population genetic models can explicitly address how both natural selection and genetic mechanisms including mutation, segregation, and recombination affect allele and genotype frequency dynamics over time.
-These models typically assume that evolution changes the frequencies of a fixed set of genotypes and the emergence of new genotypes through the large-effect mutations (i.e., the creation of new alleles or loci) occurs on a longer timescale than that of the model.
+These models typically assume that evolution changes the frequencies of a fixed set of genotypes and that the emergence of new genotypes through the large-effect mutations (i.e., the creation of new alleles and loci) occurs on a longer timescale than that of the model.
 For this reason, these population genetic models are often said to represent a ``short-term'' evolutionary process.

### 2022-10-10 manuscript.tex
moar writing

@@
 This question was tackled by mathematical population biologists beginning in the late 1970s and through 1980s and 1990s \cite[e..g,][]{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Eshel:1982,Hofbauer:Schuster:1982,Lessard:1984,Cressman:1988,Cressman:Hines:1984,Cressman:Hofbauer:1996,Hammerstein:1996,Weissing:1996,Eshel:1996,Eshel:Feldman:1984}.
-Early work in the simplest haploid asexual model with discrete strategies showed that the ESSs of obtained from condition \eqref{eq:ess:def} must be stable equilibria of haploid population genetic model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980}.
+Early work in the simplest haploid asexual population genetic model with discrete strategies showed that the ESSs of obtained from condition \eqref{eq:ess:def} must be stable equilibria of the haploid asexual model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980}.
+Specifically, suppose there are two possible phenotypes $1$ and $2$ with the payoff matrix $A = {a_{xy}}$ where $a_{xy} = \w(x,y)$ is the payoff or increase to fitness of an individual of type $x$ interacting with an individual of type $y$.
+Then, applying the ESS condition \eqref{eq:ess:def} to the payoff matrix $A$ gives us the stable equilibria of the haploid population genetic model.
 This result isn't particularly surprising since natural selection is the only evolutionary force acting in the haploid asexual model.
-However, this idea was extended further by Ilan Eshel and Sabin Lessard
+However, this correspondence was extended to a diploid sexual population with two possible phenotypes and a single genetic locus where dominance is possible between homologous alleles and Mendelian segregation breaks up diploid genotypes every generation \cite{Eshel:1982,Lessard:1984}.
+Specifically, suppose there are $k$ possible alleles $A_{1},\ldots,A_{k}$ where diploid genotype $A_{i} A_{j}$ chooses phenotype $1$ with probability $v_{ij}$ and phenotype $2$ with probability $1-v_{ij}$.
+Eshel and Lessard show  \cite{Eshel:1982,Lessard:1984} that
 

### 2022-10-26 manuscript.tex
write write write

--- 2022-10-10 manuscript.tex
@@
 \newcommand{\ess}[1]{#1^*}
+\renewcommand{\vec}[1]{\mathbf{#1}}
 
@@
 
+\section{How does ESS related to population-genetic dynamics?}
+
 Notably, the conditions in \eqref{eq:ess:def} do not reference the genetic basis of the trait $X$ nor do they describe how the frequency or mean value of the trait changes over time.
@@
 This question was tackled by mathematical population biologists beginning in the late 1970s and through 1980s and 1990s \cite[e..g,][]{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Eshel:1982,Hofbauer:Schuster:1982,Lessard:1984,Cressman:1988,Cressman:Hines:1984,Cressman:Hofbauer:1996,Hammerstein:1996,Weissing:1996,Eshel:1996,Eshel:Feldman:1984}.
-Early work in the simplest haploid asexual model with discrete strategies showed that the ESSs of obtained from condition \eqref{eq:ess:def} must be stable equilibria of haploid population genetic model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980}.
+Early work in the simplest haploid asexual population genetic model with discrete strategies showed that the ESSs of obtained from condition \eqref{eq:ess:def} must be stable equilibria of the haploid asexual model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980}.
+Specifically, suppose there are two possible phenotypes $1$ and $2$ with the payoff matrix $A = \{a_{xy}\}$ where $a_{xy} = \w(x,y)$ is the payoff or increase to fitness of an individual of type $x$ interacting with an individual of type $y$.
+Then, the ESSs obtained by applying condition \eqref{eq:ess:def} to the payoff matrix $A$ are stable equilibria of a haploid asexual population genetic model with two genotypes, one for each phenotype \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Hofbauer:Sigmund:1998}.
 This result isn't particularly surprising since natural selection is the only evolutionary force acting in the haploid asexual model.
-However, this idea was extended further by Ilan Eshel and Sabin Lessard
+Importantly, a similar  correspondence between ESSs and population genetic models also applies to a diploid sexual population reproducing in discrete time.
+Suppose there are two possible phenotypes controlled by a single genetic locus with $k$ possible alleles $A_{1},\ldots,A_{k}$ where dominance is possible between alleles and Mendelian segregation breaks up diploid genotypes every generation \cite{Eshel:1982,Lessard:1984}.
+Let diploid genotype $A_{i} A_{j}$ choose phenotype $1$ with probability $v_{ij}$ and phenotype $2$ with probability $1-v_{ij}$.
+If the $\vec{p} = (p_{1}, \ldots, p_{k})$ is the vector of allele frequencies for alleles $A_{1},\ldots,A_{k}$, then the mean frequency of phenotype $1$ in the population (in the next generation) is
+\begin{equation*}
+  z(\vec{p}) = \sum_{i=1}^{k} p_{i} p_{j} v_{ij} \: .
+\end{equation*}
+Eshel and Lessard showed \cite{Eshel:1982,Lessard:1984} that ESSs $\ess{x}$ of the payoff matrix $A$ are stably maintained in the diploid population model when those ESSs are produced by some vector of allele frequencies $\ess{\vec{p}}$ such that $\ess{x} = z(\ess{\vec{p}})$.
+This correspondence also holds for continuous-time diploid models \cite{Sigmund:1987,Cressman:1988,Cressman:Hofbauer:1996,Hofbauer:Sigmund:1998} and for models with multiple loci when the probability of a genotype choosing a phenotype is an additive function of effects from each locus (i.e., additive non-epistasis) \cite{Eshel:1982}.
+
+\section{ESS and long-term evolution}
+
+An important limitation of the above results connecting ESS phenotypes $\ess{x}$ to equilibria in diploid single and multilocus population genetic models is that the genotype-phenotype map may not produce phenotypes that are ESS and thus those ESS phenotypes are unattainable in the short-term.
+Specifically, the $v_{ij}$ may have values so that no population composition of genotypes $A_{i} A_{j}$ yields the ESS phenotype $\ess{x}$.
+Moreover, as mentioned above, multilocus models more generally (i.e., genetic systems with non-additive interactions or epistasis) do not guarantee that genotypes with higher marginal fitness\footnote{The marginal fitness of a genotype or allele is the mean fitness of that genotype allele among all contexts in which it occurs in the population.
+For example, the marginal fitness of one allele at a single diploid locus in the mean fitness of that allele in combination with all other alleles according to their frequencies.} increase in frequency, which is a requirement for the correspondence between ESS and population genetic equilibria described above \cite{Eshel:1982,Hofbauer:Sigmund:1998}.
+Thus, the genetic system may impose constraints on either the range of possible phenotypes or on the ability of selection to increase the frequency of high marginal fitness genotypes.
+Intriguingly, Eshel and others discovered that these constraints might not be so important on a longer term evolutionary timescales.
+
+In addition to showing that ESS phenotypes $\ess{x}$ are
 
@@
 
+summary of results in chapter 22. for continuous time and two alleles or two phenotypes and linear payoffs, equilibria that produce ESS phenotypes are stable \cite{Hofbauer:Sigmund:1998}
+
 \section{Stability in dynamical systems and the ESS}
 
-For the behavior or evolutionary ecologist interested in a mathematical model of natural selection, an ESS model offers one main advantage over population genetic models: the ESS model can be formulated entirely in terms of phenotypes and ftiness and can avoid relying explicitly on genetic parameters (e.g., number of loci affecting the trait, recombination rates, etc) because the implicit genetic assumptions (i.e., additive interactions between alleles within a locus and across loci) imply a population genetic model with the same assumptions would generate the same ESSs \cite{Eshel:1982}.
+For the behavior or evolutionary ecologist interested in a mathematical model of natural selection, an ESS model offers one main advantage over population genetic models: the ESS model can be formulated entirely in terms of phenotypes and fitness and can avoid relying explicitly on genetic parameters (e.g., number of loci affecting the trait, recombination rates, etc) because the implicit genetic assumptions (i.e., additive interactions between alleles within a locus and across loci) imply a population genetic model with the same assumptions would generate the same ESSs \cite{Eshel:1982}.
 

### 2022-10-29 manuscript.tex
write more

@@
 \newcommand{\ess}[1]{#1^*}
+\newcommand{\fixp}[1]{\hat{#1}}
 \renewcommand{\vec}[1]{\mathbf{#1}}
+\newcommand{\rec}{\rho}
+\newcommand{\mut}{\mu}
 
@@
 The role of evolutionary game theory and the ESS method specifically in evolutionary biology has been at the crux of some of the most important conceptual debates in the field including the role of natural selection and adaptation vis-a-vis other forces \cite[e.g.,][]{MaynardSmith:1978,Gould:Lewontin:1979,Lewontin:1979,Orzack:Sober:1994,Gardner:2017,Kern:Hahn:2018,Jensen:Payseur:2019} and the importance of kin and group selection relative to individual selection
-\cite[e.g.,][]{Maynard-Smith:1964,Hamilton:1963,Price:1972:cov,Wilson:Wilson:2007,Leigh:2010,Akcay:Van-Cleve:2012,West:Griffin:2007,Gardner:Grafen:2009,Nowak:Tarnita:2010,Abbot:Abe:2011,Allen:Nowak:2013,Birch:2014,Birch:2017,Nowak:McAvoy:2017}.
+\cite[e.g.,][]{Maynard-Smith:1964,Hamilton:1963,Price:1972:cov,Wilson:Wilson:2007,Leigh:2010,Akcay:VanCleve:2012,West:Griffin:2007,Gardner:Grafen:2009,Nowak:Tarnita:2010,Abbot:Abe:2011,Allen:Nowak:2013,Birch:2014,Birch:2017,Nowak:McAvoy:2017}.
 These debates might be captured in part by the following two questions: \begin{ilnum} \item \label{q:I} How do ESS models that focus on individual fitness capture the effects of kin selection or group selection?
@@
 The levels of selection question became a major topic of study in evolutionary theory \cite[e.g.,][]{Lewontin:1970,Dawkins:1982,Wilson:Sober:1989,Maynard-Smith:Szathmary:1995,Wilson:1997,Michod:1999,Michod:2006,Szathmary:2015} and philosophy of biology \cite{Hull:1980,Brandon:1982,Damuth:Heisler:1988,Lloyd:1992,Lloyd:1994,Sober:Wilson:1994,Okasha:2006,Okasha:2016} and continues to generate substantial research \cite[e.g.,][]{Black:Bourrat:2020,Cooney:Mori:2022,Veit:2022}.
-For the purpose of explaining how an ESS can capture selection at multiple levels and among relatives or kin, I will argue below that the right measure of utility is the ``lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:Van-Cleve:2016,Lehmann:Mullon:2016} of a mutant allele at a single genetic locus, which can be shown to be functionally equivalent to a measure of inclusive fitness \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020}.
+For the purpose of explaining how an ESS can capture selection at multiple levels and among relatives or kin, I will argue below that the right measure of utility is the ``lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016} of a mutant allele at a single genetic locus, which can be shown to be functionally equivalent to a measure of inclusive fitness \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020}.
 
@@
 Early work in the simplest haploid asexual population genetic model with discrete strategies showed that the ESSs of obtained from condition \eqref{eq:ess:def} must be stable equilibria of the haploid asexual model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980}.
-Specifically, suppose there are two possible phenotypes $1$ and $2$ with the payoff matrix $A = \{a_{xy}\}$ where $a_{xy} = \w(x,y)$ is the payoff or increase to fitness of an individual of type $x$ interacting with an individual of type $y$.
+Specifically, suppose there are two possible phenotypes $1$ and $2$ with the payoff matrix $U = (u_{xy})$ where $u_{xy} = \w(x,y)$ is the payoff or increase to fitness of an individual of type $x$ interacting with an individual of type $y$.
 Then, the ESSs obtained by applying condition \eqref{eq:ess:def} to the payoff matrix $A$ are stable equilibria of a haploid asexual population genetic model with two genotypes, one for each phenotype \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Hofbauer:Sigmund:1998}.
@@
 Importantly, a similar  correspondence between ESSs and population genetic models also applies to a diploid sexual population reproducing in discrete time.
-Suppose there are two possible phenotypes controlled by a single genetic locus with $k$ possible alleles $A_{1},\ldots,A_{k}$ where dominance is possible between alleles and Mendelian segregation breaks up diploid genotypes every generation \cite{Eshel:1982,Lessard:1984}.
-Let diploid genotype $A_{i} A_{j}$ choose phenotype $1$ with probability $v_{ij}$ and phenotype $2$ with probability $1-v_{ij}$.
-If the $\vec{p} = (p_{1}, \ldots, p_{k})$ is the vector of allele frequencies for alleles $A_{1},\ldots,A_{k}$, then the mean frequency of phenotype $1$ in the population (in the next generation) is
+Suppose there are two possible phenotypes controlled by a single genetic locus with $n$ possible alleles $A_{1},\ldots,A_{n}$ where dominance is possible between alleles and Mendelian segregation breaks up diploid genotypes every generation \cite{Eshel:1982,Lessard:1984}.
+Let diploid genotype $A_{i} A_{j}$ choose phenotype $1$ with probability $x_{ij}$ and phenotype $2$ with probability $1-x_{ij}$.
+If the $\vec{p} = (p_{1}, \ldots, p_{n})$ is the vector of allele frequencies for alleles $A_{1},\ldots,A_{n}$, then the frequency of phenotype $1$ in the population after reproduction is
 \begin{equation*}
-  z(\vec{p}) = \sum_{i=1}^{k} p_{i} p_{j} v_{ij} \: .
+  x = x(\vec{p}) = \sum_{i,j} p_{i} p_{j} x_{ij} \: .
 \end{equation*}
-Eshel and Lessard showed \cite{Eshel:1982,Lessard:1984} that ESSs $\ess{x}$ of the payoff matrix $A$ are stably maintained in the diploid population model when those ESSs are produced by some vector of allele frequencies $\ess{\vec{p}}$ such that $\ess{x} = z(\ess{\vec{p}})$.
+Eshel and Lessard showed \cite{Eshel:1982,Lessard:1984} that ESSs $\ess{x}$ of the payoff matrix $A$ are stably maintained in the diploid population model when those ESSs are produced by some vector of allele frequencies $\ess{\vec{p}}$ such that $\ess{x} = x(\ess{\vec{p}})$.
 This correspondence also holds for continuous-time diploid models \cite{Sigmund:1987,Cressman:1988,Cressman:Hofbauer:1996,Hofbauer:Sigmund:1998} and for models with multiple loci when the probability of a genotype choosing a phenotype is an additive function of effects from each locus (i.e., additive non-epistasis) \cite{Eshel:1982}.
@@
 An important limitation of the above results connecting ESS phenotypes $\ess{x}$ to equilibria in diploid single and multilocus population genetic models is that the genotype-phenotype map may not produce phenotypes that are ESS and thus those ESS phenotypes are unattainable in the short-term.
-Specifically, the $v_{ij}$ may have values so that no population composition of genotypes $A_{i} A_{j}$ yields the ESS phenotype $\ess{x}$.
-Moreover, as mentioned above, multilocus models more generally (i.e., genetic systems with non-additive interactions or epistasis) do not guarantee that genotypes with higher marginal fitness\footnote{The marginal fitness of a genotype or allele is the mean fitness of that genotype allele among all contexts in which it occurs in the population.
+Specifically, the $x_{ij}$ may have values so that no population composition of genotypes $A_{i} A_{j}$ yields the ESS phenotype $\ess{x}$.
+Moreover, as mentioned above, multilocus systems more generally (i.e., systems with non-additive interactions or epistasis) do not guarantee that genotypes with higher marginal fitness\footnote{The marginal fitness of a genotype or allele is the mean fitness of that genotype allele among all contexts in which it occurs in the population.
 For example, the marginal fitness of one allele at a single diploid locus in the mean fitness of that allele in combination with all other alleles according to their frequencies.} increase in frequency, which is a requirement for the correspondence between ESS and population genetic equilibria described above \cite{Eshel:1982,Hofbauer:Sigmund:1998}.
 Thus, the genetic system may impose constraints on either the range of possible phenotypes or on the ability of selection to increase the frequency of high marginal fitness genotypes.
-Intriguingly, Eshel and others discovered that these constraints might not be so important on a longer term evolutionary timescales.
 
-In addition to showing that ESS phenotypes $\ess{x}$ are
+Intriguingly, Eshel, Feldman, and others discovered that these constraints might not be so important on a longer term evolutionary timescales.
+The important idea is that the conditions under which an genetic equilibrium is stable given a fixed set of alleles or genotypes does not determine whether a new allele or genotype outside of that fixed set might invade the population when rare.
+If an equilibrium is stable given the fixed set of alleles or genotypes, it is called ``internally stable'' \cite{Eshel:1996}.
+If that equilibrium also resists invasion by new alleles or genotypes outside the fixed set, it is called ``externally stable'' \cite{Liberman:1988,Eshel:1996}.
+Equilibria that are internally stable may not be externally stable.
+For example, suppose there is heterozygote advantage in a population with two alleles at a single diploid locus.
+The internally stable equilibrium is a polymorphism of both alleles.
+However, a new allele can invade the population if it generates the original heterozygote fitness in a heterozygote with either original allele and in a homozygote with itself.
+Hence, the internally stable polymorphism is not externally stable.
+The invasion of this new allele will lead to its fixation, which is a new internally stable equilibrium that may or may not be externally stable to further invasions depending on the the genotype-phenotype map (e.g., how the genotypes $A_{i} A_{j}$ generate fitness values $w_{ij}$).
+Given a fixed set of genotypes, the evolution of a fixed set of genotypes to an internally stable equilibrium is called ``short-term evolution'' whereas the  process of new mutant invasions that move the population from one internally stable equilibrium to another is called ``long-term evolution'' \cite{Eshel:1991,Eshel:1996,Hammerstein:1996,Weissing:1996,Eshel:Feldman:1998,VanCleve:2015,VanCleve:2020} and eventually leads to an externally stable equilibrium.
+
+Remarkably, despite the additional complexity of the long-term process, the equilibria of the long-term evolution process, not the short-term process, are what most closely align to what is predicted by the ESS criteria in \eqref{eq:ess:def}.
+Eshel, Feldman, Lessard and others demonstrated that externally stable equilibria are those that generate phenotypes $\ess{x}$ that are ESS \cite{Eshel:Feldman:1984,Liberman:1988,Eshel:1996,Hammerstein:Selten:1994,Hammerstein:1996,Weissing:1996,Eshel:Feldman:1998}.
+Specifically, suppose there are two phenotypes determined by two loci $A$ and $B$ with alleles $A_{1},\ldots,A_{n}$ and $B_{1},\ldots,B_{m}$, respectively, and with recombination rate $0<\rec\le1/2$ between the loci.
+The vector $\vec{p} = (\ldots,p_{ij},\ldots)$ tracks the frequencies of the chromosome $A_{i} B_{j}$ after selection and recombination.
+Individuals with diploid genotype $A_{i}B_{j}/A_{k}B_{l}$ chooses phenotype 1 with probability $x_{ijkl}$ and phenotype 2 with probability $1 - x_{ijkl}$.
+The frequency of phenotype 1 in the population after reproduction is
+\begin{equation*}
+  x = x(\vec{p}) = \sum_{i,j,k,l} p_{ij} p_{kl} x_{ijkl}
+\end{equation*}
+and the mean fitness of an individual with genotype $A_{i}B_{j}/A_{k}B_{l}$ interacting with other individuals in the population according to their frequencies is $w(x_{ijkl}, x)$ (since the fitness function is linear).
+The mean fitness in the population is then
+\begin{equation*}
+  \mean{w} = \sum_{ijkl} w(x_{ijkl}, x) p_{ij} p_{kl} = w(x, x) \: .
+\end{equation*}
+Now, assume that a new mutant allele $A_{\mut}$ (increasing the total number of alleles at locus $A$ to $n+1$) arrives at low frequency in a population at an internally stable equilibrium $\fixp{p}$ with mean phenotype $\fixp{x} = x(\fixp{p})$.
+An individual with this new allele chooses phenotype 1 with probability $x_{\mut jkl}$ if its genotype is $A_{\mut}B_{j}/A_{k}B_{l}$ and phenotype 2 with probability $1 - x_{\mut jkl}$.
+A linear stability analysis shows that this allele will either initially increase or decrease in frequency at a geometric rate $\lambda$ that is the largest eigenvalue of a Jacobian matrix \cite{Edelstein-Keshet:2005}.
+The eigenvalue $\lambda$ has an associated normalized eigenvector $\vec{v}$ that measures the frequency of mutant chromosomes (i.e., chromosomes $A_{\mut} B_{j}$) once the invasion process has stabilized (i.e., has reached stationarity \cite{Caswell:2006}).
+Eshel and Feldman \cite{Eshel:Feldman:1984} showed that this geometric growth rate is
+\begin{equation}
+  \label{eq:lambda:twolocus}
+  \lambda = \frac{w(x_{\mut}, \fixp{x})}{w(\fixp{x}, \fixp{x})}
+\end{equation}
+where $\lambda>0$ and
+\begin{equation*}
+  x_{\mut} = \sum_{jkl} v_{j} \fixp{p}_{kl} x_{\mut jkl}
+\end{equation*}
+is the frequency of phenotype 1 among mutant individuals once the invasion process has stabilized.
+Note that since the mutant allele is rare, mutant individuals can have only one mutant chromosome; the other chromosome has no mutant alleles at locus $A$ and remains at the internal equilibrium frequency $\fixp{p}_{kl}$.
 
@@
 Population genetic models showed that \cite{Cavalli-Sforza:Feldman:1978,Eshel:1982}
-
-Long-term evolution cites \cite{Eshel:1996,Hammerstein:1996,Weissing:1996}
-
-Though the idea of fitness optimization by itself has some intuitive appeal, high fitness individuals produce more surviving offspring who themselves are likely to be more fit than average, it needs an explicit evolutionary model in order to be evaluated mathematically and quantitatively.
 

### 2022-11-02 manuscript.tex
got modifier examples done

@@
 
+\usepackage{tikz}
+\usetikzlibrary{arrows}
+
 %%% new commands and macros
+
 \newcommand{\der}{\mathop{}\!\mathrm{d}}
@@
 \newcommand{\fixp}[1]{\hat{#1}}
-\renewcommand{\vec}[1]{\mathbf{#1}}
+\renewcommand{\vec}[1]{\symbf{#1}}
 \newcommand{\rec}{\rho}
 \newcommand{\mut}{\mu}
+
+\newtheorem{result}{Result}
+
+\newcommand{\xaxisandtickspip}{
+\begin{tikzpicture}[>=angle 90,scale=1.05]
+  % x-axis
+  \draw[] (0,0)--(3,0);
+  % Tick marks and annotations
+  %\path (1.5,-0.8) node [below]{$\mu_M$};
+
+  \foreach \x/\nodeLabel in {0/$10^{-3}$,
+    1.1115/$10^{-2}$,
+    2.2231/$10^{-1}$,
+    3/$0.5$}
+  {
+    \draw (\x,-0.1) -- (\x,0.1);
+    \path[anchor=center] (\x,-0.5) node[rotate=90] {\footnotesize\nodeLabel};
+  }
+\end{tikzpicture}
+}
+
+\newcommand{\yaxisandtickspip}{
+\begin{tikzpicture}[>=angle 90,scale=1.05]
+  % y-axis
+  \draw[] (0,0)--(0,3);
+  % Tick marks and annotations
+  %\path (-0.8,1.5) node [left]{$\mu_m$};
+
+  \foreach \y/\nodeLabel in {0/$10^{-3}$,
+    1.1115/$10^{-2}$,
+    2.2231/$10^{-1}$,
+    3/$0.5$}
+  {
+    \draw (-0.1,\y) -- (0.1,\y);
+    \path[anchor=center] (-0.5,\y) node {\footnotesize\nodeLabel};
+  }
+\end{tikzpicture}
+}
+
+%%%
 
@@
 Fitness $\w$ for the moment is simply the survival rate (i.e., viability selection) but as we'll see we can define much more general fitness measures.
-The stability condition \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} says that a phenotype $\ess{x}$ is an ESS when no alternative strategy $y$ can improve upon the fitness that $\ess{x}$ receives when interacting with itself:
+The stability condition \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} says that a phenotype $\ess{x}$ is an ESS, or is evolutionarily stable (ES), when no alternative strategy $y$ can improve upon the fitness that $\ess{x}$ receives when interacting with itself:
 \begin{subequations}
@@
 This question was tackled by mathematical population biologists beginning in the late 1970s and through 1980s and 1990s \cite[e..g,][]{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Eshel:1982,Hofbauer:Schuster:1982,Lessard:1984,Cressman:1988,Cressman:Hines:1984,Cressman:Hofbauer:1996,Hammerstein:1996,Weissing:1996,Eshel:1996,Eshel:Feldman:1984}.
-Early work in the simplest haploid asexual population genetic model with discrete strategies showed that the ESSs of obtained from condition \eqref{eq:ess:def} must be stable equilibria of the haploid asexual model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980}.
+Early work in the simplest haploid asexual population genetic model with discrete strategies showed that the ESSs obtained from condition \eqref{eq:ess:def} must be stable equilibria of the haploid asexual model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980}.
 Specifically, suppose there are two possible phenotypes $1$ and $2$ with the payoff matrix $U = (u_{xy})$ where $u_{xy} = \w(x,y)$ is the payoff or increase to fitness of an individual of type $x$ interacting with an individual of type $y$.
@@
 \end{equation*}
-Now, assume that a new mutant allele $A_{\mut}$ (increasing the total number of alleles at locus $A$ to $n+1$) arrives at low frequency in a population at an internally stable equilibrium $\fixp{p}$ with mean phenotype $\fixp{x} = x(\fixp{p})$.
+Now, assume that a new mutant allele $A_{\mut}$ (increasing the total number of alleles at locus $A$ to $n+1$) arrives at low frequency in a population at an internally stable equilibrium $\fixp{\vec{p}}$ with mean phenotype $\fixp{x} = x(\fixp{\vec{p}})$.
 An individual with this new allele chooses phenotype 1 with probability $x_{\mut jkl}$ if its genotype is $A_{\mut}B_{j}/A_{k}B_{l}$ and phenotype 2 with probability $1 - x_{\mut jkl}$.
@@
 The eigenvalue $\lambda$ has an associated normalized eigenvector $\vec{v}$ that measures the frequency of mutant chromosomes (i.e., chromosomes $A_{\mut} B_{j}$) once the invasion process has stabilized (i.e., has reached stationarity \cite{Caswell:2006}).
-Eshel and Feldman \cite{Eshel:Feldman:1984} showed that this geometric growth rate is
+Eshel and Feldman \cite{Eshel:Feldman:1984,Eshel:Feldman:1998} showed that this geometric growth rate is
 \begin{equation}
@@
 where $\lambda>0$ and
-\begin{equation*}
+\begin{equation}
+  \label{eq:mut:pheno:twolocus}
   x_{\mut} = \sum_{jkl} v_{j} \fixp{p}_{kl} x_{\mut jkl}
-\end{equation*}
+\end{equation}
 is the frequency of phenotype 1 among mutant individuals once the invasion process has stabilized.
 Note that since the mutant allele is rare, mutant individuals can have only one mutant chromosome; the other chromosome has no mutant alleles at locus $A$ and remains at the internal equilibrium frequency $\fixp{p}_{kl}$.
+A necessary condition for the equilibrium with phenotype $\fixp{x}$ to be externally stable is that $\lambda \le 1$ or from equation \eqref{eq:lambda:twolocus}
+\begin{equation*}
+  w(x_{\mut}, \fixp{x}) \le w(\fixp{x}, \fixp{x})
+\end{equation*}
+for all possible mutant $A_{\mut}$ that generate mutant phenotype $x_{\mut}$.
+This condition is quite close to the ESS condition in \eqref{eq:ess:def} and can be used to prove that the following result for randomly mating diploid populations with two-locus genetic systems \cite{Eshel:Feldman:1984,Eshel:1996,Eshel:Feldman:1998}:
+\begin{result}
+  \label{res:twolocus}
+  Any phenotype $\fixp{x}$ generated by an externally stable equilibrium is an ESS.
+Moreover, for ESS phenotype $\ess{x}$ generated by an internally stable equilibrium $\fixp{\vec{p}}$, $\fixp{\vec{p}}$ is externally stable.
+\end{result}
+Result \eqref{eq:lambda:twolocus} likely extends to more than two loci given work by Liberman \cite{Liberman:1988}.
+
+One important conclusion from Result \ref{res:twolocus} is that the ESS condition of Maynard Smith and Price \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} is an appropriate stability condition for the long-term evolution process.
+More generally, the fact that the ESS condition alone can describe the end points of long-term evolution (given the the general genetic assumptions above) is a mathematical demonstration of the importance of natural selection on phenotypes in determining long-term evolutionary trajectories.
+Thus, Result \eqref{eq:lambda:twolocus} provides a robust mathematical justification for the use of the ESS method or fitness optimization in building evolutionary models.
+
+However, Result \eqref{eq:lambda:twolocus} also shows the importance of genetics in the long-term evolution process.
+ESS phenotypes of the long-term process are produced by externally stable genotypic equilibria, and the conditions for the stability of those equilibria depend on specific assumptions about the genetic system of the evolving population (e.g., random mating and autosomal loci).
+If those assumptions change, then the ES phenotypes may depend on parameters of the genetic system.
+For example, the ES sex ratio \cite{Fisher:1930} in a randomly mating population is 1:1 when the ratio is determined by an diploid autosomal gene \cite{Hamilton:1967,Eshel:Feldman:1982}, but it can take other values when the gene is sex-linked or the genetic system is more complex like in insect that are haplodiploid \cite{Hamilton:1967,Eshel:Feldman:1982a}.
+A more recent example involves the evolution of bet-hedging where organisms alternative between different phenotypes with a specific probability or rate when the environmental cycles between different conditions \cite{Seger:Brockmann:1987,Slatkin:1974}.
+These phenotypic shifts can be due to genetic mutations \cite{Leigh:1970,Ishii:Matsuda:1989} or switching between different epigenetic states  \cite{Kussell:Leibler:2005,Salathe:VanCleve:2009} where control of the mutation or switching rate is determined by a ``modifier'' locus separate for the genetic or epigenetic locus controlling the phenotype.
+Due to the changing environment, this scenario does not fit neatly into the conditions required for Result \eqref{eq:lambda:twolocus}.
+In fact, an analysis of the external stability of the switching/mutation rate shows that the switching rate that is ES depends on the recombination rate between phenotypic locus and the modifier locus \cite{Liberman:VanCleve:2011}.
+For example, Figure \ref{fig:pip:switching}, shows pairwise invasibility plots for switching rate with different recombination rates and given weakly asymmetric selection in two different environments that alternate every $50$ generations.
+White regions are where mutant alleles with a switching rate on the veritcal axis can invade a population with a ``resident'' rate given by the horizontal axis; black regions are when such a mutant cannot invade.
+The plots show that a rate of $~1/50$ is ES when the recombination rate is zero (similar to older results in \cite{Leigh:1970,Lachmann:Jablonka:1996}).
+The ES switching rate decreases as recombination increases until there is an abrupt shift to an ES rate of zero switching.
+In addition to the ES sex ratio and switching rate, there are other examples where the genetic parameters of the modifier locus affect the ES value at that locus including dispersal and recombination in changing environments \cite{Carja:Liberman:2014}.
+
+\begin{figure}[ht!]
+  \begin{tabular}{r@{}r@{\hspace{-1.0ex}}c@{}c@{}c@{}c}
+    & & $R=0.0$ & $R=0.05$ & $R=0.1$ & $R=0.15$ \\
+    \cmidrule(lr){3-3}  \cmidrule(lr){4-4}  \cmidrule(lr){5-5} \cmidrule(lr){6-6}
+    \raisebox{-0.2cm}{\rotatebox{90}{Mutant switching rate}} & \raisebox{-0.2cm}{\yaxisandtickspip} &
+    \includegraphics[width=1.25in]{pip_n50_s1_0-01_s2_0-015_r_0-0.pdf} &
+    \includegraphics[width=1.25in]{pip_n50_s1_0-01_s2_0-015_r_0-05.pdf} &
+    \includegraphics[width=1.25in]{pip_n50_s1_0-01_s2_0-015_r_0-1.pdf} &
+    \includegraphics[width=1.25in]{pip_n50_s1_0-01_s2_0-015_r_0-15.pdf} \\
+    & & \xaxisandtickspip & \xaxisandtickspip & \xaxisandtickspip & \xaxisandtickspip \\
+    & & \multicolumn{4}{c}{Resident switching rate}
+  \end{tabular}
+  \caption{}
+  \label{fig:pip:switching}
+\end{figure}
 
@@
 
+This work was supported by National Science Foundation awards \#1846260 and \#1953223 to J.V.C.
+
 \clearpage

### 2022-11-02 manuscript.tex
small fixes

@@
 \end{result}
-Result \eqref{eq:lambda:twolocus} likely extends to more than two loci given work by Liberman \cite{Liberman:1988}.
+Result \ref{res:twolocus} likely extends to more than two loci given work by Liberman \cite{Liberman:1988}.
 
@@
 More generally, the fact that the ESS condition alone can describe the end points of long-term evolution (given the the general genetic assumptions above) is a mathematical demonstration of the importance of natural selection on phenotypes in determining long-term evolutionary trajectories.
-Thus, Result \eqref{eq:lambda:twolocus} provides a robust mathematical justification for the use of the ESS method or fitness optimization in building evolutionary models.
+Thus, Result \ref{res:twolocus} provides a robust mathematical justification for the use of the ESS method or fitness optimization in building evolutionary models.
 
-However, Result \eqref{eq:lambda:twolocus} also shows the importance of genetics in the long-term evolution process.
-ESS phenotypes of the long-term process are produced by externally stable genotypic equilibria, and the conditions for the stability of those equilibria depend on specific assumptions about the genetic system of the evolving population (e.g., random mating and autosomal loci).
-If those assumptions change, then the ES phenotypes may depend on parameters of the genetic system.
-For example, the ES sex ratio \cite{Fisher:1930} in a randomly mating population is 1:1 when the ratio is determined by an diploid autosomal gene \cite{Hamilton:1967,Eshel:Feldman:1982}, but it can take other values when the gene is sex-linked or the genetic system is more complex like in insect that are haplodiploid \cite{Hamilton:1967,Eshel:Feldman:1982a}.
-A more recent example involves the evolution of bet-hedging where organisms alternative between different phenotypes with a specific probability or rate when the environmental cycles between different conditions \cite{Seger:Brockmann:1987,Slatkin:1974}.
-These phenotypic shifts can be due to genetic mutations \cite{Leigh:1970,Ishii:Matsuda:1989} or switching between different epigenetic states  \cite{Kussell:Leibler:2005,Salathe:VanCleve:2009} where control of the mutation or switching rate is determined by a ``modifier'' locus separate for the genetic or epigenetic locus controlling the phenotype.
-Due to the changing environment, this scenario does not fit neatly into the conditions required for Result \eqref{eq:lambda:twolocus}.
-In fact, an analysis of the external stability of the switching/mutation rate shows that the switching rate that is ES depends on the recombination rate between phenotypic locus and the modifier locus \cite{Liberman:VanCleve:2011}.
-For example, Figure \ref{fig:pip:switching}, shows pairwise invasibility plots for switching rate with different recombination rates and given weakly asymmetric selection in two different environments that alternate every $50$ generations.
-White regions are where mutant alleles with a switching rate on the veritcal axis can invade a population with a ``resident'' rate given by the horizontal axis; black regions are when such a mutant cannot invade.
-The plots show that a rate of $~1/50$ is ES when the recombination rate is zero (similar to older results in \cite{Leigh:1970,Lachmann:Jablonka:1996}).
-The ES switching rate decreases as recombination increases until there is an abrupt shift to an ES rate of zero switching.
-In addition to the ES sex ratio and switching rate, there are other examples where the genetic parameters of the modifier locus affect the ES value at that locus including dispersal and recombination in changing environments \cite{Carja:Liberman:2014}.
-
-\begin{figure}[ht!]
+\begin{figure}[t!]
+  \small
   \begin{tabular}{r@{}r@{\hspace{-1.0ex}}c@{}c@{}c@{}c}
@@
 \end{figure}
+
+However, Result \ref{res:twolocus} also shows the importance of genetics in the long-term evolution process.
+ESS phenotypes of the long-term process are produced by externally stable genotypic equilibria, and the conditions for the stability of those equilibria depend on specific assumptions about the genetic system of the evolving population (e.g., random mating and autosomal loci).
+If those assumptions change, then the ES phenotypes may depend on parameters of the genetic system.
+For example, the ES sex ratio \cite{Fisher:1930} in a randomly mating population is 1:1 when the ratio is determined by an diploid autosomal gene \cite{Hamilton:1967,Eshel:Feldman:1982}, but it can take other values when the gene is sex-linked or the genetic system is more complex like in insect that are haplodiploid \cite{Hamilton:1967,Eshel:Feldman:1982a}.
+A more recent example involves the evolution of bet-hedging where organisms alternative between different phenotypes with a specific probability or rate when the environmental cycles between different conditions \cite{Seger:Brockmann:1987,Slatkin:1974}.
+These phenotypic shifts can be due to genetic mutations \cite{Leigh:1970,Ishii:Matsuda:1989} or switching between different epigenetic states  \cite{Kussell:Leibler:2005,Salathe:VanCleve:2009} where control of the mutation or switching rate is determined by a ``modifier'' locus separate for the genetic or epigenetic locus controlling the phenotype.
+Due to the changing environment, this scenario does not fit neatly into the conditions required for Result \ref{res:twolocus}.
+In fact, an analysis of the external stability of the switching/mutation rate shows that the switching rate that is ES depends on the recombination rate between phenotypic locus and the modifier locus \cite{Liberman:VanCleve:2011}.
+For example, Figure \ref{fig:pip:switching}, shows pairwise invasibility plots for switching rate with different recombination rates and given weakly asymmetric selection in two different environments that alternate every $50$ generations.
+White regions are where mutant alleles with a switching rate on the veritcal axis can invade a population with a ``resident'' rate given by the horizontal axis; black regions are when such a mutant cannot invade.
+The plots show that a rate of $~1/50$ is ES when the recombination rate is zero (similar to older results in \cite{Leigh:1970,Lachmann:Jablonka:1996}).
+The ES switching rate decreases as recombination increases until there is an abrupt shift to an ES rate of zero switching.
+In addition to the ES sex ratio and switching rate, there are other examples where the genetic parameters of the modifier locus affect the ES value at that locus including dispersal and recombination in changing environments \cite{Carja:Liberman:2014}.
 

### 2022-11-08 manuscript.tex
edits and added plot files

@@
 \newcommand{\mut}{\mu}
+\newcommend{\eig}{\lambda}
 
@@
 The two conditions in \eqref{eq:ess:def} are equivalent to requiring that phenotype $\ess{x}$ receives higher average fitness than any alternative phenotype $y$ when $y$ is rare in the population \cite{Maynard-Smith:1974,Bishop:Cannings:1976}.
+If condition \eqref{eq:ess:def} holds as a strict inequality, then $\ess{x}$ is a ``strict'' ESS.
 
-\section{How does ESS related to population-genetic dynamics?}
+\section{How is ESS related to population-genetic dynamics?}
 
@@
 \section{ESS and long-term evolution}
+\label{sec:long-term}
 
@@
 \end{equation*}
-Now, assume that a new mutant allele $A_{\mut}$ (increasing the total number of alleles at locus $A$ to $n+1$) arrives at low frequency in a population at an internally stable equilibrium $\fixp{\vec{p}}$ with mean phenotype $\fixp{x} = x(\fixp{\vec{p}})$.
+Now, assume that a new mutant allele $A_{\mut}$ (increasing the total number of alleles at locus $A$ to $n+1$) arrives at low frequency in a population at an internally stable equilibrium $\fixp{\vec{p}}$ with ``resident'' or wild-type mean phenotype $\fixp{x} = x(\fixp{\vec{p}})$.
 An individual with this new allele chooses phenotype 1 with probability $x_{\mut jkl}$ if its genotype is $A_{\mut}B_{j}/A_{k}B_{l}$ and phenotype 2 with probability $1 - x_{\mut jkl}$.
-A linear stability analysis shows that this allele will either initially increase or decrease in frequency at a geometric rate $\lambda$ that is the largest eigenvalue of a Jacobian matrix \cite{Edelstein-Keshet:2005}.
-The eigenvalue $\lambda$ has an associated normalized eigenvector $\vec{v}$ that measures the frequency of mutant chromosomes (i.e., chromosomes $A_{\mut} B_{j}$) once the invasion process has stabilized (i.e., has reached stationarity \cite{Caswell:2006}).
+A linear stability analysis shows that this allele will either initially increase or decrease in frequency at a geometric rate $\eig$ that is the largest eigenvalue of a Jacobian matrix \cite{Edelstein-Keshet:2005}.
+The eigenvalue $\eig$ has an associated normalized eigenvector $\vec{v}$ that measures the frequency of mutant chromosomes (i.e., chromosomes $A_{\mut} B_{j}$) once the invasion process has stabilized (i.e., has reached stationarity \cite{Caswell:2006}).
 Eshel and Feldman \cite{Eshel:Feldman:1984,Eshel:Feldman:1998} showed that this geometric growth rate is
@@
   \label{eq:lambda:twolocus}
-  \lambda = \frac{w(x_{\mut}, \fixp{x})}{w(\fixp{x}, \fixp{x})}
+  \eig = \frac{w(x_{\mut}, \fixp{x})}{w(\fixp{x}, \fixp{x})}
 \end{equation}
-where $\lambda>0$ and
+where $\eig>0$ and
 \begin{equation}
@@
 Note that since the mutant allele is rare, mutant individuals can have only one mutant chromosome; the other chromosome has no mutant alleles at locus $A$ and remains at the internal equilibrium frequency $\fixp{p}_{kl}$.
-A necessary condition for the equilibrium with phenotype $\fixp{x}$ to be externally stable is that $\lambda \le 1$ or from equation \eqref{eq:lambda:twolocus}
+A necessary condition for the equilibrium with phenotype $\fixp{x}$ to be externally stable is that $\eig \le 1$ or from equation \eqref{eq:lambda:twolocus}
 \begin{equation*}
@@
 \end{result}
+\noindent
 Result \ref{res:twolocus} likely extends to more than two loci given work by Liberman \cite{Liberman:1988}.
@@
 One important conclusion from Result \ref{res:twolocus} is that the ESS condition of Maynard Smith and Price \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} is an appropriate stability condition for the long-term evolution process.
-More generally, the fact that the ESS condition alone can describe the end points of long-term evolution (given the the general genetic assumptions above) is a mathematical demonstration of the importance of natural selection on phenotypes in determining long-term evolutionary trajectories.
-Thus, Result \ref{res:twolocus} provides a robust mathematical justification for the use of the ESS method or fitness optimization in building evolutionary models.
+More generally, given the the general genetic and demographic assumptions above, the fact that the ESS condition alone can describe the end points of long-term evolution is a mathematical demonstration of the importance of natural selection on phenotypes in determining long-term evolutionary trajectories.
+Thus, under the right assumptions, Result \ref{res:twolocus} provides a robust mathematical justification for the use of the ESS method in building evolutionary models.
+
+\section{Invasion and lineage fitness and ESSs}
 
@@
   \end{tabular}
-  \caption{}
+  \caption{Pairwise invasibility plots for the swithing/mutation rate model of \cite{Liberman:VanCleve:2011} when alternation between each of the two environments occurs every $n=50$ generations.
+The fitness costs of being in the wrong environments are $s_{1} = 0.01$ and $s_{2} = 0.015$ and the recombination rate between the phenotypic locus and the modifier is given above each plot.
+White regions are where a mutant modifier allele with a switching rate given on the horizontal axis can invade a population fixed for a modifier allele with switching rate given on the vertical axis, and black regions are where the mutant cannot invade; in other words, the leading eigenvalue of external stability matrix $\mathcal{L}_{\mathrm{ex}}$ in equation (17) of Liberman et al.
+\cite{Liberman:VanCleve:2011} is greater than one in the white regions and less than one in the black regions.}
   \label{fig:pip:switching}
@@
 
-However, Result \ref{res:twolocus} also shows the importance of genetics in the long-term evolution process.
-ESS phenotypes of the long-term process are produced by externally stable genotypic equilibria, and the conditions for the stability of those equilibria depend on specific assumptions about the genetic system of the evolving population (e.g., random mating and autosomal loci).
+While Result \ref{res:twolocus} emphasizes the role of ESSs and phenotypic evolution in the long-term process, the external stability approach used to prove \ref{res:twolocus} reveals a more general viewpoint where ESSs can depend in important ways on genetic and demographic parameters of a population.
+The external stability analysis assumes that the population is at some genetic equilibrium and is subject to potential invasion by a new mutant allele at low frequency and looks for a genetic equilibrium and its resulting phenotype that resists invasion by any mutant allele within a set of phenotypic effects.
+Before the mathematical machinery of external stability was developed, Hamilton appreciated the conceptual importance of uninvadable phenotypes in his study of the sex ratio, and he called sex ratios that resisted invasion ``unbeatable'' phenotypes \cite{Hamilton:1967}.
+As we saw above, determining whether a phenotype can be unbeatable and its genetic equilibria externally stable involves calculating the magnitude of the eigenvalue $\eig$ associated with the linear dynamical system that approximates the dynamics of the mutant allele frequency.
+Assuming that $\eig >0$, the phenotype $\ess{x}$ is unbeatable whenever $\eig(\ess{x},y) < 1$ for all $y \ne \ess{x}$ (within some set of effects).
+
+%% a few things:
+%% \item have to say \eig(x,x) = 1
+%% \item have to work with strict ESS
+%% \item maybe have to say this is for continuum of pure strategies?
+%% \item maybe I can dodge that?
+%% \item
+
+external stability analysis assumes that new mutant allele arises in a population at low enough frequency that so that its dynamics are well approximated by a linear approximation of the genotype frequency dynamical equations.
+These linear dynamical systems either grow or contract geometrically depending on the magnitude of the eigenvalue $\eig$ of the (Jacobian) matrix where growth occurs for $\eig > 1$ (assuming $\eig>0$) and contraction for $\eig < 1$.
+Thus, rare mutant alleles will either successfully invade and reach some nonzero frequency, after which point their dynamics are no longer linear, or they will go extinct.
+The magnitude of the eigenvalue $\eig$ measures the geometric growth rate (e.g., equation \eqref{eq:lambda:twolocus}) of the mutant genotype relative to the resident genotypes and so is really a measure of the fitness of the invading mutant genotype and is often called ``invasion fitness'' \cite{Metz:Nisbet:1992,Heino:Metz:1998} (relatedly, ``invasion exponent'' \cite{Rand:Wilson:1994}).
+Regardless of the complexity of the genetics, demography, or ecology of a population, we can in principle perform an external stability analysis\footnote{This is also referred to as an ``evolutionary invasion analysis''.} that uses invasion fitness to predict the fate of rare mutants.
+
+ESS phenotypes of the long-term process are produced by externally stable genetic equilibria, and the conditions for the stability of those equilibria depend on specific assumptions about the genetic system of the evolving population (e.g., random mating and autosomal loci).
+
+A deeper aspect of how genotypic and phenotypic processes are enmeshed in Result \ref{res:twolocus} comes from looking at how external stability generates the ESS condition, or how phenotypic evolutionary stability emerges out of genotypic evolutionary stability.
+
+Given a measure of invasion fitness, we can then look for a phenotype $x$ that cannot be invaded by any other phenotype;
+Hamilton called such phenotypes ``unbeatable'' \cite{Hamilton:1967}.
+An unbeatable phenotype $\ess{x}$ is one where all possible mutant phenotypes $y$ have an invasion fitness $\eig(y,\ess{x}) < 1$.
+In fact, an unbeatable phenotype $\ess{x}$ is a strict ESSs \cite{Eshel:Feldman:1998}.
+To see this, recall that from equation \eqref{eq:lambda:twolocus} that the invasion fitness is just the ratio of the mutant fitness against the resident to the resident fitness against itself, or $\eig(y,\ess{x}) = w(y,\ess{x})/w(\ess{x}, \ess{x})$.
+Since $\eig < 1$ for all mutant $y$ phenotypes, this means $w(y,\ess{x}) < w(\ess{x},\ess{x})$, which is the definition of a strict ESS from equation \eqref{eq:ess:def:a}.
+Thus, finding an ESS $\ess{x}$ means calculating the right invasion fitness for a population and showing that $\eig(\ess{x},y) < 1$ for all possible $y$.
+
+As we saw in section \ref{sec:long-term}, the invasion fitness $\eig$ has an associated eigenvector $v$ that measures the frequency of mutant genotypes during the invasion process.
+
 If those assumptions change, then the ES phenotypes may depend on parameters of the genetic system.
@@
 These phenotypic shifts can be due to genetic mutations \cite{Leigh:1970,Ishii:Matsuda:1989} or switching between different epigenetic states  \cite{Kussell:Leibler:2005,Salathe:VanCleve:2009} where control of the mutation or switching rate is determined by a ``modifier'' locus separate for the genetic or epigenetic locus controlling the phenotype.
-Due to the changing environment, this scenario does not fit neatly into the conditions required for Result \ref{res:twolocus}.
+Due to the the fitness effects of a changing environment, this scenario does not fit neatly into the conditions required for Result \ref{res:twolocus}.
 In fact, an analysis of the external stability of the switching/mutation rate shows that the switching rate that is ES depends on the recombination rate between phenotypic locus and the modifier locus \cite{Liberman:VanCleve:2011}.
 For example, Figure \ref{fig:pip:switching}, shows pairwise invasibility plots for switching rate with different recombination rates and given weakly asymmetric selection in two different environments that alternate every $50$ generations.
-White regions are where mutant alleles with a switching rate on the veritcal axis can invade a population with a ``resident'' rate given by the horizontal axis; black regions are when such a mutant cannot invade.
+White regions are where mutant alleles with a switching rate on the veritcal axis can invade a population with a resident rate for the rest of the population given by the horizontal axis; black regions are when such a mutant cannot invade.
 The plots show that a rate of $~1/50$ is ES when the recombination rate is zero (similar to older results in \cite{Leigh:1970,Lachmann:Jablonka:1996}).
@@
 In addition to the ES sex ratio and switching rate, there are other examples where the genetic parameters of the modifier locus affect the ES value at that locus including dispersal and recombination in changing environments \cite{Carja:Liberman:2014}.
-
-%%%
-%%% old scraps and ramblings
-%%%
-
-summary of results in chapter 22. for continuous time and two alleles or two phenotypes and linear payoffs, equilibria that produce ESS phenotypes are stable \cite{Hofbauer:Sigmund:1998}
-
-\section{Stability in dynamical systems and the ESS}
-
-For the behavior or evolutionary ecologist interested in a mathematical model of natural selection, an ESS model offers one main advantage over population genetic models: the ESS model can be formulated entirely in terms of phenotypes and fitness and can avoid relying explicitly on genetic parameters (e.g., number of loci affecting the trait, recombination rates, etc) because the implicit genetic assumptions (i.e., additive interactions between alleles within a locus and across loci) imply a population genetic model with the same assumptions would generate the same ESSs \cite{Eshel:1982}.
-
-\section{Long-term evolution}
-
-Following the development of the ESS approach by Maynard Smith and Price \cite{Maynard-Smith:Price:1973}, mathematical biologists evaluated the approach in light of the older and established theory population genetics.
-Population genetic models can explicitly address how both natural selection and genetic mechanisms including mutation, segregation, and recombination affect allele and genotype frequency dynamics over time.
-These models typically assume that evolution changes the frequencies of a fixed set of genotypes and that the emergence of new genotypes through the large-effect mutations (i.e., the creation of new alleles and loci) occurs on a longer timescale than that of the model.
-For this reason, these population genetic models are often said to represent a ``short-term'' evolutionary process.
-Population genetic models showed that \cite{Cavalli-Sforza:Feldman:1978,Eshel:1982}
 

### 2022-11-08 matlab/AeqFreq.m

@@ -1,0 +1,49 @@
+% Calculate one-locus equilibrium frequency for an allele in a fluctuating
+% environment.
+%
+% Jeremy Van Cleve <vancleve@stanford.edu>
+%
+% 2008/08/13: creation date
+
+function x = AeqFreq(mu, w11, w31, w12, w32, n)
+
+fh = @(y) fNestedMapC(y, mu, w11, w31, w12, w32, n) - y;
+
+initvals = linspace(0.1, 0.9, 5);
+errorf = 1;
+
+%disp(['mu ' num2str(mu) ' w11 ' num2str(w11) ' w31 ' num2str(w31) ' w12 ' num2str(w12) ' w32 ' num2str(w32) ...
+%      ' n ' num2str(n)]);
+%x = bisection(fh, 0, 1, 1e-6, 1e6);
+
+if (sign(fh(0))*sign(fh(1)) < 0)
+    x = fzero(fh, [0 1]);
+else
+    for i=1:5
+        x = fzero(fh, initvals(i));
+        if (x <= 1 && x >= 0)
+            errorf = 0;
+            break;
+        end
+    end
+end
+
+% if (errorf)
+%     error(['no root for mu ' num2str(mu) ' w1 ' num2str(w1) ' w3 ' num2str(w3) ' n ' num2str(n)]);
+% end
+
+% calculate nested map
+function f = fNestedMap(x, mu, w11, w31, w12, w32, n)
+
+f = x;
+for i=1:n
+    f = fMap(f, mu, w11, w31);
+end
+for i=1:n
+    f = fMap(f, mu, w32, w12);
+end
+
+% calculate one step iteration of allele frequency recursion
+function f = fMap(x, mu, w1, w3)
+    
+f = ((1-mu)*w1*x + mu*w3*(1-x)) / (w1*x + w3*(1-x));

### 2022-11-08 matlab/eigModInvas.m

@@ -1,0 +1,114 @@
+% Calculate leading eigenvalue (and derivative with respect to resident mutation rate)
+% of product of jacobian matrices for modifier allele invasion
+%
+% Jeremy Van Cleve <vancleve@stanford.edu>
+%
+% 2009/08/13: creation date
+
+function [ev dev d2ev] = eigModInvas(muRes, muMut, w11, w31, w12, w32, r, n)
+
+% calculate allele frequencies and mean fitness in cycle
+orbit = zeros(2*n,1);
+wbar = zeros(2*n,1);
+orbit(1) = AeqFreq(muRes, w11, w31, w12, w32, n);
+wbarprod = 1;
+for i=2:n+1
+    orbit(i) = fMap(orbit(i-1), muRes, w11, w31);
+    wbar(i-1) = w11*orbit(i-1) + w31*(1-orbit(i-1));
+    wbarprod = wbarprod * wbar(i-1);
+end
+for i=n+2:2*n
+    orbit(i) = fMap(orbit(i-1), muRes, w32, w12);
+    wbar(i-1) = w32*orbit(i-1) + w12*(1-orbit(i-1));
+    wbarprod = wbarprod * wbar(i-1);
+end
+wbar(2*n) = w11*orbit(2*n) + w31*(1-orbit(2*n));
+wbarprod = wbarprod * wbar(2*n);
+
+jprod = eye(2);
+jprod2 = eye(2);
+djprod = zeros(2);
+d2jprod = zeros(2);
+for i=1:n
+    jprod = jac(orbit(i), muMut, w11, w31, r) * jprod;
+    if (nargout > 2)
+        d2jprod = jac(orbit(i), muRes, w11, w31, r) * d2jprod ...
+                 + 2*djac(orbit(i), w11, w31, r) * djprod;
+    end
+    if (nargout > 1)
+        djprod = jac(orbit(i), muRes, w11, w31, r) * djprod ...
+                 + djac(orbit(i), w11, w31, r) * jprod2;
+        jprod2 = jac(orbit(i), muRes, w11, w31, r) * jprod2;
+    end
+end
+for i=n+1:2*n
+    jprod = jac(orbit(i), muMut, w32, w12, r) * jprod;
+    if (nargout > 2)
+        d2jprod = jac(orbit(i), muRes, w32, w12, r) * d2jprod ...
[…]

### 2022-11-08 matlab/eigPIP.m

@@ -1,0 +1,17 @@
+% Calculate pairwise invasibility plot using leading eigen value of
+% jacboian of mutation rate modifier
+%
+% Jeremy Van Cleve <vancleve@stanford.edu>
+%
+% 2009/10/05: creation date
+
+function pip = eigPIP(muvals, w11, w31, w12, w32, r, n)
+
+lenmuvals = length(muvals);
+pip = zeros(lenmuvals, lenmuvals);
+
+for i=1:lenmuvals
+    for j=1:lenmuvals
+        pip(i,j) = eigModInvas(muvals(j), muvals(i), w11, w31, w12, w32, r, n);
+    end
+end

### 2022-11-08 matlab/fNestedMapC.c

@@ -1,0 +1,61 @@
+/*
+ * =============================================================
+ * MEX file
+ *
+ * Jeremy Van Cleve <vancleve@stanford.edu>
+ * =============================================================
+ */
+ 
+#include "mex.h"
+#include <math.h>
+
+double fMapC(double x, double mu, double w1, double w3)
+{
+  return ((1-mu)*w1*x + mu*w3*(1-x)) / (w1*x + w3*(1-x));
+}
+
+void fNestedMapC(double xin, double *xout, double mu, double w11, double w31, double w12, double w32, int n)
+{
+  int i;
+  
+  *xout = xin;
+  for (i=0; i<n; i++) {
+    *xout = fMapC(*xout, mu, w11, w31);
+  }
+  for (i=0; i<n; i++) {
+    *xout = fMapC(*xout, mu, w32, w12);
+  }
+}
+
+void mexFunction(int nlhs, mxArray *plhs[], int nrhs,
+                 const mxArray *prhs[])
+{
+  double xin, mu, w11, w31, w12, w32, n;
+  double *xout;
+
+  /* Check for correct number of input and output arguments */
+  if (nrhs < 7) {
+    mexErrMsgTxt("This function requires five arguments.");
+  }
+  if (nlhs != 1) {
+    mexErrMsgTxt("This function requires one output.");
+  }
+  
+  /* Assign each input and output. */
+  xin = mxGetScalar(prhs[0]);
+  mu = mxGetScalar(prhs[1]);
+  w11 = mxGetScalar(prhs[2]);
+  w31 = mxGetScalar(prhs[3]);
+  w12 = mxGetScalar(prhs[4]);
+  w32 = mxGetScalar(prhs[5]);
+  n = mxGetScalar(prhs[6]);
+
+  /* Create matrix for the return argument. */
+  plhs[0] = mxCreateDoubleMatrix(1,1, mxREAL);
+
+  /* Pointer to output */
+  xout = mxGetPr(plhs[0]);
+
+  /* Call the subroutine. */
+  fNestedMapC(xin, xout, mu, w11, w31, w12, w32, n);
+}

### 2022-11-08 matlab/plot_cmds.m

@@ -1,0 +1,13 @@
+muvals = [logspace(-3, log10(1/2), 101)];
+
+s1 = 0.01;
+s2 = 0.015;
+
+% r = 0.0
+plotandsave_pip(eigPIP(muvals, 1, 1-s1, 1, 1-s2, 0.0,50), 'pip_n50_s1_0-01_s2_0-015_r_0-0.pdf', muvals)
+% r = 0.05
+plotandsave_pip(eigPIP(muvals, 1, 1-s1, 1, 1-s2, 0.05, 50), 'pip_n50_s1_0-01_s2_0-015_r_0-05.pdf', muvals)
+% r = 0.1
+plotandsave_pip(eigPIP(muvals, 1, 1-s1, 1, 1-s2, 0.1, 50), 'pip_n50_s1_0-01_s2_0-015_r_0-1.pdf', muvals)
+% r = 0.15
+plotandsave_pip(eigPIP(muvals, 1, 1-s1, 1, 1-s2, 0.15, 50), 'pip_n50_s1_0-01_s2_0-015_r_0-15.pdf', muvals)

### 2022-11-08 matlab/plotandsave_pip.m

@@ -1,0 +1,15 @@
+function [] = plotandsave_pip(pipmat, outputfile, muvals)
+
+logmuvals = log10(muvals);
+lenmu = length(muvals);
+incr = (lenmu-1)/(logmuvals(end)-logmuvals(1));
+ticks = [[1] [1 2] * incr lenmu];
+
+contourf(pipmat-1,[-inf 0 inf]);
+colormap([0 0 0; 1 1 1]);
+set(gca, 'Units', 'inches', 'Position', [1 1 2 2]); axis image;
+set(gca,'xtick', ticks,'ytick',ticks,'xticklabel',[],'yticklabel',[])
+set(gcf, 'Units', 'inches', 'Position', [0 0 4 4]);
+%set(gcf, 'PaperUnits', 'inches', 'PaperSize', [2.0 2.0]); 
+
+exportgraphics(gcf, outputfile, 'BackgroundColor', 'none', 'ContentType', 'vector')

### 2022-11-08 manuscript.tex
now onto examples

@@
 \usepackage{xparse}
-\usepackage{lineno}
+
+\usepackage[displaymath, mathlines]{lineno}
+\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
+  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
+  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
+  \renewenvironment{#1}%
+     {\linenomath\csname old#1\endcsname}%
+     {\csname oldend#1\endcsname\endlinenomath}}%
+\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
+  \patchAmsMathEnvironmentForLineno{#1}%
+  \patchAmsMathEnvironmentForLineno{#1*}}%
+\AtBeginDocument{%
+\patchBothAmsMathEnvironmentsForLineno{equation}%
+\patchBothAmsMathEnvironmentsForLineno{align}%
+\patchBothAmsMathEnvironmentsForLineno{flalign}%
+\patchBothAmsMathEnvironmentsForLineno{alignat}%
+\patchBothAmsMathEnvironmentsForLineno{gather}%
+\patchBothAmsMathEnvironmentsForLineno{multline}%
+}
 
@@
 \newcommand{\mut}{\mu}
-\newcommend{\eig}{\lambda}
+\newcommand{\eig}{\lambda}
 
@@
 These debates might be captured in part by the following two questions: \begin{ilnum} \item \label{q:I} How do ESS models that focus on individual fitness capture the effects of kin selection or group selection?
-\item \label{q:II} How does an ESS account for mutation, genetic drift, and other evolutionary forces?
+\item \label{q:II} How does an ESS account for recombination, mutation, genetic drift and other evolutionary forces?
 \end{ilnum} In this work, I will describe how conceptual advances in evolutionary theory since Maynard Smith and Price \cite{Maynard-Smith:Price:1973} have shed light on both of these questions and have led to a broader understanding of the ESS method and a more integrative framework for understanding how multiple evolutionary forces in complex populations can lead to diverse phenotypes.
 
-Question \ref{q:I} derives in part from how Maynard Smith introduced his work on ESSs; he argued that an ESS could provide an explanation for the evolution of behaviors where ``selection acts entirely at the individual level, but in which the success of any particular strategy depends on what strategies are adopted by other members of the population'' \cite[][p.
+Question \ref{q:I} derives in part from how Maynard Smith introduced his work on ESSs; he argued that an ESS could provide an explanation for the evolution of behaviors where ``selection acts entirely at the individual level, but in which the success of any particular strategy depends on what strategies are adopted by other members of the population'' \cite[p.
 210]{Maynard-Smith:1974} and is not ``due to group or species selection'' \cite[p.
@@
 kin selection are about the ``units'' or ``levels'' at which selection acts and how fitness should be measured to account for selection at those levels.
-Lewontin also hints at the levels of selection question in his 1961 game theory paper when he discusses the relative merits of individual-level measures of fitness like intrinsic or Malthusian growth rate $r$ versus population-level measures like mean fitness $\mean{w}$ \cite[p.
+Lewontin also hints at the levels of selection question in his 1961 game theory paper when he discusses the relative merits of individual-level measures of fitness like intrinsic or Malthusian growth rate $r$ versus population-level measures like mean fitness $\mean{w}$ \cite[pp.
 400-401]{Lewontin:1961} as analogs of utility.
@@
 The two conditions in \eqref{eq:ess:def} are equivalent to requiring that phenotype $\ess{x}$ receives higher average fitness than any alternative phenotype $y$ when $y$ is rare in the population \cite{Maynard-Smith:1974,Bishop:Cannings:1976}.
-If condition \eqref{eq:ess:def} holds as a strict inequality, then $\ess{x}$ is a ``strict'' ESS.
+If condition \eqref{eq:ess:def} holds as a strict inequality when $y \ne \ess{x}$, then $\ess{x}$ is a ``strict'' ESS.
 
@@
 Note that since the mutant allele is rare, mutant individuals can have only one mutant chromosome; the other chromosome has no mutant alleles at locus $A$ and remains at the internal equilibrium frequency $\fixp{p}_{kl}$.
-A necessary condition for the equilibrium with phenotype $\fixp{x}$ to be externally stable is that $\eig \le 1$ or from equation \eqref{eq:lambda:twolocus}
+A sufficient condition for the equilibrium with phenotype $\fixp{x}$ to be externally stable is that $\eig < 1$ or from equation \eqref{eq:lambda:twolocus}
 \begin{equation*}
-  w(x_{\mut}, \fixp{x}) \le w(\fixp{x}, \fixp{x})
+  w(x_{\mut}, \fixp{x}) < w(\fixp{x}, \fixp{x})
 \end{equation*}
 for all possible mutant $A_{\mut}$ that generate mutant phenotype $x_{\mut}$.
-This condition is quite close to the ESS condition in \eqref{eq:ess:def} and can be used to prove that the following result for randomly mating diploid populations with two-locus genetic systems \cite{Eshel:Feldman:1984,Eshel:1996,Eshel:Feldman:1998}:
+This condition is in fact the condition for a strict ESS from \eqref{eq:ess:def} and can be used to prove that the following result for randomly mating diploid populations with two-locus genetic systems \cite{Eshel:Feldman:1984,Eshel:1996,Eshel:Feldman:1998}:
 \begin{result}
@@
 
-\section{Invasion and lineage fitness and ESSs}
+\section{Invasion fitness and ESSs}
 
@@
 
-While Result \ref{res:twolocus} emphasizes the role of ESSs and phenotypic evolution in the long-term process, the external stability approach used to prove \ref{res:twolocus} reveals a more general viewpoint where ESSs can depend in important ways on genetic and demographic parameters of a population.
-The external stability analysis assumes that the population is at some genetic equilibrium and is subject to potential invasion by a new mutant allele at low frequency and looks for a genetic equilibrium and its resulting phenotype that resists invasion by any mutant allele within a set of phenotypic effects.
+While Result \ref{res:twolocus} emphasizes the role of ESSs and phenotypic evolution in the long-term process, the external stability approach used to prove \ref{res:twolocus} reveals a more general viewpoint where ESSs can depend in important ways on the genetic and demographic parameters of a population.
+External stability analysis assumes that the population is at some genetic equilibrium where new mutant alleles can invade at low frequency and looks for a genetic equilibrium and its resulting phenotype that resists invasion by any mutant allele within a set of phenotypic effects.
 Before the mathematical machinery of external stability was developed, Hamilton appreciated the conceptual importance of uninvadable phenotypes in his study of the sex ratio, and he called sex ratios that resisted invasion ``unbeatable'' phenotypes \cite{Hamilton:1967}.
-As we saw above, determining whether a phenotype can be unbeatable and its genetic equilibria externally stable involves calculating the magnitude of the eigenvalue $\eig$ associated with the linear dynamical system that approximates the dynamics of the mutant allele frequency.
-Assuming that $\eig >0$, the phenotype $\ess{x}$ is unbeatable whenever $\eig(\ess{x},y) < 1$ for all $y \ne \ess{x}$ (within some set of effects).
+As we saw above, determining whether a phenotype can be unbeatable and its genetic equilibria externally stable involves calculating the magnitude of the eigenvalue $\eig$ associated with the linear dynamical system that approximates the dynamics of the mutant allele frequency when rare.
+The magnitude of the eigenvalue $\eig$ measures the geometric growth rate (e.g., equation \eqref{eq:lambda:twolocus}) of all mutant genotypes in a population of resident genotypes.
+If there are multiple possible genotypes with the mutant allele, as is the case for the two-locus model in section \ref{sec:long-term} where there may be many possible alleles at the non-mutant locus, then mutant genotype fractions during the invasion process are given by the normalized right eigenvector of the linear system $\vec{v}$.
+Thus, the $\eig$ really measures the fitness of the whole mutant genetic lineage.
+The growth rate $\eig$ is often called the ``invasion fitness'' \cite{Metz:Nisbet:1992,Heino:Metz:1998}\footnote{When invasion occurs at more complex attractors like cycles, the ``invasion exponent'' is used, which is $\log |\eig|$ \cite{Rand:Wilson:1994}} and sometimes called the ``lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016,Lehmann:Rousset:2020}.
+In effect, the invasion or lineage fitness measures the number of offspring copies that a mutant allele produces among all the genotypes in the mutant lineage \cite{Lehmann:Mullon:2016}.
+Assuming that $\eig >0$, the phenotype $\ess{x}$ is unbeatable whenever
+\begin{equation}
+  \label{eq:unbeatable}
+  \eig(y, \ess{x}) < 1 \qquad \text{for all } y \ne x \: .
+\end{equation}
+Regardless of the complexity of the genetics, demography, or ecology of a population, we can in principle perform an external stability analysis\footnote{This is also referred to as an ``evolutionary invasion analysis''.} and use invasion fitness to determine whether there are unbeatable phenotypes.
 
-%% a few things:
-%% \item have to say \eig(x,x) = 1
-%% \item have to work with strict ESS
-%% \item maybe have to say this is for continuum of pure strategies?
-%% \item maybe I can dodge that?
-%% \item
+It should already be evident given Result \ref{res:twolocus} that the unbeatable phenotypes obtained via external stability are closely connected to phenotypes obtained via an ESS analysis using condition \eqref{eq:ess:def}.
+In the general case for an arbitrary genetic and demographic scenario, we assume only that we have the invasion fitness $\eig(y, \ess{x})$ of a rare mutant $y$ in the population that otherwise has mean phenotype $\ess{x}$.
+Starting with the condition for a strict ESS, $w(\ess{x}, \ess{x}) > w(y, \ess{x})$ for $y \ne \ess{x}$ from \eqref{eq:ess:def:a}, we convert that condition to one in terms of ``relative fitness'' by dividing by the mean population fitness $w(\ess{x}, \ess{x})$ to obtain
+\begin{equation}
+  \label{eq:strict:relative}
+  \frac{w(y, \ess{x})}{w(\ess{x}, \ess{x})} = \omega(y, \ess{x}) < 1 \qquad \text{for all } y \ne x
+\end{equation}
+where $\omega(y, \ess{x})$ is the relative fitness of a rare mutant population with mean phenotype $y$ in a population with mean phenotype $\ess{x}$.
+The strict ESS condition in \eqref{eq:strict:relative} using relative fitness is identical to the condition for uninvadability in condition \eqref{eq:unbeatable} using invasion fitness, which means that unbeatable or uninvadable phentoypes are strict ESSs\footnote{The relationship between invasion fitness and non-strict ESSs is more complex and depends on specific details of the genetics of the population.
+For example, see the section entitled ``Long-term Stability of an ESS'' in \cite{Eshel:Feldman:1998}}.
 
-external stability analysis assumes that new mutant allele arises in a population at low enough frequency that so that its dynamics are well approximated by a linear approximation of the genotype frequency dynamical equations.
-These linear dynamical systems either grow or contract geometrically depending on the magnitude of the eigenvalue $\eig$ of the (Jacobian) matrix where growth occurs for $\eig > 1$ (assuming $\eig>0$) and contraction for $\eig < 1$.
-Thus, rare mutant alleles will either successfully invade and reach some nonzero frequency, after which point their dynamics are no longer linear, or they will go extinct.
-The magnitude of the eigenvalue $\eig$ measures the geometric growth rate (e.g., equation \eqref{eq:lambda:twolocus}) of the mutant genotype relative to the resident genotypes and so is really a measure of the fitness of the invading mutant genotype and is often called ``invasion fitness'' \cite{Metz:Nisbet:1992,Heino:Metz:1998} (relatedly, ``invasion exponent'' \cite{Rand:Wilson:1994}).
-Regardless of the complexity of the genetics, demography, or ecology of a population, we can in principle perform an external stability analysis\footnote{This is also referred to as an ``evolutionary invasion analysis''.} that uses invasion fitness to predict the fate of rare mutants.
+Since we can in principle calculate invasion fitness for populations with complex genetics and demography, we can obtain strict ESSs in those populations even though Result \ref{res:twolocus} may not apply.
+However, those ESSs will almost certainly depend on genetic and demographic parameters of the population as the invasion fitness captures the effect of genetic and demographic processes on the growth of invading mutant lineages\footnote{see also the discussion of ``evolutionary genetic stability'' in \cite[pp.
+505--506]{Eshel:1996}}.
+The fact these ESSs incorporate genetic and demographic forces alongside natural selection shows that the ESS method is a more general tool than potentially originally envisioned by Maynard Smith and other proponents of the phentoypic gambit.
+Below, we will describe two cases that capture this flexibility: group- and class-structured populations with kin selection and populations in variable environments with bet-hedging phenotypes.
 
-ESS phenotypes of the long-term process are produced by externally stable genetic equilibria, and the conditions for the stability of those equilibria depend on specific assumptions about the genetic system of the evolving population (e.g., random mating and autosomal loci).
+\section{Group structure, inclusive fitness, and ESS}
 
-A deeper aspect of how genotypic and phenotypic processes are enmeshed in Result \ref{res:twolocus} comes from looking at how external stability generates the ESS condition, or how phenotypic evolutionary stability emerges out of genotypic evolutionary stability.
+\section{Bet-hedging, recombination, and ESS}
 
-Given a measure of invasion fitness, we can then look for a phenotype $x$ that cannot be invaded by any other phenotype;
-Hamilton called such phenotypes ``unbeatable'' \cite{Hamilton:1967}.
-An unbeatable phenotype $\ess{x}$ is one where all possible mutant phenotypes $y$ have an invasion fitness $\eig(y,\ess{x}) < 1$.
-In fact, an unbeatable phenotype $\ess{x}$ is a strict ESSs \cite{Eshel:Feldman:1998}.
-To see this, recall that from equation \eqref{eq:lambda:twolocus} that the invasion fitness is just the ratio of the mutant fitness against the resident to the resident fitness against itself, or $\eig(y,\ess{x}) = w(y,\ess{x})/w(\ess{x}, \ess{x})$.
-Since $\eig < 1$ for all mutant $y$ phenotypes, this means $w(y,\ess{x}) < w(\ess{x},\ess{x})$, which is the definition of a strict ESS from equation \eqref{eq:ess:def:a}.
-Thus, finding an ESS $\ess{x}$ means calculating the right invasion fitness for a population and showing that $\eig(\ess{x},y) < 1$ for all possible $y$.
-
-As we saw in section \ref{sec:long-term}, the invasion fitness $\eig$ has an associated eigenvector $v$ that measures the frequency of mutant genotypes during the invasion process.
-
-If those assumptions change, then the ES phenotypes may depend on parameters of the genetic system.
-For example, the ES sex ratio \cite{Fisher:1930} in a randomly mating population is 1:1 when the ratio is determined by an diploid autosomal gene \cite{Hamilton:1967,Eshel:Feldman:1982}, but it can take other values when the gene is sex-linked or the genetic system is more complex like in insect that are haplodiploid \cite{Hamilton:1967,Eshel:Feldman:1982a}.
 A more recent example involves the evolution of bet-hedging where organisms alternative between different phenotypes with a specific probability or rate when the environmental cycles between different conditions \cite{Seger:Brockmann:1987,Slatkin:1974}.
@@
 
-This work was supported by National Science Foundation awards \#1846260 and \#1953223 to J.V.C.
+This work was supported by National Science Foundation awards \#1846260 and \#1953223.
 

### 2022-11-08 manuscript.tex
another edit

@@
 505--506]{Eshel:1996}}.
-The fact these ESSs incorporate genetic and demographic forces alongside natural selection shows that the ESS method is a more general tool than potentially originally envisioned by Maynard Smith and other proponents of the phentoypic gambit.
+For example, the ES sex ratio depends on the genetic structure of the population: in a randomly mating population, the sex ratio is 1:1 when the ratio is determined by an diploid autosomal gene \cite{Fisher:1958,Hamilton:1967,Eshel:Feldman:1982}, but it can take other values when the gene is sex-linked or the genetic system is more complex like in insects that are haplodiploid \cite{Hamilton:1967,Eshel:Feldman:1982a}.
+The fact that ESSs can incorporate genetic and demographic forces alongside natural selection shows that the ESS method is a more general tool than potentially originally envisioned by Maynard Smith and other proponents of the phentoypic gambit.
 Below, we will describe two cases that capture this flexibility: group- and class-structured populations with kin selection and populations in variable environments with bet-hedging phenotypes.

### 2022-11-11 manuscript.tex
yay, closer to finishing the IF part

@@
 \setmathfont[range=bfit/{latin,Latin,greek,Greek}]{Minion Pro Bold It}
+\setmathfont[range=\mathcal]{Asana Math}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newcommand{\eig}{\lambda}
+\newcommand{\numc}{\mathcal{C}}
+\newcommand{\cost}{}
+\newcommand{\benefit}{}
 
@@
 
-\newpage
-
-\section{Outline/Brainstorm}
-
-\begin{enumerate}
-  \item Intro
-        \begin{enumerate}
-          \item Little intro about Maynard Smith and Price
-          \item Game theory as general optimization and search for fixed points
-          \item Connect back to Hamilton 1964 and goal of optimization with relatedness
-          \item Evolutionary process is dynamic but still has ``fixed points''
-          \item Evolutionary change can be decomposed and each piece affects location and stability of fixed points
-          \item Evol Game theory captures selection but not demography and genetic forces, which can have strong effects.
-        \end{enumerate}
-  \item Ok, so thinking more about this, the two main pieces in certain ways are the fitness or success measure and whether maximizing it leads a correct predictions.
-pop mean fitness as measure fails due to genetics and due to frequency dependence.
-hamilton suggested inclusive fitness.
-later authors showed that this is equivalent to an invasion fitness perspective.
-maximization of this leads to ESS.
-Stochasticity though due to mutation/drift/etc leads to stationary distribution...
-  \item Extra things
-        \begin{enumerate}
-          \item emphasize the incompleteness of the review. Maybe cite a few ``relevant'' Perc papers?
-        \end{enumerate}
-\end{enumerate}
-
-%%
-the flow for the intro should be something like: maynard smith setup this false dichotomy: ESS is about ``individual'' level selection when really there are two issues, ESS as tool of optimality and then the application of that tool to kin/group models.
-this requires explaining how ESS is a tool of optimality and how a particular optimality measure, namely lineage growth rate, accounts for kin and group selection since it can captures selection at all levels from the gene on up.
-accounting for other evolutionary forces is more difficult but one way forward uses the fixation probability and allows mutation/demography etc to affect fixation probabilities and hence transition rates.
-
 \clearpage
@@
 The levels of selection question became a major topic of study in evolutionary theory \cite[e.g.,][]{Lewontin:1970,Dawkins:1982,Wilson:Sober:1989,Maynard-Smith:Szathmary:1995,Wilson:1997,Michod:1999,Michod:2006,Szathmary:2015} and philosophy of biology \cite{Hull:1980,Brandon:1982,Damuth:Heisler:1988,Lloyd:1992,Lloyd:1994,Sober:Wilson:1994,Okasha:2006,Okasha:2016} and continues to generate substantial research \cite[e.g.,][]{Black:Bourrat:2020,Cooney:Mori:2022,Veit:2022}.
-For the purpose of explaining how an ESS can capture selection at multiple levels and among relatives or kin, I will argue below that the right measure of utility is the ``lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016} of a mutant allele at a single genetic locus, which can be shown to be functionally equivalent to a measure of inclusive fitness \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020}.
+For the purpose of explaining how an ESS can capture selection at multiple levels and among relatives or kin, I will argue below that the right measure of utility is the ``invasion fitness'' \cite{Metz:Nisbet:1992,Heino:Metz:1998} or ```lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016} of a mutant allele at a single genetic locus, which can be shown to be functionally equivalent to a measure of inclusive fitness \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020}.
 
@@
 I will argue below that the ESS method is more general than originally imagined by Maynard Smith and Price \cite{Maynard-Smith:Price:1973} in that in can in fact be incorporated into an evolutionary framework where natural selection and other evolutionary forces combine to shape the short and long-term evolution of traits.
-
-%%%
-%%% Detailed outline for post intro
-%%%
-
-What is ESS? ESS is just phenotypes and fitness and an equilibrium and stability condition. This matches pop gen for simple genetics \cite{Eshel:1982}.
-
-ESS really is a model of a ``long-term evolution'' process of invasion of mutant by resident where fitness determines invasion. This is the ``unbeatable strategy'' of Hamilton.
-
-Long-term evolution depends on a fitness ``measure'' so ``fitness'' isn't just individual fitness.
-Really, we track the frequency or density of individual alleles (i.e., gene's eye view).
-This is lineage growth rate.
-
-Lineage growth rate allows us to measure ESS in complex structure populations with kin and group selection.
-We measure the growth rate of the lineage in all the genetic and demographic environments it finds itself in.
-
-Genetic environments can include other loci linked by recombination. This is the case of the modifier theory for recombination, mutation, migration, etc. Example from Liberman 2011.
-
-Recognizing the importance of genetic parameters in ESS is the opposite of early work to justify ESS which tried to find conditions when ESS predicted same as pop gen but genetic parameters didn't appear.
-
-One way forward to understanding the complex phenotypic evolution including social evolution should involve using ESS theory that incorporates more genetics including multiple loci.
-This is notably complementary to methods that study trait coevolution, e.g., Mullon and Lehmann.
-
-%%%
-%%% Fleshed out outline material
-%%%
 
@@
 \begin{equation*}
-  \mean{w} = \sum_{ijkl} w(x_{ijkl}, x) p_{ij} p_{kl} = w(x, x) \: .
+  \mean{w} = \sum_{i,j,k,l} w(x_{ijkl}, x) p_{ij} p_{kl} = w(x, x) \: .
 \end{equation*}
@@
 A linear stability analysis shows that this allele will either initially increase or decrease in frequency at a geometric rate $\eig$ that is the largest eigenvalue of a Jacobian matrix \cite{Edelstein-Keshet:2005}.
-The eigenvalue $\eig$ has an associated normalized eigenvector $\vec{v}$ that measures the frequency of mutant chromosomes (i.e., chromosomes $A_{\mut} B_{j}$) once the invasion process has stabilized (i.e., has reached stationarity \cite{Caswell:2006}).
+The eigenvalue $\eig$ has an associated normalized eigenvector $\vec{u}$ that measures the frequency of mutant chromosomes (i.e., chromosomes $A_{\mut} B_{j}$) once the invasion process has stabilized (i.e., has reached stationarity \cite{Caswell:2006}).
 Eshel and Feldman \cite{Eshel:Feldman:1984,Eshel:Feldman:1998} showed that this geometric growth rate is
 \begin{equation}
-  \label{eq:lambda:twolocus}
+  \label{eq:eig:twolocus}
   \eig = \frac{w(x_{\mut}, \fixp{x})}{w(\fixp{x}, \fixp{x})}
@@
   \label{eq:mut:pheno:twolocus}
-  x_{\mut} = \sum_{jkl} v_{j} \fixp{p}_{kl} x_{\mut jkl}
+  x_{\mut} = \sum_{j,k,l} u_{j} \fixp{p}_{kl} x_{\mut jkl}
 \end{equation}
@@
 Note that since the mutant allele is rare, mutant individuals can have only one mutant chromosome; the other chromosome has no mutant alleles at locus $A$ and remains at the internal equilibrium frequency $\fixp{p}_{kl}$.
-A sufficient condition for the equilibrium with phenotype $\fixp{x}$ to be externally stable is that $\eig < 1$ or from equation \eqref{eq:lambda:twolocus}
+A sufficient condition for the equilibrium with phenotype $\fixp{x}$ to be externally stable is that $\eig < 1$ or from equation \eqref{eq:eig:twolocus}
 \begin{equation*}
@@
 
-\begin{figure}[t!]
-  \small
-  \begin{tabular}{r@{}r@{\hspace{-1.0ex}}c@{}c@{}c@{}c}
-    & & $R=0.0$ & $R=0.05$ & $R=0.1$ & $R=0.15$ \\
-    \cmidrule(lr){3-3}  \cmidrule(lr){4-4}  \cmidrule(lr){5-5} \cmidrule(lr){6-6}
-    \raisebox{-0.2cm}{\rotatebox{90}{Mutant switching rate}} & \raisebox{-0.2cm}{\yaxisandtickspip} &
-    \includegraphics[width=1.25in]{pip_n50_s1_0-01_s2_0-015_r_0-0.pdf} &
-    \includegraphics[width=1.25in]{pip_n50_s1_0-01_s2_0-015_r_0-05.pdf} &
-    \includegraphics[width=1.25in]{pip_n50_s1_0-01_s2_0-015_r_0-1.pdf} &
-    \includegraphics[width=1.25in]{pip_n50_s1_0-01_s2_0-015_r_0-15.pdf} \\
-    & & \xaxisandtickspip & \xaxisandtickspip & \xaxisandtickspip & \xaxisandtickspip \\
-    & & \multicolumn{4}{c}{Resident switching rate}
-  \end{tabular}
-  \caption{Pairwise invasibility plots for the swithing/mutation rate model of \cite{Liberman:VanCleve:2011} when alternation between each of the two environments occurs every $n=50$ generations.
-The fitness costs of being in the wrong environments are $s_{1} = 0.01$ and $s_{2} = 0.015$ and the recombination rate between the phenotypic locus and the modifier is given above each plot.
-White regions are where a mutant modifier allele with a switching rate given on the horizontal axis can invade a population fixed for a modifier allele with switching rate given on the vertical axis, and black regions are where the mutant cannot invade; in other words, the leading eigenvalue of external stability matrix $\mathcal{L}_{\mathrm{ex}}$ in equation (17) of Liberman et al.
-\cite{Liberman:VanCleve:2011} is greater than one in the white regions and less than one in the black regions.}
-  \label{fig:pip:switching}
-\end{figure}
-
 While Result \ref{res:twolocus} emphasizes the role of ESSs and phenotypic evolution in the long-term process, the external stability approach used to prove \ref{res:twolocus} reveals a more general viewpoint where ESSs can depend in important ways on the genetic and demographic parameters of a population.
@@
 As we saw above, determining whether a phenotype can be unbeatable and its genetic equilibria externally stable involves calculating the magnitude of the eigenvalue $\eig$ associated with the linear dynamical system that approximates the dynamics of the mutant allele frequency when rare.
-The magnitude of the eigenvalue $\eig$ measures the geometric growth rate (e.g., equation \eqref{eq:lambda:twolocus}) of all mutant genotypes in a population of resident genotypes.
-If there are multiple possible genotypes with the mutant allele, as is the case for the two-locus model in section \ref{sec:long-term} where there may be many possible alleles at the non-mutant locus, then mutant genotype fractions during the invasion process are given by the normalized right eigenvector of the linear system $\vec{v}$.
+The magnitude of the eigenvalue $\eig$ measures the geometric growth rate (e.g., equation \eqref{eq:eig:twolocus}) of all mutant genotypes in a population of resident genotypes.
+If there are multiple possible genotypes with the mutant allele, as is the case for the two-locus model in section \ref{sec:long-term} where there may be many possible alleles at the non-mutant locus, then mutant genotype fractions during the invasion process are given by the normalized right eigenvector of the linear system $\vec{u}$.
 Thus, the $\eig$ really measures the fitness of the whole mutant genetic lineage.
@@
 
-\section{Group structure, inclusive fitness, and ESS}
+\section{Inclusive fitness, group-structured populations, and ESS}
 
-\section{Bet-hedging, recombination, and ESS}
+In a populations where individuals live in groups and dispersal is limited, individuals are more genetically related to their groupmates than to individuals in different groups \cite{Wright:1943,Wright:1951,Hamilton:1970}, which means there is a possibility for kin selection to drive the evolution of prosocial phenotypes \cite{Hamilton:1964,Rousset:2004,Lehmann:Rousset:2014}.
+Moreover, group structure allows for competition among groups for resources and the possibility of selection among groups favoring more prosocial groups given sufficient group integrity \cite{Hamilton:1975,Wilson:Pollock:1992,Traulsen:Nowak:2006,Gardner:West:2006,Lehmann:Rousset:2010}.
+In fact, the Price equation \cite{Price:1970,Price:1972:cov} can be used to show that kin and group selection perspectives both lead to the same predictions when applied to the same population model \cite{Queller:1992,Lehmann:Keller:2007,Bijma:Wade:2008,Gardner:Grafen:2009,Marshall:2011,Akcay:VanCleve:2012}.
+Thus, we should expect that predicting an ESS for prosocial behavior in a group structured population by calculating invasion fitness will capture the effects of both kin and group selection.
+To see this, we will outline the invasion fitness analysis of this scenario performed by Lehmann and colleagues \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020}.
+This analysis also provides a formal justification for the intuition of Maynard Smith \cite{MaynardSmith:1978} and others that a correct fitness measure for ESS analysis is Hamilton's inclusive fitness \cite{Hamilton:1964} by showing that inclusive fitness is monotonically related to invasion fitness.
+
+Suppose a population is composed of an infinite number of groups each containing $n$ individuals; groups can be thought as demes or patches each  equally connected to one another at some rate of migration or dispersal (i.e., Wright's island model \cite{Wright:1931}).
+Each individual can belong to one of $\numc$ demographic classes, such as age, sex, or caste, such as workers and queens in social insects.
+Phenotypes are produced by alleles at a single haploid genetic locus.
+Alleles can be pleiotropic and code for specific phenotypes for each class: phenotype $\vec{x}=(x_{1},\ldots,x_{\numc})$ where $x_{s}$ is the phenotype for class $s$.
+We assume that once at a genetic and demographic equilibrium with phenotype $\vec{x}$, the population stabilizes at $n_{s}$ individuals in class $s$ in each group where the group size is just the sum of all the individuals in each class or $n = \sum_{s} n_{s}$.
+
+We introduce a mutant phenotype $\vec{y}$ at low frequency into a population where all individuals are resident phenotype $\vec{x}$.
+One way to track the number of mutant individuals is to track the number of groups with $k$ mutants.
+Since each individual belongs to a specific class, we actually track how many mutant individuals $k_{s}$ are in each class $s$ in a group or $\vec{k} = (k_{1},\ldots,k_{\numc})$ where $k = \sum_{s} k_{s}$.
+When the mutant phenotype is rare, growth of the mutant population can be modeled using a matrix population model \cite{Caswell:2006} or multitype branching process \cite{Kimmel:Axelrod:2015} with a projection matrix $\vec{A}(\vec{y}, \vec{x}) = [a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x})]$ where the $a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x})$ element measures the expected number of groups with $\vec{k}'$ mutants with phenotype $\vec{y}$ that result from the reproduction occurring in a group with $\vec{k}$ mutants in a population resident for phenotype $\vec{x}$.
+The largest eigenvalue $\eig(\vec{y}, \vec{x})$ of the projection matrix $\vec{A}(\vec{y}, \vec{x})$ is the invasion fitness in this population and satisfies
+\begin{equation}
+  \label{eq:Au:eigu}
+  \vec{A}(\vec{y}, \vec{x}) \vec{u}(\vec{y}, \vec{x}) = \eig(\vec{y}, \vec{x}) \vec{u}(\vec{y}, \vec{x})
+\end{equation}
+where the associated eigenvector $\vec{u}(\vec{y}, \vec{x})$ measures the relative frequency of groups with $\vec{k}$ mutants when the invasion process has stabilized.
+The elements $a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x})$ of the projection matrix can be rewritten in terms of an individual (relative) fitness function $\omega_{s's\vec{k}}(\vec{y}, \vec{x})$, which measures the number of class $s'$ mutant offspring in the next generation produced by a class $s$ mutant individual in a group with $\vec{k}$ mutants.
+Relative fitness is the right measure here since we measure offspring production after dispersal, density dependent survival, and all other life-cycle events.
+The projection matrix and fitness function are then related through the expression for the total number of class $s'$ offspring produced by a mutant individuals in a group with $\vec{k}$ mutants:
+\begin{equation}
+  \label{eq:a:w}
+  \sum_{k'} k_{s'} k_{s'}' a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x}) = \sum_{s} k_{s} \omega_{s's\vec{k}}(\vec{y}, \vec{x}) \: .
+\end{equation}
+\citeauthor{Lehmann:Mullon:2016} \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020} use equation \eqref{eq:a:w} to show that the invasion (or lineage) fitness is given by
+\begin{equation}
+  \label{eq:eig:class}
+  \eig(\vec{y}, \vec{x}) = \sum_{\vec{k},s',s} \omega_{s's\vec{k}}(\vec{y}, \vec{x}) q_{\vec{k},s}(\vec{y}, \vec{x})
+\end{equation}
+where $q_{\vec{k},s}(\vec{y}, \vec{x})$ is the probability (when the invasion process has stabilized) of a randomly sampled mutant allele from the mutant lineage is in a class $s$ individual in a group with $\vec{k}$ mutants and is given by
+\begin{equation*}
+  q_{\vec{k},s}(\vec{y}, \vec{x}) = \frac{k_{s} u_{\vec{k}}(\vec{y}, \vec{x})}{\sum_{\vec{k},s} k_{s} u_{\vec{k}}(\vec{y}, \vec{x})} \: .
+\end{equation*}
+The invasion fitness expression in equation \eqref{eq:eig:class} for a group- and class-structured population is conceptually identical to the one for a single population with two-loci in equation \eqref{eq:eig:twolocus}: both expressions show that the mutant successfully invades when mutant alleles more than replace themselves when averaged across all the genetic contexts (i.e., possible alleles at other loci) and demographic contexts (i.e., possible group compositions by mutant allele frequency in each class) in which the mutant lineage occurs during their invasion process.
+
+The invasion fitness in equation \eqref{eq:eig:class} can also be expressed rewritten in terms of inclusive fitness to isolate the effect of mutant allele expression on the fitness of the focal individual (direct effect) and on genetically related individuals (indirect effect).
+There are some important technical subtleties as to how this can be accomplished that we will leave out to keep the presentation simple (see ``Supplement B'' in \cite{Lehmann:Rousset:2020} for details).
+The essence of the method is the same as the one found in many analyses that derive inclusive fitness using the Price equation \cite[e.g.,]{Queller:1992,Frank:1998,Lehmann:Keller:2006,Akcay:VanCleve:2012}: the fitness of a focal individual, here $\omega_{s's\vec{k}}(\vec{y}, \vec{x})$, decomposed into terms that reflects the marginal fitness effects of ``changing'' an allele from expressing the resident phenotype $x_{s}$ to the mutant phenotype $y_{s}$ in a class $s$ individual.
+The cost term
+
+\section{Variable environments, recombination, and ESS}
+
+\begin{figure}[t!]
+  \small
+  \begin{tabular}{r@{}r@{\hspace{-1.0ex}}c@{}c@{}c@{}c}
+    & & $R=0.0$ & $R=0.05$ & $R=0.1$ & $R=0.15$ \\
+    \cmidrule(lr){3-3}  \cmidrule(lr){4-4}  \cmidrule(lr){5-5} \cmidrule(lr){6-6}
+    \raisebox{-0.2cm}{\rotatebox{90}{Mutant switching rate}} & \raisebox{-0.2cm}{\yaxisandtickspip} &
+    \includegraphics[width=1.25in]{pip_n50_s1_0-01_s2_0-015_r_0-0.pdf} &
+    \includegraphics[width=1.25in]{pip_n50_s1_0-01_s2_0-015_r_0-05.pdf} &
+    \includegraphics[width=1.25in]{pip_n50_s1_0-01_s2_0-015_r_0-1.pdf} &
+    \includegraphics[width=1.25in]{pip_n50_s1_0-01_s2_0-015_r_0-15.pdf} \\
+    & & \xaxisandtickspip & \xaxisandtickspip & \xaxisandtickspip & \xaxisandtickspip \\
+    & & \multicolumn{4}{c}{Resident switching rate}
+  \end{tabular}
+  \caption{Pairwise invasibility plots for the swithing/mutation rate model of \cite{Liberman:VanCleve:2011} when alternation between each of the two environments occurs every $n=50$ generations.
+The fitness costs of being in the wrong environments are $s_{1} = 0.01$ and $s_{2} = 0.015$ and the recombination rate between the phenotypic locus and the modifier is given above each plot.
+White regions are where a mutant modifier allele with a switching rate given on the horizontal axis can invade a population fixed for a modifier allele with switching rate given on the vertical axis, and black regions are where the mutant cannot invade; in other words, the leading eigenvalue of external stability matrix $\mathcal{L}_{\mathrm{ex}}$ in equation (17) of Liberman et al.
+\cite{Liberman:VanCleve:2011} is greater than one in the white regions and less than one in the black regions.}
+  \label{fig:pip:switching}
+\end{figure}
 
@@
 
-\section{Acknowledgements}
+% \section{Acknowledgements}
 
@@
 
+%% Outline/Brainstorm
+
+\begin{enumerate}
+  \item Intro
+        \begin{enumerate}
+          \item Little intro about Maynard Smith and Price
+          \item Game theory as general optimization and search for fixed points
+          \item Connect back to Hamilton 1964 and goal of optimization with relatedness
+          \item Evolutionary process is dynamic but still has ``fixed points''
+          \item Evolutionary change can be decomposed and each piece affects location and stability of fixed points
+          \item Evol Game theory captures selection but not demography and genetic forces, which can have strong effects.
+        \end{enumerate}
+  \item Ok, so thinking more about this, the two main pieces in certain ways are the fitness or success measure and whether maximizing it leads a correct predictions.
+pop mean fitness as measure fails due to genetics and due to frequency dependence.
+hamilton suggested inclusive fitness.
+later authors showed that this is equivalent to an invasion fitness perspective.
+maximization of this leads to ESS.
+Stochasticity though due to mutation/drift/etc leads to stationary distribution...
+  \item Extra things
+        \begin{enumerate}
+          \item emphasize the incompleteness of the review. Maybe cite a few ``relevant'' Perc papers?
+        \end{enumerate}
+\end{enumerate}
+
+%%
+the flow for the intro should be something like: maynard smith setup this false dichotomy: ESS is about ``individual'' level selection when really there are two issues, ESS as tool of optimality and then the application of that tool to kin/group models.
+this requires explaining how ESS is a tool of optimality and how a particular optimality measure, namely lineage growth rate, accounts for kin and group selection since it can captures selection at all levels from the gene on up.
+accounting for other evolutionary forces is more difficult but one way forward uses the fixation probability and allows mutation/demography etc to affect fixation probabilities and hence transition rates.
+
+%%%
+%%% Detailed outline for post intro
+%%%
+
+What is ESS? ESS is just phenotypes and fitness and an equilibrium and stability condition. This matches pop gen for simple genetics Eshel:1982.
+
+ESS really is a model of a ``long-term evolution'' process of invasion of mutant by resident where fitness determines invasion. This is the ``unbeatable strategy'' of Hamilton.
+
+Long-term evolution depends on a fitness ``measure'' so ``fitness'' isn't just individual fitness.
+Really, we track the frequency or density of individual alleles (i.e., gene's eye view).
+This is lineage growth rate.
+
+Lineage growth rate allows us to measure ESS in complex structure populations with kin and group selection.
+We measure the growth rate of the lineage in all the genetic and demographic environments it finds itself in.
+
+Genetic environments can include other loci linked by recombination. This is the case of the modifier theory for recombination, mutation, migration, etc. Example from Liberman 2011.
+
+Recognizing the importance of genetic parameters in ESS is the opposite of early work to justify ESS which tried to find conditions when ESS predicted same as pop gen but genetic parameters didn't appear.
+
+One way forward to understanding the complex phenotypic evolution including social evolution should involve using ESS theory that incorporates more genetics including multiple loci.
+This is notably complementary to methods that study trait coevolution, e.g., Mullon and Lehmann.
+
+%%%
+%%% Fleshed out outline material
+%%%
+
 maybe my argument is then (see below) that population genetic model that include many forces can be converted into long-term evolution models.

### 2022-11-14 manuscript.tex
edits plus new abstract

@@
 \newcommand{\eig}{\lambda}
-\newcommand{\numc}{\mathcal{C}}
+\newcommand{\numc}{\mathcal{S}}
 \newcommand{\cost}{}
@@
 
-The application of game theory to animal behavior by Maynard Smith and Price invigorated evolutionary biology by suggesting that the complex dynamics among natural selection, genes, and behavior could be analyzed using an evolutionary refinement of the Nash equilibrium, namely the evolutionarily stable strategy (ESS)\@.
+  The application of game theory to animal behavior by Maynard Smith and Price invigorated evolutionary biology by suggesting that the complex dynamics among natural selection, genes, and behavior could be analyzed using an evolutionary refinement of the Nash equilibrium, namely the evolutionarily stable strategy (ESS).
 As implied by ``equilibrium'' and ``stable'', an ESS is a predicted long-term outcome that an evolutionary process may reach via natural selection and mutation.
-ESS analysis has been particularly important in highlighting the importance of social dilemmas in biology where the evolution of cooperative behavior can be undermined by the individual fitness costs of that behavior.
-However, even simple games like the stag-hunt or coordination game can have multiple ESSs, which begs the question, which ESS is more likely to evolve?
-Unfortunately, ESS analysis does not reveal which ESS is more likely to evolve nor does it even show how genetic or demographic factors affect the speed of approach to an ESS.
-Progress on determining which ESSs are more ``realistic'' has been made using tools from stochastic processes and population and quantitative genetics including stochastic stability, absorption times, and G-matrix theory.
-In this paper, we will review the use of these tools and provide novel insight into how they can predict social evolution as a function of genetic and demographic processes.
+
+  In addition to its intuitiveness, ESS analysis makes modeling phenotypic evolution relatively simple as it often does not require the modeler to state their assumptions about the genetics and demography of a population in contrast to explicit population or quantitative genetic models.
+Though this simplicity is useful, it can obscure the degree to which ESS analysis is applicable in populations with more realistic genetic and demography: for example how does ESS analysis handle  interactions with relatives (kin selection), selection across levels (group selection), and variable environments when phenotypes are affected by multiple genes?
+
+  In this paper, I will review the history of the ESS concept and show how early uncertainty about the method lead to deeper theoretical work linking ESS analysis to very general population models through the theory of ``long-term evolution''.
+A crucial tool in long-term evolution is the concept of ``invasion fitness'', which can be calculated for populations with complex genetics and demography and can be used to predict ESSs.
+These ESSs often depend crucially on demographic and genetic parameters.
+Given the increasing abundance of genetic and long-term demographic data in natural populations, this means that ESS analysis will likely be an important tool in understanding these data with evolutionary models.
 
@@
 Let $\w(x,y)$ be the fitness that an individual with phenotype or strategy $x$ obtains when interacting with an individual with phenotype or strategy $y$.
-The phenotypes can be drawn from a set of discrete or continous set of values $X$ that represent a single trait like body size or propensity to cooperate.
+The phenotypes can be drawn from a set of discrete or continuous set of values $X$ that represent a single trait like body size or propensity to cooperate.
 Fitness $\w$ for the moment is simply the survival rate (i.e., viability selection) but as we'll see we can define much more general fitness measures.
@@
 Moreover, group structure allows for competition among groups for resources and the possibility of selection among groups favoring more prosocial groups given sufficient group integrity \cite{Hamilton:1975,Wilson:Pollock:1992,Traulsen:Nowak:2006,Gardner:West:2006,Lehmann:Rousset:2010}.
-In fact, the Price equation \cite{Price:1970,Price:1972:cov} can be used to show that kin and group selection perspectives both lead to the same predictions when applied to the same population model \cite{Queller:1992,Lehmann:Keller:2007,Bijma:Wade:2008,Gardner:Grafen:2009,Marshall:2011,Akcay:VanCleve:2012}.
+In fact, the Price equation \cite{Price:1970,Price:1972:cov} can be used to show that kin and group selection perspectives on fitness both lead to the same predictions when applied to the same population model \cite{Queller:1992,Lehmann:Keller:2007,Bijma:Wade:2008,Gardner:West:2007,Marshall:2011,Akcay:VanCleve:2012}.
 Thus, we should expect that predicting an ESS for prosocial behavior in a group structured population by calculating invasion fitness will capture the effects of both kin and group selection.
@@
 \end{equation}
-\citeauthor{Lehmann:Mullon:2016} \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020} use equation \eqref{eq:a:w} to show that the invasion (or lineage) fitness is given by
+\citeauthor{Lehmann:Mullon:2016} \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020} use equation \eqref{eq:a:w} to show that the invasion (or lineage) fitness of mutant phenotype $\vec{x}$ in a resident population with phenotype $\vec{y}$ is given by
 \begin{equation}
@@
 
-The invasion fitness in equation \eqref{eq:eig:class} can also be expressed rewritten in terms of inclusive fitness to isolate the effect of mutant allele expression on the fitness of the focal individual (direct effect) and on genetically related individuals (indirect effect).
+The invasion fitness in equation \eqref{eq:eig:class} can also be expressed in terms of inclusive fitness in order to isolate the effect on fitness of mutant allele expression in the focal individual (direct effect) and in genetically related individuals (indirect effect).
 There are some important technical subtleties as to how this can be accomplished that we will leave out to keep the presentation simple (see ``Supplement B'' in \cite{Lehmann:Rousset:2020} for details).
-The essence of the method is the same as the one found in many analyses that derive inclusive fitness using the Price equation \cite[e.g.,]{Queller:1992,Frank:1998,Lehmann:Keller:2006,Akcay:VanCleve:2012}: the fitness of a focal individual, here $\omega_{s's\vec{k}}(\vec{y}, \vec{x})$, decomposed into terms that reflects the marginal fitness effects of ``changing'' an allele from expressing the resident phenotype $x_{s}$ to the mutant phenotype $y_{s}$ in a class $s$ individual.
-The cost term
+The essence of the method is the same as the one found in many analyses that derive inclusive fitness using the Price equation \cite[e.g.,]{Queller:1992,Frank:1998,Lehmann:Keller:2006,Akcay:VanCleve:2012}: the fitness of a focal individual, here $\omega_{s's\vec{k}}(\vec{y}, \vec{x})$, is decomposed into cost and benefit terms reflecting the marginal effects on fitness\footnote{Marginal effects in this context means additive effects of the phenotype on fitness and often measured with partial regression coefficients.} of expressing the mutant phenotype $y_{s}$ relative to the resident phenotype $x_{s}$.
+The cost term $c_{s's}(\vec{y}, \vec{x})$ measures the marginal decrease in class $s'$ offspring produced by a class $s$ individual due to expression of the mutant allele in that individual.
+The benefit term $b_{s'\sigma \leftarrow s}(\vec{y}, \vec{x})$ measures the marginal increase in the number of class $s'$ offspring produced individuals of class $\sigma$ due to the mutant allele expression by a groupmate of class $s$.
+The invasion fitness can then be expressed as (using an ``actor-modulated'' approach) \cite{Lehmann:Rousset:2020}
+\begin{equation}
+  \label{eq:eig:IF}
+  \eig(\vec{y}, \vec{x}) = 1 + \sum_{s',s} q_{s}(\vec{y}, \vec{x}) v_{s'}(\vec{y}, \vec{x})
+  \left( - c_{s's}(\vec{y}, \vec{x}) + \sum_{\sigma} b_{s'\sigma \leftarrow s}(\vec{y}, \vec{x}) r_{\sigma|s}(\vec{y}, \vec{x}) \right)
+\end{equation}
+where $r_{\sigma|s}(\vec{y}, \vec{x})$ is genetic relatedness and is the probability that a random groupmate of class $\sigma$ has the mutant allele given an individual of class $s$ has the mutant allele, $v_{s'}(\vec{y}, \vec{x})$ is the normalized reproductive value of class $s'$, and $q_{s}(\vec{y}, \vec{x})$ is the probability that a mutant individual in the invasion process is in class $s$ (see ref.
+\cite{Lehmann:Rousset:2020} for technical definitions of these quantities).
+Equation \eqref{eq:eig:IF} shows how invasion fitness can be partitioned into the direct effects of mutant allele expression, which are cost terms, and the indirect effects, which are the benefit times relatedness terms.
+Thus, kin selection and the evolution of altruistic traits are easily captured with an ESS analysis of invasion fitness.
+
+Given that previous work using the Price equation has shown that kin selection and group selection partitions of fitness are simply alternative ways of expressing genotype frequency change in a population \cite{Queller:1992,Bijma:Wade:2008,Gardner:West:2007,Marshall:2011,Akcay:VanCleve:2012}, it should be possible to rewrite invasion fitness so as to group terms that constitute between-group selection, $\omega_{G,s's}(\vec{y}, \vec{x})$, and those that constitute within-group selection, $\omega_{\Delta G,s's}(\vec{y}, \vec{x})$.
+In fact, we can do this by applying a ``neighbor-modulated'' approach from refs.
+\cite{Queller:1992,Akcay:VanCleve:2012} to the results from ref.
+\cite{Lehmann:Rousset:2020} and obtaining
+\begin{equation}
+  \label{eq:eig:GS}
+  \eig(\vec{y}, \vec{x}) = 1 + \sum_{s',s} q_{s}(\vec{y}, \vec{x}) v_{s'}(\vec{y}, \vec{x})
+  \left( \omega_{G,s's}(\vec{y}, \vec{x}) + \omega_{\Delta G,s's}(\vec{y}, \vec{x}) \right)
+\end{equation}
+where, dropping the dependence on $\vec{x}$ and $\vec{y}$ for ease of presentation,
+\begin{equation}
+  \label{eq:wg:wdeltag}
+  \begin{split}
+    \omega_{G,s's} & = \left(b_{s's \leftarrow s} - c_{s's} \right) \left(\frac{1}{n_{s}} + \frac{n_{s}-1}{n_{s}} r_{\sigma|s} \right) + \sum_{\sigma \ne s} b_{s's \leftarrow \sigma} r_{\sigma|s} \\
+    \omega_{\Delta G,s's} & = - \left(b_{s's \leftarrow s} + c_{s's}\right) \frac{n_{s}-1}{n_{s}} \left( 1 - r_{\sigma|s} \right) \: .
+  \end{split}
+\end{equation}
+We can see immediately that if mutant allele expression is costly to the individual itself, $c_{s's}>0$, helps others, $b_{s's \leftarrow \sigma}>0$ for all classes $\sigma$, and the benefits outweigh the costs within a class, $b_{s's \leftarrow s} - c_{s's}>0$, then the mutant allele expression is positively selected by between-group selection and negatively selected by within-group selection.
+Since both the inclusive fitness and group selection partitions of invasion fitness (equations \eqref{eq:eig:IF} and \eqref{eq:eig:GS}, respectively) are expressed using the same benefit, cost, and relatedness functions, a singular ESS analysis using invasion can answer questions derived both both perspectives.
 
@@
 
-\begin{enumerate}
-  \item Intro
-        \begin{enumerate}
-          \item Little intro about Maynard Smith and Price
-          \item Game theory as general optimization and search for fixed points
-          \item Connect back to Hamilton 1964 and goal of optimization with relatedness
-          \item Evolutionary process is dynamic but still has ``fixed points''
-          \item Evolutionary change can be decomposed and each piece affects location and stability of fixed points
-          \item Evol Game theory captures selection but not demography and genetic forces, which can have strong effects.
-        \end{enumerate}
-  \item Ok, so thinking more about this, the two main pieces in certain ways are the fitness or success measure and whether maximizing it leads a correct predictions.
+\section{extra stuff}
+- Intro
+  - Little intro about Maynard Smith and Price
+  - Game theory as general optimization and search for fixed points
+  - Connect back to Hamilton 1964 and goal of optimization with relatedness
+  - Evolutionary process is dynamic but still has ``fixed points''
+  - Evolutionary change can be decomposed and each piece affects location and stability of fixed points
+  - Evol Game theory captures selection but not demography and genetic forces, which can have strong effects.
+- Ok, so thinking more about this, the two main pieces in certain ways are the fitness or success measure and whether maximizing it leads a correct predictions.
 pop mean fitness as measure fails due to genetics and due to frequency dependence.
@@
 Stochasticity though due to mutation/drift/etc leads to stationary distribution...
-  \item Extra things
-        \begin{enumerate}
-          \item emphasize the incompleteness of the review. Maybe cite a few ``relevant'' Perc papers?
-        \end{enumerate}
-\end{enumerate}
+- Extra things
+  - emphasize the incompleteness of the review. Maybe cite a few ``relevant'' Perc papers?
 

### 2022-11-15 manuscript.tex
on to conclusions!

@@
 
-\usepackage[displaymath, mathlines]{lineno}
+\usepackage[mathlines]{lineno}
 \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
@@
 
-\title{\scshape Evolutionary dynamics, equilibrium selection, and what population and quantitative genetics have taught evolutionary game theory}
+\title{\scshape Evolutionarily stable strategy analysis and its links to demography and genetics through invasion fitness}
 \author{Jeremy Van Cleve}
@@
 
-  The application of game theory to animal behavior by Maynard Smith and Price invigorated evolutionary biology by suggesting that the complex dynamics among natural selection, genes, and behavior could be analyzed using an evolutionary refinement of the Nash equilibrium, namely the evolutionarily stable strategy (ESS).
-As implied by ``equilibrium'' and ``stable'', an ESS is a predicted long-term outcome that an evolutionary process may reach via natural selection and mutation.
+  Evolutionarily stable strategy (ESS) analysis pioneered by Maynard Smith and Price took off in part because it often does not require explicit assumptions about the genetics and demography of a population in contrast to population genetic models.
+Though this simplicity is useful, it obscures the degree to which ESS analysis applies to in populations with more realistic genetics and demography: for example, how does ESS analysis handle interactions kin selection, group selection, and variable environments when phenotypes are affected by multiple genes?
 
-  In addition to its intuitiveness, ESS analysis makes modeling phenotypic evolution relatively simple as it often does not require the modeler to state their assumptions about the genetics and demography of a population in contrast to explicit population or quantitative genetic models.
-Though this simplicity is useful, it can obscure the degree to which ESS analysis is applicable in populations with more realistic genetic and demography: for example how does ESS analysis handle  interactions with relatives (kin selection), selection across levels (group selection), and variable environments when phenotypes are affected by multiple genes?
-
-  In this paper, I will review the history of the ESS concept and show how early uncertainty about the method lead to deeper theoretical work linking ESS analysis to very general population models through the theory of ``long-term evolution''.
-A crucial tool in long-term evolution is the concept of ``invasion fitness'', which can be calculated for populations with complex genetics and demography and can be used to predict ESSs.
-These ESSs often depend crucially on demographic and genetic parameters.
-Given the increasing abundance of genetic and long-term demographic data in natural populations, this means that ESS analysis will likely be an important tool in understanding these data with evolutionary models.
+  In this paper, I review the history of the ESS concept and show how early uncertainty about the method lead to important mathematical theory linking ESS analysis to general population-genetic models.
+I use this theory to emphasize the link between ESS analysis and the concept of ``invasion fitness''.
+I give examples of how invasion fitness can measure kin selection, group selection, and the evolution of linked modifier genes in response to variable environments.
+The ESSs in these examples depend crucially on demographic and genetic parameters, which highlights how ESS analysis will continue to be an important tool in understanding evolutionary patterns even as models must address the increasing abundance of genetic and long-term demographic data in natural populations.
 
@@
 \noindent {\bfseries Key words}\\
+ESS, lineage fitness, kin selection, group selection, inclusive fitness, reduction principle, variable environments
 
@@
 
-A more recent example involves the evolution of bet-hedging where organisms alternative between different phenotypes with a specific probability or rate when the environmental cycles between different conditions \cite{Seger:Brockmann:1987,Slatkin:1974}.
-These phenotypic shifts can be due to genetic mutations \cite{Leigh:1970,Ishii:Matsuda:1989} or switching between different epigenetic states  \cite{Kussell:Leibler:2005,Salathe:VanCleve:2009} where control of the mutation or switching rate is determined by a ``modifier'' locus separate for the genetic or epigenetic locus controlling the phenotype.
-Due to the the fitness effects of a changing environment, this scenario does not fit neatly into the conditions required for Result \ref{res:twolocus}.
-In fact, an analysis of the external stability of the switching/mutation rate shows that the switching rate that is ES depends on the recombination rate between phenotypic locus and the modifier locus \cite{Liberman:VanCleve:2011}.
-For example, Figure \ref{fig:pip:switching}, shows pairwise invasibility plots for switching rate with different recombination rates and given weakly asymmetric selection in two different environments that alternate every $50$ generations.
-White regions are where mutant alleles with a switching rate on the veritcal axis can invade a population with a resident rate for the rest of the population given by the horizontal axis; black regions are when such a mutant cannot invade.
-The plots show that a rate of $~1/50$ is ES when the recombination rate is zero (similar to older results in \cite{Leigh:1970,Lachmann:Jablonka:1996}).
+One of the important features of Result \ref{res:twolocus} is that shows how under some genetic and demographic scenarios (i.e., a single randomly mating population in a constant environment) an ESS phenotype maybe independent of some genetic parameters like the recombination rate between the loci affecting the phenotype.
+Another example of this kind of result is the ``reduction principle'' of Feldman, Liberman, and colleagues \cite{Feldman:Liberman:1986,Liberman:Feldman:1986,Liberman:Feldman:1986a,Liberman:Feldman:1989,Altenberg:Liberman:2017}, which says that for large, randomly mating populations with viability selection in a constant environment, the only uninvadable or unbeatable rates of recombination, mutation, and migration are the lowest possible rates (typically zero).
+Like Result \ref{res:twolocus}, the reduction principle is proved by analyzing the external stability of an internally stable equilibrium and calculating the invasion fitness of a rare mutant allele.
+However, alleles at the genetic locus where external stability is tested, which is called the modifier locus, do not directly affect fitness and instead modify rates of recombination, mutation, or migration.
+Given the connection between invasion fitness and ESS analysis, the reduction principle shows that rates of zero recombination, mutation, or migration are strict ESSs.
+In effect, the reduction principle says natural selection on populations at equilibrium in constant environments acts to reduce genetic and demographic processes that increase genetic variation; this is intuitive as such genetic variation must results in lower fitness genotypes since the population is already at an equilibrium and the environment is constant.
+Like Result \ref{res:twolocus}, the reduction principle holds independent of the recombination rate between the modifier locus and major loci that affect fitness.
+
+In environments that are not constant, the reduction principle does not hold as genetic variation can be beneficial depending on the environmental state.
+Thus, natural selection may favor genetic and demographic mechanisms that generate genetic variation like positive rates of recombination, mutation, and migration.
+In fact, environmental variation, particularly over time, is one of the most well-studied and important factors supporting the evolution of genotypic and phenotypic variation and the evolution of the mechanisms that promote such variation like recombination \cite{Charlesworth:1976,Sasaki:Iwasa:1987,Otto:Michalakis:1998}, mutation \cite{Leigh:1970,Ishii:Matsuda:1989,Lachmann:Jablonka:1996}, migration \cite{Gillespie:1981,McPeek:Holt:1992,Blanquart:Gandon:2011}, and phenotypic plasticity \cite{Caswell:1983,Via:Lande:1985,Gavrilets:Scheiner:1993,Jong:1995} and bet-hedging \cite{Slatkin:1974,Seger:Brockmann:1987,Kussell:Leibler:2005,Salathe:VanCleve:2009}.
+Recent work shows that not only are the ES recombination, mutation, and migration evolve rates positive in variable environments, but they also depends on genetic parameters like the recombination rate between the major and modifier loci.
+In the case of mutation rate evolution, \citeauthor{Liberman:VanCleve:2011} \cite{Liberman:VanCleve:2011} use an analysis of the external stability of the mutation rate to show that the ES mutation rate depends on the recombination rate between phenotypic locus and the modifier locus.
+For example, Figure \ref{fig:pip:switching}, shows pairwise invasibility plots for switching rate with different recombination rates and given weakly asymmetric selection in two different environments that alternate every $T=50$ generations.
+White regions are where mutant alleles with a switching rate on the vertical axis can invade a population with a resident rate for the rest of the population given by the horizontal axis; black regions are when such a mutant cannot invade.
+The plots show that a rate of $~1/50$ is ES when the recombination rate is zero, similar to older results in \cite{Leigh:1970,Lachmann:Jablonka:1996} that show the ES rate is $~1/T$.
 The ES switching rate decreases as recombination increases until there is an abrupt shift to an ES rate of zero switching.
-In addition to the ES sex ratio and switching rate, there are other examples where the genetic parameters of the modifier locus affect the ES value at that locus including dispersal and recombination in changing environments \cite{Carja:Liberman:2014}.
+\citeauthor{Carja:Liberman:2014} \cite{Carja:Liberman:2014} generalize this approach to the evolution of recombination, mutation, and migration in fluctuating environments.
+They first show that the rate of environmental fluctuation has the same effect on ES recombination and migration rates as it does on ES mutation rates, namely that slower rates of fluctuation lead to slower recombination, mutation, and migration rates.
+In other words, as the rate of environmental change slows, so do the ES rates of genetic and demographic processes that generate genetic variation.
+Ref.
+\cite{Carja:Liberman:2014} also shows that these ES rates all decrease with increasing rates of recombination between the major and modifier loci.
+This can be understood as due to the fact that the strength of indirect selection at the modifier locus decreases as increased recombination erodes linkage disequilibrium between the major and modifier loci.
 
-% \section{Acknowledgements}
+\section{Conclusions and the future of ESS analysis}
+
+When Maynard Smith and Price \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} introduced ESS analysis, the revolutions in molecular genetics and genomics where still two and three decades off, respectively, and evolutionary biology was being roiled by questions and debates regarding the role of natural selection in the evolution of phenotypic novelty broadly \cite{Gould:Lewontin:1979} and in the evolution of animal and human behavior specifically \cite{Lewontin:1977,Hamilton:1977}.
+These debates drew sharp lines between proponents of evolutionary theory derived from explicit genetic and demographic assumptions and theories of phenotypic change based on natural selection alone, namely fitness optimization and ESS analysis.
+Even among those who agreed that ESS analysis was an important approach, questions remained about what measure of fitness to optimize, individual or inclusive fitness, and how to incorporate multilevel selection.
+In the decades following Maynard Smith and Price \cite{Maynard-Smith:Price:1973} however, mathematical analysis bridged some of the division by showing how ESSs can be viewed as the evolutionary attractors of a long-term evolutionary process that also incorporates short-term evolutionary change due to natural selection, mutation, recombination, and other evolutionary forces.
+Mathematical analysis of the long-term process also revealed that invasion fitness as a measure of external stability is the appropriate way to measure evolutionary success and that it can incorporate kin and group selection effects in group structure populations.
+
+\cite{Leimar:2009,Mullon:Keller:2016,Mullon:Keller:2018,Mullon:Lehmann:2019}
+
+\section{Acknowledgements}
+
+J.V.C. would like to acknowledge the hard work and extraordinary patience of the editors and staff who brought this special issue together.
 
@@
 
-%%
-the flow for the intro should be something like: maynard smith setup this false dichotomy: ESS is about ``individual'' level selection when really there are two issues, ESS as tool of optimality and then the application of that tool to kin/group models.
-this requires explaining how ESS is a tool of optimality and how a particular optimality measure, namely lineage growth rate, accounts for kin and group selection since it can captures selection at all levels from the gene on up.
-accounting for other evolutionary forces is more difficult but one way forward uses the fixation probability and allows mutation/demography etc to affect fixation probabilities and hence transition rates.
-
 %%%
@@
 This is notably complementary to methods that study trait coevolution, e.g., Mullon and Lehmann.
-
-%%%
-%%% Fleshed out outline material
-%%%
-
-maybe my argument is then (see below) that population genetic model that include many forces can be converted into long-term evolution models.
-
-So the idea should be something like ESS is really a mathematical formulation of the long-term evolution process which can include additional evolutionary forces other than natural selection.
-
-The argument here is that invasion fitness / lineage fitness and its finite population analogue, fixation probability, are really just measures of evolutionary success that can include any potential evolutionary force.
-The maxima of these measures then yield the stable fixed equilibria of a long-term evolutionary process in a deterministic model or the modes of probability density in a stochastic model.
-Can appeal to physical notions of mean-field (?), potential wells, large deviation theory.
-Also appeal to stochastic stability and equilibrium selection from game theory (which borrows from physics).
-ESS then represents attracting points in phenotypic space and is informative even in the non-zero mutation and finite population size cases.

### 2022-11-16 manuscript.tex
first draft done

@@
 for all possible mutant $A_{\mut}$ that generate mutant phenotype $x_{\mut}$.
-This condition is in fact the condition for a strict ESS from \eqref{eq:ess:def} and can be used to prove that the following result for randomly mating diploid populations with two-locus genetic systems \cite{Eshel:Feldman:1984,Eshel:1996,Eshel:Feldman:1998}:
+This condition is in fact the condition for a strict ESS from \eqref{eq:ess:def} and can be used to prove that the following result for large, randomly mating, diploid populations with two-locus genetic systems \cite{Eshel:Feldman:1984,Eshel:1996,Eshel:Feldman:1998}:
 \begin{result}
@@
 505--506]{Eshel:1996}}.
-For example, the ES sex ratio depends on the genetic structure of the population: in a randomly mating population, the sex ratio is 1:1 when the ratio is determined by an diploid autosomal gene \cite{Fisher:1958,Hamilton:1967,Eshel:Feldman:1982}, but it can take other values when the gene is sex-linked or the genetic system is more complex like in insects that are haplodiploid \cite{Hamilton:1967,Eshel:Feldman:1982a}.
+For example, the ES sex ratio depends on the genetic structure of the population: in a large, randomly mating population, the sex ratio is 1:1 when the ratio is determined by an diploid autosomal gene \cite{Fisher:1958,Hamilton:1967,Eshel:Feldman:1982}, but it can take other values when the gene is sex-linked or the genetic system is more complex like in insects that are haplodiploid \cite{Hamilton:1967,Eshel:Feldman:1982a}.
 The fact that ESSs can incorporate genetic and demographic forces alongside natural selection shows that the ESS method is a more general tool than potentially originally envisioned by Maynard Smith and other proponents of the phentoypic gambit.
@@
 
-One of the important features of Result \ref{res:twolocus} is that shows how under some genetic and demographic scenarios (i.e., a single randomly mating population in a constant environment) an ESS phenotype maybe independent of some genetic parameters like the recombination rate between the loci affecting the phenotype.
+One of the important features of Result \ref{res:twolocus} is that shows how under some genetic and demographic scenarios (i.e., a large, randomly mating population in a constant environment) an ESS phenotype maybe independent of some genetic parameters like the recombination rate between the loci affecting the phenotype.
 Another example of this kind of result is the ``reduction principle'' of Feldman, Liberman, and colleagues \cite{Feldman:Liberman:1986,Liberman:Feldman:1986,Liberman:Feldman:1986a,Liberman:Feldman:1989,Altenberg:Liberman:2017}, which says that for large, randomly mating populations with viability selection in a constant environment, the only uninvadable or unbeatable rates of recombination, mutation, and migration are the lowest possible rates (typically zero).
@@
 Even among those who agreed that ESS analysis was an important approach, questions remained about what measure of fitness to optimize, individual or inclusive fitness, and how to incorporate multilevel selection.
-In the decades following Maynard Smith and Price \cite{Maynard-Smith:Price:1973} however, mathematical analysis bridged some of the division by showing how ESSs can be viewed as the evolutionary attractors of a long-term evolutionary process that also incorporates short-term evolutionary change due to natural selection, mutation, recombination, and other evolutionary forces.
+In the decades following Maynard Smith and Price \cite{Maynard-Smith:Price:1973} however, mathematical analysis bridged some of the division by showing how ESSs can be viewed as the evolutionary attractors of a long-term evolutionary process that builds on short-term evolutionary change due to natural selection, mutation, recombination, and other evolutionary forces.
 Mathematical analysis of the long-term process also revealed that invasion fitness as a measure of external stability is the appropriate way to measure evolutionary success and that it can incorporate kin and group selection effects in group structure populations.
 
-\cite{Leimar:2009,Mullon:Keller:2016,Mullon:Keller:2018,Mullon:Lehmann:2019}
+One of the initial goals of the long-term evolution approach was to determine the degree to which ESS analyses can be independent of the underlying genetics and demography of the population.
+Put another way, how justifiable is the phenotypic gambit?
+Result \ref{res:twolocus} shows that for large and randomly mating populations, ESSs can be fairly independent of genetic parameters like mutation and recombination rates and even the number of loci (for mixed ESSs; see refs.
+\cite{Eshel:Feldman:1998,Eshel:Feldman:2001}).
+At the time, this independence was important for supporting the use of ESS analysis across a broad range of species with different genetics and demography.
+However, this independence may no longer be as useful given the growing abundance of genomic, transcriptomic, and long-term demographic data for natural populations.
+Instead, we might rather want to make use of these data to refine our evolutionary models, and ESS may be a less useful approach if it cannot take advantage of these data.
+However, long-term evolution and consequently ESSs are in fact sensitive to genetics and demography for many traits and populations more complex than Result \ref{res:twolocus} assumes.
+The threshold for such complexity can be quite low as in the case of sex ratio evolution where details of the genetic system can be crucial \cite{Hamilton:1967,Eshel:Feldman:1982a,Wu:1983,Taylor:Jaenike:2002} and in the case of recombination, mutation, and migration rate evolution in variable environments where the ES rates depend on the recombination rate between the major and modifier loci.
+
+All of the ESS analysis so far as concerned a single phenotypic trait.
+However, the principle of invasion fitness can also be applied to multiple evolving traits \cite{Leimar:2009,Mullon:Keller:2016,Mullon:Lehmann:2019}.
+When multiple phenotypes coevolve, issues of pleiotropy and genetic constraint immediately arise as mutations may not cause independent effects along the different phenotypic dimensions.
+Thus, even in the case of two coevolving phenotypes, genetic constraints due to the underlying genetic architecture can have strong effects on long-term evolution and the resulting ESSs.
+The study of coevolving phenotypes is nothing new to quantitative genetics \cite{Lande:1979,Lande:Arnold:1983,Phillips:Arnold:1989} where the important of genetic constrains has long been emphasized.
+In fact, recent work by \citeauthor{Mullon:Lehmann:2019} \cite{Mullon:Lehmann:2019} ties together the quantitative genetic and invasion fitness approaches for coevolving traits.
+By building from an invasion fitness perspective that is comfortable with fitness as a complex function of species ecology and behavior, this synthetic approach shows how pleiotropy, demography, and behavior can interact to shape the coevolution of two synergistic social traits \cite{Mullon:Lehmann:2019}.
+
+Many conceptual divides in evolutionary theory have been bridged in recent decades including divides surrounding ESS analysis and its role in genotypic and phenotypic evolution and in kin and group selection.
+Given these advances, it is hard to be pessmistic about the prospect that evolutionary theory can tackle big open questions about the emergence and persistence of biodiversity.
+Even questions with global consequences like climate change, biodiversity loss, and global conflict may yet be advanced by a more unified evolutionary theory that can handle address the simultaneous action of genetic, phenotypic, social, and cultural change.
 

### 2022-11-16 manuscript.tex
delete notes

@@
 \end{document}
-
-%% Outline/Brainstorm
-
-\section{extra stuff}
-- Intro
-  - Little intro about Maynard Smith and Price
-  - Game theory as general optimization and search for fixed points
-  - Connect back to Hamilton 1964 and goal of optimization with relatedness
-  - Evolutionary process is dynamic but still has ``fixed points''
-  - Evolutionary change can be decomposed and each piece affects location and stability of fixed points
-  - Evol Game theory captures selection but not demography and genetic forces, which can have strong effects.
-- Ok, so thinking more about this, the two main pieces in certain ways are the fitness or success measure and whether maximizing it leads a correct predictions.
-pop mean fitness as measure fails due to genetics and due to frequency dependence.
-hamilton suggested inclusive fitness.
-later authors showed that this is equivalent to an invasion fitness perspective.
-maximization of this leads to ESS.
-Stochasticity though due to mutation/drift/etc leads to stationary distribution...
-- Extra things
-  - emphasize the incompleteness of the review. Maybe cite a few ``relevant'' Perc papers?
-
-%%%
-%%% Detailed outline for post intro
-%%%
-
-What is ESS? ESS is just phenotypes and fitness and an equilibrium and stability condition. This matches pop gen for simple genetics Eshel:1982.
-
-ESS really is a model of a ``long-term evolution'' process of invasion of mutant by resident where fitness determines invasion. This is the ``unbeatable strategy'' of Hamilton.
-
-Long-term evolution depends on a fitness ``measure'' so ``fitness'' isn't just individual fitness.
-Really, we track the frequency or density of individual alleles (i.e., gene's eye view).
-This is lineage growth rate.
-
-Lineage growth rate allows us to measure ESS in complex structure populations with kin and group selection.
-We measure the growth rate of the lineage in all the genetic and demographic environments it finds itself in.
-
-Genetic environments can include other loci linked by recombination. This is the case of the modifier theory for recombination, mutation, migration, etc. Example from Liberman 2011.
-
-Recognizing the importance of genetic parameters in ESS is the opposite of early work to justify ESS which tried to find conditions when ESS predicted same as pop gen but genetic parameters didn't appear.
-
-One way forward to understanding the complex phenotypic evolution including social evolution should involve using ESS theory that incorporates more genetics including multiple loci.
-This is notably complementary to methods that study trait coevolution, e.g., Mullon and Lehmann.

### 2022-11-16 manuscript.tex
slight edits

@@
   Evolutionarily stable strategy (ESS) analysis pioneered by Maynard Smith and Price took off in part because it often does not require explicit assumptions about the genetics and demography of a population in contrast to population genetic models.
-Though this simplicity is useful, it obscures the degree to which ESS analysis applies to in populations with more realistic genetics and demography: for example, how does ESS analysis handle interactions kin selection, group selection, and variable environments when phenotypes are affected by multiple genes?
+Though this simplicity is useful, it obscures the degree to which ESS analysis applies to populations with more realistic genetics and demography: for example, how does ESS analysis handle interactions kin selection, group selection, and variable environments when phenotypes are affected by multiple genes?
 
@@
 \noindent {\bfseries Key words}\\
-ESS, lineage fitness, kin selection, group selection, inclusive fitness, reduction principle, variable environments
+ESS, lineage fitness, kin selection, group selection, reduction principle, variable environments
 
@@
 These debates might be captured in part by the following two questions: \begin{ilnum} \item \label{q:I} How do ESS models that focus on individual fitness capture the effects of kin selection or group selection?
-\item \label{q:II} How does an ESS account for recombination, mutation, genetic drift and other evolutionary forces?
+\item \label{q:II} How does an ESS account for recombination, mutation, migration, and other evolutionary forces?
 \end{ilnum} In this work, I will describe how conceptual advances in evolutionary theory since Maynard Smith and Price \cite{Maynard-Smith:Price:1973} have shed light on both of these questions and have led to a broader understanding of the ESS method and a more integrative framework for understanding how multiple evolutionary forces in complex populations can lead to diverse phenotypes.
@@
 
-The origin of question \ref{q:II} rests in a different set of debates in evolutionary biology regarding the relative role of natural selection vis-a-vis other evolutionary forces, such as mutation, recombination, and gene flow, in explaining organismal phenotypes.
+The origin of question \ref{q:II} rests in a different set of debates in evolutionary biology regarding the relative role of natural selection vis-a-vis other evolutionary forces, such as mutation, recombination, and migration, in explaining organismal phenotypes.
 Early in the 20th century, R.
@@
 The FTNS states that the increase in the mean fitness of a population is equal to the genetic variance in fitness and thus seems to imply that populations always become better adapted to their environments and even that fitness is maximized over the long term.
-It's hard to overstate the importance of the FTNS in shaping the direction of evolutionary theory.
+It is hard to overstate the importance of the FTNS in shaping the direction of evolutionary theory.
 The FTNS came to typify the idea that evolutionary change is dominated by natural selection as a fitness optimizing force.
@@
 Thus, he proposed the ESS method developed by Price and himself as the appropriate tool for predicting phenotypic evolution and particularly for social traits.
-While Maynard Smith acknowledged that ESS models make a number of biological assumptions including that traits have a simple genetic basis (i.e., no dominance, epistasis, etc) and that appropriate genetic variation exits \cite{MaynardSmith:1978}, he believed that the limitations imposed by these assumptions are reasonable.
+While Maynard Smith acknowledged that ESS models make a number of biological assumptions including that traits have a simple genetic basis and that appropriate genetic variation exits \cite{MaynardSmith:1978}, he believed that the limitations imposed by these assumptions are reasonable.
 This view that simplifying away genetic complexities or constraints is generally reasonable was termed the ``phenotypic gambit'' by Alan Grafen \cite{Grafen:1984} and came to dominate evolutionary theory in behavioral and evolutionary ecology \cite{Houston:McNamara:1999}.
@@
 The definition of an ESS is relatively simple and only involves a measure of fitness, a set of phenotypes, and a stability condition.
-Let $\w(x,y)$ be the fitness that an individual with phenotype or strategy $x$ obtains when interacting with an individual with phenotype or strategy $y$.
-The phenotypes can be drawn from a set of discrete or continuous set of values $X$ that represent a single trait like body size or propensity to cooperate.
+Let $\w(x,y)$ be the fitness that an individual with phenotype or strategy $x$ receives when interacting with an individual with phenotype or strategy $y$.
+The phenotypes can be drawn from a set of discrete or continuous set of values that represents a single trait like body size or propensity to cooperate.
 Fitness $\w$ for the moment is simply the survival rate (i.e., viability selection) but as we'll see we can define much more general fitness measures.
@@
 
-Notably, the conditions in \eqref{eq:ess:def} do not reference the genetic basis of the trait $X$ nor do they describe how the frequency or mean value of the trait changes over time.
+Notably, the conditions in \eqref{eq:ess:def} do not reference the genetic basis of the trait $x$ nor do they describe how the frequency or mean value of the trait changes over time.
 Population genetics models, on the other hand, are built to measure the dynamical process of evolution via the combined effects of natural selection and segregation, recombination, mutation, genetic drift, and other evolutionary forces \cite{Crow:Kimura:1970,Ewens:2004}.
@@
 Early work in the simplest haploid asexual population genetic model with discrete strategies showed that the ESSs obtained from condition \eqref{eq:ess:def} must be stable equilibria of the haploid asexual model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980}.
-Specifically, suppose there are two possible phenotypes $1$ and $2$ with the payoff matrix $U = (u_{xy})$ where $u_{xy} = \w(x,y)$ is the payoff or increase to fitness of an individual of type $x$ interacting with an individual of type $y$.
+Specifically, suppose there are two possible phenotypes $x$ and $y$ with the payoff matrix $U = (u_{xy})$ where $u_{xy} = \w(x,y)$ is the payoff or increase to fitness of an individual of type $x$ interacting with an individual of type $y$.
 Then, the ESSs obtained by applying condition \eqref{eq:ess:def} to the payoff matrix $A$ are stable equilibria of a haploid asexual population genetic model with two genotypes, one for each phenotype \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Hofbauer:Sigmund:1998}.
@@
 Importantly, a similar  correspondence between ESSs and population genetic models also applies to a diploid sexual population reproducing in discrete time.
-Suppose there are two possible phenotypes controlled by a single genetic locus with $n$ possible alleles $A_{1},\ldots,A_{n}$ where dominance is possible between alleles and Mendelian segregation breaks up diploid genotypes every generation \cite{Eshel:1982,Lessard:1984}.
-Let diploid genotype $A_{i} A_{j}$ choose phenotype $1$ with probability $x_{ij}$ and phenotype $2$ with probability $1-x_{ij}$.
-If the $\vec{p} = (p_{1}, \ldots, p_{n})$ is the vector of allele frequencies for alleles $A_{1},\ldots,A_{n}$, then the frequency of phenotype $1$ in the population after reproduction is
+Suppose there are two possible phenotypes controlled by a single genetic locus with $m$ possible alleles $A_{1},\ldots,A_{m}$ where dominance is possible between alleles and Mendelian segregation breaks up diploid genotypes every generation \cite{Eshel:1982,Lessard:1984}.
+Let diploid genotype $A_{i} A_{j}$ choose phenotype $x$ with probability $x_{ij}$ and phenotype $y$ with probability $1-x_{ij}$.
+If the $\vec{p} = (p_{1}, \ldots, p_{m})$ is the vector of allele frequencies for alleles $A_{1},\ldots,A_{m}$, then the frequency of phenotype $x$ in the population after reproduction is
 \begin{equation*}
@@
 The invasion of this new allele will lead to its fixation, which is a new internally stable equilibrium that may or may not be externally stable to further invasions depending on the the genotype-phenotype map (e.g., how the genotypes $A_{i} A_{j}$ generate fitness values $w_{ij}$).
-Given a fixed set of genotypes, the evolution of a fixed set of genotypes to an internally stable equilibrium is called ``short-term evolution'' whereas the  process of new mutant invasions that move the population from one internally stable equilibrium to another is called ``long-term evolution'' \cite{Eshel:1991,Eshel:1996,Hammerstein:1996,Weissing:1996,Eshel:Feldman:1998,VanCleve:2015,VanCleve:2020} and eventually leads to an externally stable equilibrium.
+Given a fixed set of genotypes, the evolution of those genotypes to an internally stable equilibrium is called ``short-term evolution'' whereas the process of new mutant invasions that move the population from one internally stable equilibrium to another is called ``long-term evolution'' \cite{Eshel:1991,Eshel:1996,Hammerstein:1996,Weissing:1996,Eshel:Feldman:1998,VanCleve:2015,VanCleve:2020} and eventually leads to an externally stable equilibrium.
 
@@
 Eshel, Feldman, Lessard and others demonstrated that externally stable equilibria are those that generate phenotypes $\ess{x}$ that are ESS \cite{Eshel:Feldman:1984,Liberman:1988,Eshel:1996,Hammerstein:Selten:1994,Hammerstein:1996,Weissing:1996,Eshel:Feldman:1998}.
-Specifically, suppose there are two phenotypes determined by two loci $A$ and $B$ with alleles $A_{1},\ldots,A_{n}$ and $B_{1},\ldots,B_{m}$, respectively, and with recombination rate $0<\rec\le1/2$ between the loci.
+Specifically, suppose there are two alternative phenotypes each determined by two loci $A$ and $B$ with alleles $A_{1},\ldots,A_{m_{A}}$ and $B_{1},\ldots,B_{m_{B}}$, respectively, and with recombination rate $0<\rec\le1/2$ between the loci.
 The vector $\vec{p} = (\ldots,p_{ij},\ldots)$ tracks the frequencies of the chromosome $A_{i} B_{j}$ after selection and recombination.
-Individuals with diploid genotype $A_{i}B_{j}/A_{k}B_{l}$ chooses phenotype 1 with probability $x_{ijkl}$ and phenotype 2 with probability $1 - x_{ijkl}$.
-The frequency of phenotype 1 in the population after reproduction is
+Individuals with diploid genotype $A_{i}B_{j}/A_{k}B_{l}$ chooses phenotype $x$ with probability $x_{ijkl}$ and phenotype $y$ with probability $1 - x_{ijkl}$.
+The frequency of phenotype $x$ in the population after reproduction is
 \begin{equation*}
@@
 Now, assume that a new mutant allele $A_{\mut}$ (increasing the total number of alleles at locus $A$ to $n+1$) arrives at low frequency in a population at an internally stable equilibrium $\fixp{\vec{p}}$ with ``resident'' or wild-type mean phenotype $\fixp{x} = x(\fixp{\vec{p}})$.
-An individual with this new allele chooses phenotype 1 with probability $x_{\mut jkl}$ if its genotype is $A_{\mut}B_{j}/A_{k}B_{l}$ and phenotype 2 with probability $1 - x_{\mut jkl}$.
-A linear stability analysis shows that this allele will either initially increase or decrease in frequency at a geometric rate $\eig$ that is the largest eigenvalue of a Jacobian matrix \cite{Edelstein-Keshet:2005}.
+An individual with this new allele chooses phenotype $x$ with probability $x_{\mut jkl}$ if its genotype is $A_{\mut}B_{j}/A_{k}B_{l}$ and phenotype $y$ with probability $1 - x_{\mut jkl}$.
+A linear stability analysis shows that this allele will either initially increase or decrease in frequency at a geometric rate $\eig$ that is the magnitude of the largest eigenvalue of a Jacobian matrix \cite{Edelstein-Keshet:2005}.
 The eigenvalue $\eig$ has an associated normalized eigenvector $\vec{u}$ that measures the frequency of mutant chromosomes (i.e., chromosomes $A_{\mut} B_{j}$) once the invasion process has stabilized (i.e., has reached stationarity \cite{Caswell:2006}).
@@
 \end{equation}
-is the frequency of phenotype 1 among mutant individuals once the invasion process has stabilized.
+is the frequency of phenotype $x$ among mutant individuals once the invasion process has stabilized.
 Note that since the mutant allele is rare, mutant individuals can have only one mutant chromosome; the other chromosome has no mutant alleles at locus $A$ and remains at the internal equilibrium frequency $\fixp{p}_{kl}$.
@@
   Any phenotype $\fixp{x}$ generated by an externally stable equilibrium is an ESS.
-Moreover, for ESS phenotype $\ess{x}$ generated by an internally stable equilibrium $\fixp{\vec{p}}$, $\fixp{\vec{p}}$ is externally stable.
+Moreover, if an ESS $\ess{x}$ is generated by an internally stable equilibrium $\fixp{\vec{p}}$, $\fixp{\vec{p}}$ is externally stable.
 \end{result}
@@
 While Result \ref{res:twolocus} emphasizes the role of ESSs and phenotypic evolution in the long-term process, the external stability approach used to prove \ref{res:twolocus} reveals a more general viewpoint where ESSs can depend in important ways on the genetic and demographic parameters of a population.
-External stability analysis assumes that the population is at some genetic equilibrium where new mutant alleles can invade at low frequency and looks for a genetic equilibrium and its resulting phenotype that resists invasion by any mutant allele within a set of phenotypic effects.
+External stability analysis assumes that the population is at some genetic equilibrium where new mutant alleles can invade at low frequency and looks for a genetic equilibrium and its resulting phenotype that resists invasion by any mutant allele (within a set of possible phenotypic effects).
 Before the mathematical machinery of external stability was developed, Hamilton appreciated the conceptual importance of uninvadable phenotypes in his study of the sex ratio, and he called sex ratios that resisted invasion ``unbeatable'' phenotypes \cite{Hamilton:1967}.
@@
 The growth rate $\eig$ is often called the ``invasion fitness'' \cite{Metz:Nisbet:1992,Heino:Metz:1998}\footnote{When invasion occurs at more complex attractors like cycles, the ``invasion exponent'' is used, which is $\log |\eig|$ \cite{Rand:Wilson:1994}} and sometimes called the ``lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016,Lehmann:Rousset:2020}.
-In effect, the invasion or lineage fitness measures the number of offspring copies that a mutant allele produces among all the genotypes in the mutant lineage \cite{Lehmann:Mullon:2016}.
+In effect, the invasion or lineage fitness measures the average number of offspring copies that produced by a mutant allele sampled randomly from among all the genotypes in the mutant lineage \cite{Lehmann:Mullon:2016}.
 Assuming that $\eig >0$, the phenotype $\ess{x}$ is unbeatable whenever
@@
 In the general case for an arbitrary genetic and demographic scenario, we assume only that we have the invasion fitness $\eig(y, \ess{x})$ of a rare mutant $y$ in the population that otherwise has mean phenotype $\ess{x}$.
-Starting with the condition for a strict ESS, $w(\ess{x}, \ess{x}) > w(y, \ess{x})$ for $y \ne \ess{x}$ from \eqref{eq:ess:def:a}, we convert that condition to one in terms of ``relative fitness'' by dividing by the mean population fitness $w(\ess{x}, \ess{x})$ to obtain
+Starting with the condition for a strict ESS, $w(\ess{x}, \ess{x}) > w(y, \ess{x})$ for $y \ne \ess{x}$ from \eqref{eq:ess:def:a}, we can convert that condition to one in terms of ``relative fitness'' by dividing by the mean population fitness $w(\ess{x}, \ess{x})$ to obtain
 \begin{equation}
@@
 The strict ESS condition in \eqref{eq:strict:relative} using relative fitness is identical to the condition for uninvadability in condition \eqref{eq:unbeatable} using invasion fitness, which means that unbeatable or uninvadable phentoypes are strict ESSs\footnote{The relationship between invasion fitness and non-strict ESSs is more complex and depends on specific details of the genetics of the population.
-For example, see the section entitled ``Long-term Stability of an ESS'' in \cite{Eshel:Feldman:1998}}.
+For example, see the section entitled ``Long-term Stability of an ESS'' in ref.
+\cite{Eshel:Feldman:1998}}.
 
@@
 505--506]{Eshel:1996}}.
-For example, the ES sex ratio depends on the genetic structure of the population: in a large, randomly mating population, the sex ratio is 1:1 when the ratio is determined by an diploid autosomal gene \cite{Fisher:1958,Hamilton:1967,Eshel:Feldman:1982}, but it can take other values when the gene is sex-linked or the genetic system is more complex like in insects that are haplodiploid \cite{Hamilton:1967,Eshel:Feldman:1982a}.
+For example, the ESS for the sex ratio of a population depends on the genetic structure of the population: in a large, randomly mating population, the sex ratio is 1:1 when the ratio is determined by an diploid autosomal gene \cite{Fisher:1958,Hamilton:1967,Eshel:Feldman:1982}, but it can take other values when the gene is sex-linked or the genetic system is more complex like in insects that are haplodiploid \cite{Hamilton:1967,Eshel:Feldman:1982a}.
 The fact that ESSs can incorporate genetic and demographic forces alongside natural selection shows that the ESS method is a more general tool than potentially originally envisioned by Maynard Smith and other proponents of the phentoypic gambit.
@@
 
-In a populations where individuals live in groups and dispersal is limited, individuals are more genetically related to their groupmates than to individuals in different groups \cite{Wright:1943,Wright:1951,Hamilton:1970}, which means there is a possibility for kin selection to drive the evolution of prosocial phenotypes \cite{Hamilton:1964,Rousset:2004,Lehmann:Rousset:2014}.
+In a populations where individuals live in groups and migration (or dispersal) is limited, individuals are more genetically related to their groupmates than to individuals in different groups \cite{Wright:1943,Wright:1951,Hamilton:1970}, which means there is a possibility for kin selection to drive the evolution of prosocial phenotypes \cite{Hamilton:1964,Rousset:2004,Lehmann:Rousset:2014}.
 Moreover, group structure allows for competition among groups for resources and the possibility of selection among groups favoring more prosocial groups given sufficient group integrity \cite{Hamilton:1975,Wilson:Pollock:1992,Traulsen:Nowak:2006,Gardner:West:2006,Lehmann:Rousset:2010}.
@@
 
-Suppose a population is composed of an infinite number of groups each containing $n$ individuals; groups can be thought as demes or patches each  equally connected to one another at some rate of migration or dispersal (i.e., Wright's island model \cite{Wright:1931}).
-Each individual can belong to one of $\numc$ demographic classes, such as age, sex, or caste, such as workers and queens in social insects.
+Suppose a population is composed of an infinite number of groups each containing $n$ individuals; groups can be thought as demes or patches each  equally connected to one another at some rate of migration (i.e., Wright's island model \cite{Wright:1931}).
+Each individual can belong to one of $\numc$ demographic classes, such as age, sex, or caste (e.g., as workers and queens in social insects).
 Phenotypes are produced by alleles at a single haploid genetic locus.
@@
 The elements $a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x})$ of the projection matrix can be rewritten in terms of an individual (relative) fitness function $\omega_{s's\vec{k}}(\vec{y}, \vec{x})$, which measures the number of class $s'$ mutant offspring in the next generation produced by a class $s$ mutant individual in a group with $\vec{k}$ mutants.
-Relative fitness is the right measure here since we measure offspring production after dispersal, density dependent survival, and all other life-cycle events.
+Relative fitness is the right measure here since we measure offspring production after migration, density dependent survival, and all other life-cycle events.
 The projection matrix and fitness function are then related through the expression for the total number of class $s'$ offspring produced by a mutant individuals in a group with $\vec{k}$ mutants:
@@
   \label{eq:a:w}
-  \sum_{k'} k_{s'} k_{s'}' a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x}) = \sum_{s} k_{s} \omega_{s's\vec{k}}(\vec{y}, \vec{x}) \: .
+  \sum_{k'} k_{s'}' a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x}) = \sum_{s} k_{s} \omega_{s's\vec{k}}(\vec{y}, \vec{x}) \: .
 \end{equation}
@@
 \end{equation}
-where $q_{\vec{k},s}(\vec{y}, \vec{x})$ is the probability (when the invasion process has stabilized) of a randomly sampled mutant allele from the mutant lineage is in a class $s$ individual in a group with $\vec{k}$ mutants and is given by
+where $q_{\vec{k},s}(\vec{y}, \vec{x})$ is the probability (when the invasion process has stabilized) a randomly sampled mutant allele from the mutant lineage is in a class $s$ individual in a group with $\vec{k}$ mutants and is given by
 \begin{equation*}
@@
 \end{equation*}
-The invasion fitness expression in equation \eqref{eq:eig:class} for a group- and class-structured population is conceptually identical to the one for a single population with two-loci in equation \eqref{eq:eig:twolocus}: both expressions show that the mutant successfully invades when mutant alleles more than replace themselves when averaged across all the genetic contexts (i.e., possible alleles at other loci) and demographic contexts (i.e., possible group compositions by mutant allele frequency in each class) in which the mutant lineage occurs during their invasion process.
+The invasion fitness expression in equation \eqref{eq:eig:class} for a group- and class-structured population is conceptually identical to the one for a single population with two-loci in equation \eqref{eq:eig:twolocus}: both expressions show that the mutant successfully invades when mutant alleles more than replace themselves when averaged across all the genetic contexts (i.e., possible alleles at other loci) and demographic contexts (i.e., possible group compositions in terms of mutant allele frequency in each class) in which the mutant lineage occurs during the invasion process.
 
 The invasion fitness in equation \eqref{eq:eig:class} can also be expressed in terms of inclusive fitness in order to isolate the effect on fitness of mutant allele expression in the focal individual (direct effect) and in genetically related individuals (indirect effect).
-There are some important technical subtleties as to how this can be accomplished that we will leave out to keep the presentation simple (see ``Supplement B'' in \cite{Lehmann:Rousset:2020} for details).
-The essence of the method is the same as the one found in many analyses that derive inclusive fitness using the Price equation \cite[e.g.,]{Queller:1992,Frank:1998,Lehmann:Keller:2006,Akcay:VanCleve:2012}: the fitness of a focal individual, here $\omega_{s's\vec{k}}(\vec{y}, \vec{x})$, is decomposed into cost and benefit terms reflecting the marginal effects on fitness\footnote{Marginal effects in this context means additive effects of the phenotype on fitness and often measured with partial regression coefficients.} of expressing the mutant phenotype $y_{s}$ relative to the resident phenotype $x_{s}$.
+There are some important technical subtleties as to how this can be accomplished that we will leave out to keep the presentation simple \cite[see ``Supplement B'' in ref.
+][for details]{Lehmann:Rousset:2020}.
+The essence of the method is the same as the one found in many analyses that derive inclusive fitness using the Price equation \cite[e.g.,][]{Queller:1992,Frank:1998,Lehmann:Keller:2006,Akcay:VanCleve:2012}: the fitness of a focal individual, here $\omega_{s's\vec{k}}(\vec{y}, \vec{x})$, is decomposed into cost and benefit terms reflecting the marginal effects on fitness\footnote{Marginal effects in this context means additive effects of the phenotype on fitness and often measured with partial regression coefficients.} of expressing the mutant phenotype $y_{s}$ relative to the resident phenotype $x_{s}$.
 The cost term $c_{s's}(\vec{y}, \vec{x})$ measures the marginal decrease in class $s'$ offspring produced by a class $s$ individual due to expression of the mutant allele in that individual.
@@
   \eig(\vec{y}, \vec{x}) = 1 + \sum_{s',s} q_{s}(\vec{y}, \vec{x}) v_{s'}(\vec{y}, \vec{x})
-  \left( - c_{s's}(\vec{y}, \vec{x}) + \sum_{\sigma} b_{s'\sigma \leftarrow s}(\vec{y}, \vec{x}) r_{\sigma|s}(\vec{y}, \vec{x}) \right)
+  \left( - c_{s's}(\vec{y}, \vec{x}) + \sum_{\sigma} b_{s'\sigma \leftarrow s}(\vec{y}, \vec{x}) r_{\sigma|s}(\vec{y}, \vec{x}) \right) \: ,
 \end{equation}
-where $r_{\sigma|s}(\vec{y}, \vec{x})$ is genetic relatedness and is the probability that a random groupmate of class $\sigma$ has the mutant allele given an individual of class $s$ has the mutant allele, $v_{s'}(\vec{y}, \vec{x})$ is the normalized reproductive value of class $s'$, and $q_{s}(\vec{y}, \vec{x})$ is the probability that a mutant individual in the invasion process is in class $s$ (see ref.
+which is the inclusive fitness \cite{Hamilton:1964} of the mutant allele.
+In equation \eqref{eq:eig:IF}, $r_{\sigma|s}(\vec{y}, \vec{x})$ is genetic relatedness and is the probability that a random groupmate of class $\sigma$ has the mutant allele given an individual of class $s$ has the mutant allele, $v_{s'}(\vec{y}, \vec{x})$ is the normalized reproductive value of class $s'$, and $q_{s}(\vec{y}, \vec{x})$ is the probability that a mutant individual in the invasion process is in class $s$ (see ref.
 \cite{Lehmann:Rousset:2020} for technical definitions of these quantities).
@@
 Given that previous work using the Price equation has shown that kin selection and group selection partitions of fitness are simply alternative ways of expressing genotype frequency change in a population \cite{Queller:1992,Bijma:Wade:2008,Gardner:West:2007,Marshall:2011,Akcay:VanCleve:2012}, it should be possible to rewrite invasion fitness so as to group terms that constitute between-group selection, $\omega_{G,s's}(\vec{y}, \vec{x})$, and those that constitute within-group selection, $\omega_{\Delta G,s's}(\vec{y}, \vec{x})$.
-In fact, we can do this by applying a ``neighbor-modulated'' approach from refs.
+In fact, if we apply a ``neighbor-modulated'' approach from refs.
 \cite{Queller:1992,Akcay:VanCleve:2012} to the results from ref.
-\cite{Lehmann:Rousset:2020} and obtaining
+\cite{Lehmann:Rousset:2020}, we obtain
 \begin{equation}
@@
 We can see immediately that if mutant allele expression is costly to the individual itself, $c_{s's}>0$, helps others, $b_{s's \leftarrow \sigma}>0$ for all classes $\sigma$, and the benefits outweigh the costs within a class, $b_{s's \leftarrow s} - c_{s's}>0$, then the mutant allele expression is positively selected by between-group selection and negatively selected by within-group selection.
-Since both the inclusive fitness and group selection partitions of invasion fitness (equations \eqref{eq:eig:IF} and \eqref{eq:eig:GS}, respectively) are expressed using the same benefit, cost, and relatedness functions, a singular ESS analysis using invasion can answer questions derived both both perspectives.
+Since both the kin selection and group selection partitions of invasion fitness (equations \eqref{eq:eig:IF} and \eqref{eq:eig:GS}, respectively) are expressed using the same benefit, cost, and relatedness functions, a single ESS analysis using invasion can answer questions derived both both perspectives.
 
@@
   \end{tabular}
-  \caption{Pairwise invasibility plots for the swithing/mutation rate model of \cite{Liberman:VanCleve:2011} when alternation between each of the two environments occurs every $n=50$ generations.
+  \caption{Pairwise invasibility plots for the mutation rate model of \cite{Liberman:VanCleve:2011} when alternation between each of the two environments occurs every $n=50$ generations.
 The fitness costs of being in the wrong environments are $s_{1} = 0.01$ and $s_{2} = 0.015$ and the recombination rate between the phenotypic locus and the modifier is given above each plot.
@@
 Another example of this kind of result is the ``reduction principle'' of Feldman, Liberman, and colleagues \cite{Feldman:Liberman:1986,Liberman:Feldman:1986,Liberman:Feldman:1986a,Liberman:Feldman:1989,Altenberg:Liberman:2017}, which says that for large, randomly mating populations with viability selection in a constant environment, the only uninvadable or unbeatable rates of recombination, mutation, and migration are the lowest possible rates (typically zero).
-Like Result \ref{res:twolocus}, the reduction principle is proved by analyzing the external stability of an internally stable equilibrium and calculating the invasion fitness of a rare mutant allele.
+Like Result \ref{res:twolocus}, the reduction principle is proved by analyzing the external stability of an internally-stable equilibrium and calculating the invasion fitness of a rare mutant allele.
 However, alleles at the genetic locus where external stability is tested, which is called the modifier locus, do not directly affect fitness and instead modify rates of recombination, mutation, or migration.
@@
 Thus, natural selection may favor genetic and demographic mechanisms that generate genetic variation like positive rates of recombination, mutation, and migration.
-In fact, environmental variation, particularly over time, is one of the most well-studied and important factors supporting the evolution of genotypic and phenotypic variation and the evolution of the mechanisms that promote such variation like recombination \cite{Charlesworth:1976,Sasaki:Iwasa:1987,Otto:Michalakis:1998}, mutation \cite{Leigh:1970,Ishii:Matsuda:1989,Lachmann:Jablonka:1996}, migration \cite{Gillespie:1981,McPeek:Holt:1992,Blanquart:Gandon:2011}, and phenotypic plasticity \cite{Caswell:1983,Via:Lande:1985,Gavrilets:Scheiner:1993,Jong:1995} and bet-hedging \cite{Slatkin:1974,Seger:Brockmann:1987,Kussell:Leibler:2005,Salathe:VanCleve:2009}.
+In fact, environmental variation, particularly over time, is one of the most well-studied and important factors supporting the evolution of genotypic and phenotypic variation and the evolution of the mechanisms that promote such variation like recombination \cite{Charlesworth:1976,Sasaki:Iwasa:1987,Otto:Michalakis:1998}, mutation \cite{Leigh:1970,Ishii:Matsuda:1989,Lachmann:Jablonka:1996}, migration \cite{Gillespie:1981,McPeek:Holt:1992,Blanquart:Gandon:2011}, phenotypic plasticity \cite{Caswell:1983,Via:Lande:1985,Gavrilets:Scheiner:1993,Jong:1995} and bet-hedging \cite{Slatkin:1974,Seger:Brockmann:1987,Kussell:Leibler:2005,Salathe:VanCleve:2009}.
 Recent work shows that not only are the ES recombination, mutation, and migration evolve rates positive in variable environments, but they also depends on genetic parameters like the recombination rate between the major and modifier loci.
 In the case of mutation rate evolution, \citeauthor{Liberman:VanCleve:2011} \cite{Liberman:VanCleve:2011} use an analysis of the external stability of the mutation rate to show that the ES mutation rate depends on the recombination rate between phenotypic locus and the modifier locus.
-For example, Figure \ref{fig:pip:switching}, shows pairwise invasibility plots for switching rate with different recombination rates and given weakly asymmetric selection in two different environments that alternate every $T=50$ generations.
-White regions are where mutant alleles with a switching rate on the vertical axis can invade a population with a resident rate for the rest of the population given by the horizontal axis; black regions are when such a mutant cannot invade.
+For example, Figure \ref{fig:pip:switching}, shows pairwise invasibility plots for the mutation rate with different recombination rates and given weakly asymmetric selection in two different environments that alternate every $T=50$ generations.
+White regions are where mutant alleles with a switching rate on the vertical axis can invade a population with a resident rate given by the horizontal axis; black regions are when such a mutant cannot invade.
 The plots show that a rate of $~1/50$ is ES when the recombination rate is zero, similar to older results in \cite{Leigh:1970,Lachmann:Jablonka:1996} that show the ES rate is $~1/T$.
@@
 In other words, as the rate of environmental change slows, so do the ES rates of genetic and demographic processes that generate genetic variation.
-Ref.
-\cite{Carja:Liberman:2014} also shows that these ES rates all decrease with increasing rates of recombination between the major and modifier loci.
+\citeauthor{Carja:Liberman:2014} also show that these ES rates all decrease with increasing rates of recombination between the major and modifier loci.
 This can be understood as due to the fact that the strength of indirect selection at the modifier locus decreases as increased recombination erodes linkage disequilibrium between the major and modifier loci.
@@
 
-When Maynard Smith and Price \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} introduced ESS analysis, the revolutions in molecular genetics and genomics where still two and three decades off, respectively, and evolutionary biology was being roiled by questions and debates regarding the role of natural selection in the evolution of phenotypic novelty broadly \cite{Gould:Lewontin:1979} and in the evolution of animal and human behavior specifically \cite{Lewontin:1977,Hamilton:1977}.
+When Maynard Smith and Price \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} introduced ESS analysis, the revolutions in molecular genetics and genomics were still two and three decades off, respectively, and evolutionary biology was being roiled by questions and debates regarding the role of natural selection in the evolution of phenotypic novelty broadly \cite{Gould:Lewontin:1979} and in the evolution of animal and human behavior specifically \cite{Lewontin:1977,Hamilton:1977}.
 These debates drew sharp lines between proponents of evolutionary theory derived from explicit genetic and demographic assumptions and theories of phenotypic change based on natural selection alone, namely fitness optimization and ESS analysis.
@@
 In the decades following Maynard Smith and Price \cite{Maynard-Smith:Price:1973} however, mathematical analysis bridged some of the division by showing how ESSs can be viewed as the evolutionary attractors of a long-term evolutionary process that builds on short-term evolutionary change due to natural selection, mutation, recombination, and other evolutionary forces.
-Mathematical analysis of the long-term process also revealed that invasion fitness as a measure of external stability is the appropriate way to measure evolutionary success and that it can incorporate kin and group selection effects in group structure populations.
+Mathematical analysis of the long-term process also revealed that invasion fitness as a measure of external stability is the appropriate way to measure evolutionary success and that it can incorporate kin and group selection effects in group structured populations.
 
-One of the initial goals of the long-term evolution approach was to determine the degree to which ESS analyses can be independent of the underlying genetics and demography of the population.
+One of the initial goals of the long-term evolution approach was to determine the degree to which ESS analyses can be independent of the underlying genetics and demography of populations.
 Put another way, how justifiable is the phenotypic gambit?
@@
 However, long-term evolution and consequently ESSs are in fact sensitive to genetics and demography for many traits and populations more complex than Result \ref{res:twolocus} assumes.
-The threshold for such complexity can be quite low as in the case of sex ratio evolution where details of the genetic system can be crucial \cite{Hamilton:1967,Eshel:Feldman:1982a,Wu:1983,Taylor:Jaenike:2002} and in the case of recombination, mutation, and migration rate evolution in variable environments where the ES rates depend on the recombination rate between the major and modifier loci.
+The threshold for such complexity can be quite low as in the case of sex ratio evolution where details of the genetic system can be crucial \cite{Hamilton:1967,Eshel:Feldman:1982a,Wu:1983,Taylor:Jaenike:2002}, in the case of kin and group structured populations where demography determines the fate of genes with social effects, and in the case of recombination, mutation, and migration rate evolution in variable environments where the ES rates depend on the recombination rate between the major and modifier loci.
 
@@
 Given these advances, it is hard to be pessmistic about the prospect that evolutionary theory can tackle big open questions about the emergence and persistence of biodiversity.
-Even questions with global consequences like climate change, biodiversity loss, and global conflict may yet be advanced by a more unified evolutionary theory that can handle address the simultaneous action of genetic, phenotypic, social, and cultural change.
+Even questions with global consequences like climate change, biodiversity loss, and global conflict may yet be advanced by a more unified evolutionary theory that can address the simultaneous action of genetic, phenotypic, social, and cultural change.
 

### 2023-01-11 manuscript.tex
typo
making it through rev 1 comments

@@
 
-\usepackage[style=prsb,natbib=true,uniquename=false,uniquelist=false,sortcites=false]{biblatex}
+\usepackage[style=prsb,natbib=true,uniquename=false,uniquelist=false,sortcites=false,sorting=oldnone]{biblatex}
 \addbibresource{refs.bib}
+
+\DeclareSortingTemplate{oldnone}{
+  \sort{\citeorder}
+}
 
@@
 \setmathfont[range=up/{num,latin,Latin,greek,Greek}]{Minion Pro}
-\setmathfont[range=it/{latin,Latin,greek,Greek}]{Minion Pro It}
-\setmathfont[range=bfup/{latin,Latin,greek,Greek}]{Minion Pro Bold}
-\setmathfont[range=bfit/{latin,Latin,greek,Greek}]{Minion Pro Bold It}
+\setmathfont[range=it/{latin,Latin,greek,Greek}]{MinionPro-It}
+\setmathfont[range=bfup/{latin,Latin,greek,Greek}]{MinionPro-Bold}
+\setmathfont[range=bfit/{latin,Latin,greek,Greek}]{MinionPro-BoldIt}
 \setmathfont[range=\mathcal]{Asana Math}
@@
   Evolutionarily stable strategy (ESS) analysis pioneered by Maynard Smith and Price took off in part because it often does not require explicit assumptions about the genetics and demography of a population in contrast to population genetic models.
-Though this simplicity is useful, it obscures the degree to which ESS analysis applies to populations with more realistic genetics and demography: for example, how does ESS analysis handle interactions kin selection, group selection, and variable environments when phenotypes are affected by multiple genes?
+Though this simplicity is useful, it obscures the degree to which ESS analysis applies to populations with more realistic genetics and demography: for example, how does ESS analysis handle complexities such as kin selection, group selection, and variable environments when phenotypes are affected by multiple genes?
 
@@
 Maynard Smith explained the ESS concept in more detail in an important paper in 1974 \cite{Maynard-Smith:1974}, introduced the famous ``Hawk-Dove'' game in his study of asymmetric games in 1976 \cite{Maynard-Smith:Parker:1976}, and summarized the nascent field of evolutionary game theory in his now classic 1982 book \cite{MaynardSmith:1982}.
-In the 50 years since the 1973 paper, evolutionary game theory has become an essential tool in evolutionary and behavioral ecology with rich theoretical work that delves into foundational evolutionary and game-theoretic concepts and mathematical and simulation models that predict ESSs for almost innumerable biological systems.
-Evolutionary game theory has also strongly influenced the social sciences including, naturally, economics, political science, psychology, and anthropology and has drawn applied mathematicians, computer scientists and physicists into mathematical biology.
+In the 50 years since the 1973 paper, evolutionary game theory has become an essential tool in evolutionary and behavioral ecology and includes rich theoretical work that delves into foundational evolutionary and game-theoretic concepts and mathematical and simulation models that predict ESSs for many specific biological systems.
+Evolutionary game theory has also strongly influenced the social sciences including, economics, political science, psychology, and anthropology and has drawn applied mathematicians, computer scientists and physicists into mathematical biology.
 
-The role of evolutionary game theory and the ESS method specifically in evolutionary biology has been at the crux of some of the most important conceptual debates in the field including the role of natural selection and adaptation vis-a-vis other forces \cite[e.g.,][]{MaynardSmith:1978,Gould:Lewontin:1979,Lewontin:1979,Orzack:Sober:1994,Gardner:2017,Kern:Hahn:2018,Jensen:Payseur:2019} and the importance of kin and group selection relative to individual selection
-\cite[e.g.,][]{Maynard-Smith:1964,Hamilton:1963,Price:1972:cov,Wilson:Wilson:2007,Leigh:2010,Akcay:VanCleve:2012,West:Griffin:2007,Gardner:Grafen:2009,Nowak:Tarnita:2010,Abbot:Abe:2011,Allen:Nowak:2013,Birch:2014,Birch:2017,Nowak:McAvoy:2017}.
+The role of evolutionary game theory and the ESS method specifically in evolutionary biology has been at the crux of some of the most important conceptual debates in the field including the role of natural selection and adaptation vis-a-vis other evolutionary forces \cite[e.g.,][]{MaynardSmith:1978,Gould:Lewontin:1979,Lewontin:1979,Orzack:Sober:1994,Gardner:2017,Kern:Hahn:2018,Jensen:Payseur:2019} and the importance of kin and group selection relative to individual selection
+\cite[e.g.,][]{Hamilton:1963,Maynard-Smith:1964,Price:1972:cov,Wilson:Wilson:2007,Leigh:2010,Akcay:VanCleve:2012,West:Griffin:2007,Gardner:Grafen:2009,Nowak:Tarnita:2010,Abbot:Abe:2011,Allen:Nowak:2013,Birch:2014,Birch:2017,Nowak:McAvoy:2017}.
 These debates might be captured in part by the following two questions: \begin{ilnum} \item \label{q:I} How do ESS models that focus on individual fitness capture the effects of kin selection or group selection?
-\item \label{q:II} How does an ESS account for recombination, mutation, migration, and other evolutionary forces?
+\item \label{q:II} How do ESS models account for recombination, mutation, migration, and other evolutionary forces?
 \end{ilnum} In this work, I will describe how conceptual advances in evolutionary theory since Maynard Smith and Price \cite{Maynard-Smith:Price:1973} have shed light on both of these questions and have led to a broader understanding of the ESS method and a more integrative framework for understanding how multiple evolutionary forces in complex populations can lead to diverse phenotypes.
@@
 210]{Maynard-Smith:1974} as might be the case for kin selection \cite{Hamilton:1964}.
-In setting up this dichotomy, Maynard Smith implies that the ESS method may not apply when kin or group selection are involved and that kin or group selection models might lead to different results when applied to the same biological scenario.
+In setting up this dichotomy, Maynard Smith implied that the ESS method may not apply when kin or group selection are involved.
 As I will describe below, evolutionary theorists have since realized (including Maynard Smith himself \cite[p.
@@
 The levels of selection question became a major topic of study in evolutionary theory \cite[e.g.,][]{Lewontin:1970,Dawkins:1982,Wilson:Sober:1989,Maynard-Smith:Szathmary:1995,Wilson:1997,Michod:1999,Michod:2006,Szathmary:2015} and philosophy of biology \cite{Hull:1980,Brandon:1982,Damuth:Heisler:1988,Lloyd:1992,Lloyd:1994,Sober:Wilson:1994,Okasha:2006,Okasha:2016} and continues to generate substantial research \cite[e.g.,][]{Black:Bourrat:2020,Cooney:Mori:2022,Veit:2022}.
-For the purpose of explaining how an ESS can capture selection at multiple levels and among relatives or kin, I will argue below that the right measure of utility is the ``invasion fitness'' \cite{Metz:Nisbet:1992,Heino:Metz:1998} or ```lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016} of a mutant allele at a single genetic locus, which can be shown to be functionally equivalent to a measure of inclusive fitness \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020}.
+For the purpose of explaining how an ESS can capture selection at multiple levels and among relatives or kin, I will argue below that the right measure of utility is the ``invasion fitness'' \cite{Metz:Nisbet:1992,Heino:Metz:1998} or ```lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016} of a mutant allele at a single genetic locus, which can be shown to be functionally equivalent to a measure of inclusive fitness \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020}, which is a concept originally proposed by W.D.
+Hamilton in relation to kin selection \cite{Hamilton:1964}.
 
@@
 Fisher's ``fundamental theorem of natural selection'' (FTNS) \cite{Fisher:1930} established a mathematical expression of the importance of natural selection.
-The FTNS states that the increase in the mean fitness of a population is equal to the genetic variance in fitness and thus seems to imply that populations always become better adapted to their environments and even that fitness is maximized over the long term.
+The FTNS states that the increase in the mean fitness of a population is equal to the genetic variance in fitness and thus seems to imply that populations always become better adapted to their environments (since genetic variance is always non-negative) and even that fitness is maximized over the long term.
 It is hard to overstate the importance of the FTNS in shaping the direction of evolutionary theory.
 The FTNS came to typify the idea that evolutionary change is dominated by natural selection as a fitness optimizing force.
-Bill Hamilton appealed to this idea in his original paper on kin selection \cite{Hamilton:1964} and theorized that ``inclusive fitness'' is the fitness quantity that is maximized.
+Hamilton appealed to this idea in his original paper on kin selection \cite{Hamilton:1964} and theorized that inclusive fitness is the fitness quantity that is maximized.
 
-Subsequent work by population geneticists revealed however that mean fitness can decrease due to frequency-dependent selection \cite{Wright:1955,Lewontin:1970} or recombination among multiple genetic loci \cite{Kojima:Kelleher:1961,Moran:1964,Karlin:1975,Akin:1979}.
+Subsequent work by population geneticists revealed the weakness of this understanding of the FTNS by showing that mean fitness can decrease due to frequency-dependent selection \cite{Wright:1955,Lewontin:1970} or recombination among multiple genetic loci \cite{Kojima:Kelleher:1961,Moran:1964,Karlin:1975,Akin:1979}.
 Further, studies in the 1960s of the rates of molecular evolution \cite{Zuckerkandl:Pauling:1965,King:Jukes:1969} and levels of polymorphism \cite{Harris:1966,Lewontin:Hubby:1966} in a number of species spurred the development of the neutral \cite{Kimura:1968,Kimura:1983:book} and nearly neutral \cite{Ohta:1974,Ohta:1992} theories \cite{Ohta:Gillespie:1996} of molecular evolution.
-These theories posited that many mutations are weakly affected by natural selection, and thus their fate is governed mostly by genetic drift \cite{Kimura:1983:book}.
-Consequently, by the 1970s, evolutionary biologists were heavily debating both the relative role of natural selection versus other forces in shaping evolutionary patterns \cite{Gillespie:Langley:1974,Gillespie:1978}, and some biologists including Lewontin specifically questioned the importance of fitness maximization \cite{Karlin:1975,Gould:Lewontin:1979}.
+These theories posited that many mutations are only weakly affected by natural selection, and thus their fate is governed mostly by genetic drift \cite{Kimura:1983:book}.
+Consequently, by the 1970s, evolutionary biologists were heavily debating the relative role of natural selection versus other forces in shaping evolutionary patterns \cite{Gillespie:Langley:1974,Gillespie:1978}, and some biologists including Lewontin specifically questioned the importance of fitness maximization \cite{Karlin:1975,Gould:Lewontin:1979}.
 Maynard Smith was a clear proponent of fitness maximization and viewed it natural to ``[assume] that evolution has occurred by natural selection'' \cite[p.
 31]{MaynardSmith:1978}.
-Thus, he proposed the ESS method developed by Price and himself as the appropriate tool for predicting phenotypic evolution and particularly for social traits.
-While Maynard Smith acknowledged that ESS models make a number of biological assumptions including that traits have a simple genetic basis and that appropriate genetic variation exits \cite{MaynardSmith:1978}, he believed that the limitations imposed by these assumptions are reasonable.
+He proposed the ESS method developed by Price and himself as the appropriate tool for predicting phenotypic evolution and particularly for social traits.
+While Maynard Smith acknowledged that ESS models make a number of biological assumptions including that traits have a simple genetic basis and that appropriate genetic variation exits \cite{MaynardSmith:1978}, he believed that the limitations imposed by these assumptions are biologically reasonable.
 This view that simplifying away genetic complexities or constraints is generally reasonable was termed the ``phenotypic gambit'' by Alan Grafen \cite{Grafen:1984} and came to dominate evolutionary theory in behavioral and evolutionary ecology \cite{Houston:McNamara:1999}.
@@
 
-The definition of an ESS is relatively simple and only involves a measure of fitness, a set of phenotypes, and a stability condition.
+The definition of an ESS is relatively simple and only involves a set of phenotypes, a measure of fitness, and a stability condition.
+Suppose that two individuals interact and the phenotypes that characterize their behavior in the interactions are measured with the variables $x$ and $y$.
+The phenotypes can be drawn from a set of discrete or continuous set of values that represents a single trait like body size or propensity to cooperate.
+For the purpose of explaining the theory more concretely, assume that individuals can choose one of two possible behaviors in the interaction, $C$ and $D$.
+The phenotypes $x$ and $y$ measure the probability that individuals in an interaction choose behavior $C$, and $1-x$ and $1-y$ are the probabilities those individuals choose behavior $D$ instead (probabilistic phenotypes are called mixed strategies in game theory \cite{Osborne:Rubinstein:1994}).
+For example, the behavior $C$ could be cooperation in the interaction and $D$ could be a lack of cooperation.
 Let $\w(x,y)$ be the fitness that an individual with phenotype or strategy $x$ receives when interacting with an individual with phenotype or strategy $y$.
-The phenotypes can be drawn from a set of discrete or continuous set of values that represents a single trait like body size or propensity to cooperate.
 Fitness $\w$ for the moment is simply the survival rate (i.e., viability selection) but as we'll see we can define much more general fitness measures.
@@
 A natural question then is how the long-run behavior of a population genetic model (e.g., its equilibrium genotypes and induced phenotypes) compares to the equilibrium phenotypes obtained from an analogous ESS model.
-This question was tackled by mathematical population biologists beginning in the late 1970s and through 1980s and 1990s \cite[e..g,][]{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Eshel:1982,Hofbauer:Schuster:1982,Lessard:1984,Cressman:1988,Cressman:Hines:1984,Cressman:Hofbauer:1996,Hammerstein:1996,Weissing:1996,Eshel:1996,Eshel:Feldman:1984}.
-Early work in the simplest haploid asexual population genetic model with discrete strategies showed that the ESSs obtained from condition \eqref{eq:ess:def} must be stable equilibria of the haploid asexual model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980}.
-Specifically, suppose there are two possible phenotypes $x$ and $y$ with the payoff matrix $U = (u_{xy})$ where $u_{xy} = \w(x,y)$ is the payoff or increase to fitness of an individual of type $x$ interacting with an individual of type $y$.
-Then, the ESSs obtained by applying condition \eqref{eq:ess:def} to the payoff matrix $A$ are stable equilibria of a haploid asexual population genetic model with two genotypes, one for each phenotype \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Hofbauer:Sigmund:1998}.
-This result isn't particularly surprising since natural selection is the only evolutionary force acting in the haploid asexual model.
-Importantly, a similar  correspondence between ESSs and population genetic models also applies to a diploid sexual population reproducing in discrete time.
-Suppose there are two possible phenotypes controlled by a single genetic locus with $m$ possible alleles $A_{1},\ldots,A_{m}$ where dominance is possible between alleles and Mendelian segregation breaks up diploid genotypes every generation \cite{Eshel:1982,Lessard:1984}.
-Let diploid genotype $A_{i} A_{j}$ choose phenotype $x$ with probability $x_{ij}$ and phenotype $y$ with probability $1-x_{ij}$.
-If the $\vec{p} = (p_{1}, \ldots, p_{m})$ is the vector of allele frequencies for alleles $A_{1},\ldots,A_{m}$, then the frequency of phenotype $x$ in the population after reproduction is
+This question was tackled by mathematical population biologists beginning in the late 1970s and through 1980s and 1990s \cite[e..g,][]{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Eshel:1982,Hofbauer:Schuster:1982,Cressman:Hines:1984,Eshel:Feldman:1984,Lessard:1984,Cressman:1988,Cressman:Hofbauer:1996,Eshel:1996,Hammerstein:1996,Weissing:1996}.
+
+Early mathematical analyses began by using the simplest population genetic model, which is one that assumes a \textit{haploid asexual} population with two genotypes, one with phenotype $x=1$ that chooses $C$ with certainty and one that with phentoype $x=0$ that chooses $D$ with certainty.
+There are four possible interacting pairs of phenotypes in this scenario yielding four fitnesses that can be collected into a so-called payoff matrix $W$:
+\begin{equation*}
+  W =
+  \begin{pmatrix}
+    \w(1,1) & \w(1,0) \\
+    \w(0,1) & \w(0,0)
+  \end{pmatrix} \, .
+\end{equation*}
+Suppose that $p$ measures the frequency of the phenotype $x=1$ that always chooses $C$.
+The equilibria of this model are either fixation of one of the two genotype resulting a population of individuals that always chooses $C$, $p=1$, or always choose $D$, $p=0$ or a polymorphism of both genotypes, $0 < p < 1$, resulting in a population where some choose $C$ and some choose $D$.
+The analogous ESS model of this scenario applies condition \eqref{eq:ess:def} to the four possible fitness values in the payoff matrix $W$ and determines which phenotype, $x=1$ or $x=0$, is ESS or whether a mixed strategy (i.e., $0 < x < 1$) is ESS.
+These early analyses showed that the values of $x$ that are an ESS are also values of $p$ that produce stable equilibria in the haploid asexual population genetic model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Hofbauer:Sigmund:1998}.
+
+This correspondence between the population genetic and ESS models isn't particularly surprising since natural selection is the only evolutionary force acting in the haploid asexual model.
+However, a similar correspondence between ESSs and population genetic models also applies to a \textit{diploid sexual} population reproducing in discrete time where Mendelian segregation breaks up diploid genotypes every generation \cite{Eshel:1982,Lessard:1984}.
+For this scenario, the phenotype is controlled by a single genetic locus with $a$ possible alleles $A_{1},\ldots,A_{a}$.
+Diploid genotype $A_{i} A_{j}$ cooperates with probability $x_{ij}$ and does not cooperate with probability $1-x_{ij}$.
+If there are no restrictions on the $x_{ij}$, then dominance is possible between the alleles.
+Let $\vec{p} = (p_{1}, \ldots, p_{a})$ be the vector of allele frequencies for alleles $A_{1},\ldots,A_{a}$.
+Given allele frequencies $\vec{p}$ in the parental generation and assuming random mating, the mean phenotype or probability of cooperation in the population after reproduction is
 \begin{equation*}
@@
 \end{equation*}
-Eshel and Lessard showed \cite{Eshel:1982,Lessard:1984} that ESSs $\ess{x}$ of the payoff matrix $A$ are stably maintained in the diploid population model when those ESSs are produced by some vector of allele frequencies $\ess{\vec{p}}$ such that $\ess{x} = x(\ess{\vec{p}})$.
+Eshel and Lessard showed \cite{Eshel:1982,Lessard:1984} that ESSs $\ess{x}$ of the payoff matrix $W$ are stably maintained in the diploid population model when those ESSs are produced by some vector of allele frequencies $\ess{\vec{p}}$ such that $\ess{x} = x(\ess{\vec{p}})$.
 This correspondence also holds for continuous-time diploid models \cite{Sigmund:1987,Cressman:1988,Cressman:Hofbauer:1996,Hofbauer:Sigmund:1998} and for models with multiple loci when the probability of a genotype choosing a phenotype is an additive function of effects from each locus (i.e., additive non-epistasis) \cite{Eshel:1982}.
@@
 An important limitation of the above results connecting ESS phenotypes $\ess{x}$ to equilibria in diploid single and multilocus population genetic models is that the genotype-phenotype map may not produce phenotypes that are ESS and thus those ESS phenotypes are unattainable in the short-term.
-Specifically, the $x_{ij}$ may have values so that no population composition of genotypes $A_{i} A_{j}$ yields the ESS phenotype $\ess{x}$.
-Moreover, as mentioned above, multilocus systems more generally (i.e., systems with non-additive interactions or epistasis) do not guarantee that genotypes with higher marginal fitness\footnote{The marginal fitness of a genotype or allele is the mean fitness of that genotype allele among all contexts in which it occurs in the population.
+Specifically, the $x_{ij}$ may have values so that no population composition $\vec{p}$ of alleles $A_{i}$ can produce the ESS phenotype $\ess{x}$.
+Moreover, as mentioned above, multilocus systems more generally (i.e., systems with non-additive interactions or epistasis) do not guarantee that genotypes with higher marginal fitness\footnote{The marginal fitness of a genotype or allele is the mean fitness of that genotype or allele among all contexts in which it occurs in the population.
 For example, the marginal fitness of one allele at a single diploid locus in the mean fitness of that allele in combination with all other alleles according to their frequencies.} increase in frequency, which is a requirement for the correspondence between ESS and population genetic equilibria described above \cite{Eshel:1982,Hofbauer:Sigmund:1998}.
@@
 
-Intriguingly, Eshel, Feldman, and others discovered that these constraints might not be so important on a longer term evolutionary timescales.
-The important idea is that the conditions under which an genetic equilibrium is stable given a fixed set of alleles or genotypes does not determine whether a new allele or genotype outside of that fixed set might invade the population when rare.
-If an equilibrium is stable given the fixed set of alleles or genotypes, it is called ``internally stable'' \cite{Eshel:1996}.
+Intriguingly, Eshel, Feldman, and others discovered that these constraints might not be so important on longer term evolutionary timescales.
+The important idea is that the conditions under which a genetic equilibrium is stable given a fixed set of alleles or genotypes does not determine whether a new and rare allele or genotype outside of that fixed set might invade the population.
+When an equilibrium is stable given a fixed set of alleles or genotypes, it is called ``internally stable'' \cite{Eshel:1996}.
 If that equilibrium also resists invasion by new alleles or genotypes outside the fixed set, it is called ``externally stable'' \cite{Liberman:1988,Eshel:1996}.
@@
 For example, suppose there is heterozygote advantage in a population with two alleles at a single diploid locus.
-The internally stable equilibrium is a polymorphism of both alleles.
+The internally stable equilibrium is a polymorphism where both alleles are present in positive frequency.
 However, a new allele can invade the population if it generates the original heterozygote fitness in a heterozygote with either original allele and in a homozygote with itself.
 Hence, the internally stable polymorphism is not externally stable.
-The invasion of this new allele will lead to its fixation, which is a new internally stable equilibrium that may or may not be externally stable to further invasions depending on the the genotype-phenotype map (e.g., how the genotypes $A_{i} A_{j}$ generate fitness values $w_{ij}$).
+The invasion of this new allele will lead to its fixation, which is a new internally stable equilibrium that may or may not be externally stable to further invasions depending on the the genotype-phenotype map (i.e., how the genotypes $A_{i} A_{j}$ generate fitness values $w_{ij}$).
 Given a fixed set of genotypes, the evolution of those genotypes to an internally stable equilibrium is called ``short-term evolution'' whereas the process of new mutant invasions that move the population from one internally stable equilibrium to another is called ``long-term evolution'' \cite{Eshel:1991,Eshel:1996,Hammerstein:1996,Weissing:1996,Eshel:Feldman:1998,VanCleve:2015,VanCleve:2020} and eventually leads to an externally stable equilibrium.
@@
 Eshel, Feldman, Lessard and others demonstrated that externally stable equilibria are those that generate phenotypes $\ess{x}$ that are ESS \cite{Eshel:Feldman:1984,Liberman:1988,Eshel:1996,Hammerstein:Selten:1994,Hammerstein:1996,Weissing:1996,Eshel:Feldman:1998}.
-Specifically, suppose there are two alternative phenotypes each determined by two loci $A$ and $B$ with alleles $A_{1},\ldots,A_{m_{A}}$ and $B_{1},\ldots,B_{m_{B}}$, respectively, and with recombination rate $0<\rec\le1/2$ between the loci.
-The vector $\vec{p} = (\ldots,p_{ij},\ldots)$ tracks the frequencies of the chromosome $A_{i} B_{j}$ after selection and recombination.
+Specifically, suppose there are two alternative phenotypes each determined by two loci $A$ and $B$ with alleles $A_{1},\ldots,A_{a}$ and $B_{1},\ldots,B_{b}$, respectively, and with recombination rate $0<\rec\le1/2$ between the loci.
+The vector $\vec{p} = (\ldots,p_{ij},\ldots)$ tracks the frequencies of the chromosomes (e.g., $A_{i} B_{j}$) after selection and recombination.
 Individuals with diploid genotype $A_{i}B_{j}/A_{k}B_{l}$ chooses phenotype $x$ with probability $x_{ijkl}$ and phenotype $y$ with probability $1 - x_{ijkl}$.
@@
 \end{equation*}
-and the mean fitness of an individual with genotype $A_{i}B_{j}/A_{k}B_{l}$ interacting with other individuals in the population according to their frequencies is $w(x_{ijkl}, x)$ (since the fitness function is linear).
+and the mean or marginal fitness of an individual with genotype $A_{i}B_{j}/A_{k}B_{l}$ interacting with other individuals in the population according to their frequencies is $w(x_{ijkl}, x)$ (since the fitness function is linear).
 The mean fitness in the population is then
@@
 \end{equation*}
-Now, assume that a new mutant allele $A_{\mut}$ (increasing the total number of alleles at locus $A$ to $n+1$) arrives at low frequency in a population at an internally stable equilibrium $\fixp{\vec{p}}$ with ``resident'' or wild-type mean phenotype $\fixp{x} = x(\fixp{\vec{p}})$.
+Now, assume that a new mutant allele $A_{\mut}$ (increasing the total number of alleles at locus $A$ to $a+1$) arrives at low frequency in a population at an internally stable equilibrium $\fixp{\vec{p}}$ with ``resident'' or wild-type mean phenotype $\fixp{x} = x(\fixp{\vec{p}})$.
 An individual with this new allele chooses phenotype $x$ with probability $x_{\mut jkl}$ if its genotype is $A_{\mut}B_{j}/A_{k}B_{l}$ and phenotype $y$ with probability $1 - x_{\mut jkl}$.
@@
 
-While Result \ref{res:twolocus} emphasizes the role of ESSs and phenotypic evolution in the long-term process, the external stability approach used to prove \ref{res:twolocus} reveals a more general viewpoint where ESSs can depend in important ways on the genetic and demographic parameters of a population.
+While Result \ref{res:twolocus} emphasizes the role of the ESS natural selection on phenotypes in the long-term evolutionary process, the external stability approach used to prove \ref{res:twolocus} reveals a more general viewpoint where ESSs can depend in important ways on the genetic and demographic parameters of a population.
 External stability analysis assumes that the population is at some genetic equilibrium where new mutant alleles can invade at low frequency and looks for a genetic equilibrium and its resulting phenotype that resists invasion by any mutant allele (within a set of possible phenotypic effects).
@@
 If there are multiple possible genotypes with the mutant allele, as is the case for the two-locus model in section \ref{sec:long-term} where there may be many possible alleles at the non-mutant locus, then mutant genotype fractions during the invasion process are given by the normalized right eigenvector of the linear system $\vec{u}$.
-Thus, the $\eig$ really measures the fitness of the whole mutant genetic lineage.
+Thus, $\eig$ really measures the fitness of the whole mutant genetic lineage.
 The growth rate $\eig$ is often called the ``invasion fitness'' \cite{Metz:Nisbet:1992,Heino:Metz:1998}\footnote{When invasion occurs at more complex attractors like cycles, the ``invasion exponent'' is used, which is $\log |\eig|$ \cite{Rand:Wilson:1994}} and sometimes called the ``lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016,Lehmann:Rousset:2020}.
-In effect, the invasion or lineage fitness measures the average number of offspring copies that produced by a mutant allele sampled randomly from among all the genotypes in the mutant lineage \cite{Lehmann:Mullon:2016}.
+In effect, the invasion or lineage fitness measures the average number of offspring copies produced by a mutant allele sampled randomly from among all the genotypes in the mutant lineage \cite{Lehmann:Mullon:2016}.
 Assuming that $\eig >0$, the phenotype $\ess{x}$ is unbeatable whenever
@@
 The strict ESS condition in \eqref{eq:strict:relative} using relative fitness is identical to the condition for uninvadability in condition \eqref{eq:unbeatable} using invasion fitness, which means that unbeatable or uninvadable phentoypes are strict ESSs\footnote{The relationship between invasion fitness and non-strict ESSs is more complex and depends on specific details of the genetics of the population.
-For example, see the section entitled ``Long-term Stability of an ESS'' in ref.
-\cite{Eshel:Feldman:1998}}.
+For example, see the section entitled ``Long-term Stability of an ESS'' in \cite{Eshel:Feldman:1998}}.
 
@@
 
-We introduce a mutant phenotype $\vec{y}$ at low frequency into a population where all individuals are resident phenotype $\vec{x}$.
-One way to track the number of mutant individuals is to track the number of groups with $k$ mutants.
+We introduce a mutant phenotype $\vec{y}$ at low frequency into a population where all individuals have the resident phenotype $\vec{x}$.
+One way to track the number of mutant individuals in the whole population is to track the number of groups with $k$ mutants.
 Since each individual belongs to a specific class, we actually track how many mutant individuals $k_{s}$ are in each class $s$ in a group or $\vec{k} = (k_{1},\ldots,k_{\numc})$ where $k = \sum_{s} k_{s}$.
 When the mutant phenotype is rare, growth of the mutant population can be modeled using a matrix population model \cite{Caswell:2006} or multitype branching process \cite{Kimmel:Axelrod:2015} with a projection matrix $\vec{A}(\vec{y}, \vec{x}) = [a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x})]$ where the $a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x})$ element measures the expected number of groups with $\vec{k}'$ mutants with phenotype $\vec{y}$ that result from the reproduction occurring in a group with $\vec{k}$ mutants in a population resident for phenotype $\vec{x}$.
-The largest eigenvalue $\eig(\vec{y}, \vec{x})$ of the projection matrix $\vec{A}(\vec{y}, \vec{x})$ is the invasion fitness in this population and satisfies
+The largest eigenvalue $\eig(\vec{y}, \vec{x})$ of the projection matrix $\vec{A}(\vec{y}, \vec{x})$ is the invasion or lineage fitness in this population and satisfies
 \begin{equation}
@@
 where the associated eigenvector $\vec{u}(\vec{y}, \vec{x})$ measures the relative frequency of groups with $\vec{k}$ mutants when the invasion process has stabilized.
-The elements $a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x})$ of the projection matrix can be rewritten in terms of an individual (relative) fitness function $\omega_{s's\vec{k}}(\vec{y}, \vec{x})$, which measures the number of class $s'$ mutant offspring in the next generation produced by a class $s$ mutant individual in a group with $\vec{k}$ mutants.
+The elements $a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x})$ of the projection matrix can be rewritten in terms of an individual (relative) fitness function $\omega_{s'\!s\vec{k}}(\vec{y}, \vec{x})$, which measures the number of class $s'$ mutant offspring in the population in the next generation produced by a class $s$ mutant individual in a group with $\vec{k}$ mutants.
 Relative fitness is the right measure here since we measure offspring production after migration, density dependent survival, and all other life-cycle events.
@@
   \label{eq:a:w}
-  \sum_{k'} k_{s'}' a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x}) = \sum_{s} k_{s} \omega_{s's\vec{k}}(\vec{y}, \vec{x}) \: .
+  \sum_{k'} k_{s'}' a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x}) = \sum_{s} k_{s} \omega_{s'\!s\vec{k}}(\vec{y}, \vec{x}) \: .
 \end{equation}
-\citeauthor{Lehmann:Mullon:2016} \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020} use equation \eqref{eq:a:w} to show that the invasion (or lineage) fitness of mutant phenotype $\vec{x}$ in a resident population with phenotype $\vec{y}$ is given by
+\citeauthor{Lehmann:Mullon:2016} \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020} use equation \eqref{eq:a:w} to show that the invasion (or lineage) fitness of mutant phenotype $\vec{y}$ in a resident population with phenotype $\vec{x}$ is given by
 \begin{equation}
   \label{eq:eig:class}
-  \eig(\vec{y}, \vec{x}) = \sum_{\vec{k},s',s} \omega_{s's\vec{k}}(\vec{y}, \vec{x}) q_{\vec{k},s}(\vec{y}, \vec{x})
+  \eig(\vec{y}, \vec{x}) = \sum_{\vec{k},s'\!,s} \omega_{s'\!s\vec{k}}(\vec{y}, \vec{x}) q_{\vec{k},s}(\vec{y}, \vec{x})
 \end{equation}
@@
 The invasion fitness in equation \eqref{eq:eig:class} can also be expressed in terms of inclusive fitness in order to isolate the effect on fitness of mutant allele expression in the focal individual (direct effect) and in genetically related individuals (indirect effect).
-There are some important technical subtleties as to how this can be accomplished that we will leave out to keep the presentation simple \cite[see ``Supplement B'' in ref.
-][for details]{Lehmann:Rousset:2020}.
-The essence of the method is the same as the one found in many analyses that derive inclusive fitness using the Price equation \cite[e.g.,][]{Queller:1992,Frank:1998,Lehmann:Keller:2006,Akcay:VanCleve:2012}: the fitness of a focal individual, here $\omega_{s's\vec{k}}(\vec{y}, \vec{x})$, is decomposed into cost and benefit terms reflecting the marginal effects on fitness\footnote{Marginal effects in this context means additive effects of the phenotype on fitness and often measured with partial regression coefficients.} of expressing the mutant phenotype $y_{s}$ relative to the resident phenotype $x_{s}$.
-The cost term $c_{s's}(\vec{y}, \vec{x})$ measures the marginal decrease in class $s'$ offspring produced by a class $s$ individual due to expression of the mutant allele in that individual.
-The benefit term $b_{s'\sigma \leftarrow s}(\vec{y}, \vec{x})$ measures the marginal increase in the number of class $s'$ offspring produced individuals of class $\sigma$ due to the mutant allele expression by a groupmate of class $s$.
+There are some important technical subtleties as to how this can be accomplished that we will leave out to keep the presentation simple \cite[see ``Supplement B'' in][for details]{Lehmann:Rousset:2020}.
+The essence of the method is the same as the one found in many analyses that derive inclusive fitness using the Price equation \cite[e.g.,][]{Queller:1992,Frank:1998,Lehmann:Keller:2006,Akcay:VanCleve:2012}: the fitness of a focal individual, here $\omega_{s'\!s\vec{k}}(\vec{y}, \vec{x})$, is decomposed into cost and benefit terms reflecting the marginal effects on fitness\footnote{Marginal effects in this context means additive effects of the phenotype on fitness and often measured with partial regression coefficients.} of expressing the mutant phenotype $y_{s}$ relative to the resident phenotype $x_{s}$.
+The cost term $c_{s'\!s}(\vec{y}, \vec{x})$ measures the marginal decrease in class $s'$ offspring produced by a class $s$ individual due to expression of the mutant allele in that individual.
+The benefit term $b_{s'\!\sigma \leftarrow s}(\vec{y}, \vec{x})$ measures the marginal increase in the number of class $s'$ offspring produced by individuals of class $\sigma$ due to the mutant allele expression by a groupmate of class $s$.
 The invasion fitness can then be expressed as (using an ``actor-modulated'' approach) \cite{Lehmann:Rousset:2020}
@@
   \label{eq:eig:IF}
-  \eig(\vec{y}, \vec{x}) = 1 + \sum_{s',s} q_{s}(\vec{y}, \vec{x}) v_{s'}(\vec{y}, \vec{x})
-  \left( - c_{s's}(\vec{y}, \vec{x}) + \sum_{\sigma} b_{s'\sigma \leftarrow s}(\vec{y}, \vec{x}) r_{\sigma|s}(\vec{y}, \vec{x}) \right) \: ,
+  \eig(\vec{y}, \vec{x}) = 1 + \sum_{s'\!,s} q_{s}(\vec{y}, \vec{x}) v_{s'}(\vec{y}, \vec{x})
+  \left( - c_{s'\!s}(\vec{y}, \vec{x}) + \sum_{\sigma} b_{s'\!\sigma \leftarrow s}(\vec{y}, \vec{x}) r_{\sigma|s}(\vec{y}, \vec{x}) \right) \: ,
 \end{equation}
 which is the inclusive fitness \cite{Hamilton:1964} of the mutant allele.
-In equation \eqref{eq:eig:IF}, $r_{\sigma|s}(\vec{y}, \vec{x})$ is genetic relatedness and is the probability that a random groupmate of class $\sigma$ has the mutant allele given an individual of class $s$ has the mutant allele, $v_{s'}(\vec{y}, \vec{x})$ is the normalized reproductive value of class $s'$, and $q_{s}(\vec{y}, \vec{x})$ is the probability that a mutant individual in the invasion process is in class $s$ (see ref.
-\cite{Lehmann:Rousset:2020} for technical definitions of these quantities).
+In equation \eqref{eq:eig:IF}, $r_{\sigma|s}(\vec{y}, \vec{x})$ is the genetic relatedness in a group and is the probability that a random groupmate of class $\sigma$ has the mutant allele given an individual of class $s$ has the mutant allele, $v_{s'}(\vec{y}, \vec{x})$ is the normalized reproductive value of class $s'$, and $q_{s}(\vec{y}, \vec{x})$ is the probability that a mutant individual in the invasion process is in class $s$ (see \cite{Lehmann:Rousset:2020} for technical definitions of these quantities).
 Equation \eqref{eq:eig:IF} shows how invasion fitness can be partitioned into the direct effects of mutant allele expression, which are cost terms, and the indirect effects, which are the benefit times relatedness terms.
@@
 
-Given that previous work using the Price equation has shown that kin selection and group selection partitions of fitness are simply alternative ways of expressing genotype frequency change in a population \cite{Queller:1992,Bijma:Wade:2008,Gardner:West:2007,Marshall:2011,Akcay:VanCleve:2012}, it should be possible to rewrite invasion fitness so as to group terms that constitute between-group selection, $\omega_{G,s's}(\vec{y}, \vec{x})$, and those that constitute within-group selection, $\omega_{\Delta G,s's}(\vec{y}, \vec{x})$.
-In fact, if we apply a ``neighbor-modulated'' approach from refs.
-\cite{Queller:1992,Akcay:VanCleve:2012} to the results from ref.
-\cite{Lehmann:Rousset:2020}, we obtain
+Given that previous work using the Price equation has shown that kin selection and group selection partitions of fitness are simply alternative ways of expressing genotype frequency change in a population \cite{Queller:1992,Bijma:Wade:2008,Gardner:West:2007,Marshall:2011,Akcay:VanCleve:2012}, it should be possible to rewrite invasion fitness so as to group terms that constitute between-group selection, $\omega_{G,s'\!s}(\vec{y}, \vec{x})$, and those that constitute within-group selection, $\omega_{\Delta G,s'\!s}(\vec{y}, \vec{x})$.
+In fact, if we apply a ``neighbor-modulated'' approach from \cite{Queller:1992,Akcay:VanCleve:2012} to the results from \cite{Lehmann:Rousset:2020}, we obtain
 \begin{equation}
   \label{eq:eig:GS}
-  \eig(\vec{y}, \vec{x}) = 1 + \sum_{s',s} q_{s}(\vec{y}, \vec{x}) v_{s'}(\vec{y}, \vec{x})
-  \left( \omega_{G,s's}(\vec{y}, \vec{x}) + \omega_{\Delta G,s's}(\vec{y}, \vec{x}) \right)
+  \eig(\vec{y}, \vec{x}) = 1 + \sum_{s'\!,s} q_{s}(\vec{y}, \vec{x}) v_{s'}(\vec{y}, \vec{x})
+  \left( \omega_{G,s'\!s}(\vec{y}, \vec{x}) + \omega_{\Delta G,s'\!s}(\vec{y}, \vec{x}) \right)
 \end{equation}
@@
   \begin{split}
-    \omega_{G,s's} & = \left(b_{s's \leftarrow s} - c_{s's} \right) \left(\frac{1}{n_{s}} + \frac{n_{s}-1}{n_{s}} r_{\sigma|s} \right) + \sum_{\sigma \ne s} b_{s's \leftarrow \sigma} r_{\sigma|s} \\
-    \omega_{\Delta G,s's} & = - \left(b_{s's \leftarrow s} + c_{s's}\right) \frac{n_{s}-1}{n_{s}} \left( 1 - r_{\sigma|s} \right) \: .
+    \omega_{G,s'\!s} & = \left(b_{s'\!s \leftarrow s} - c_{s'\!s} \right) \left(\frac{1}{n_{s}} + \frac{n_{s}-1}{n_{s}} r_{s|s} \right) + \sum_{\sigma \ne s} b_{s'\!s \leftarrow \sigma} r_{\sigma|s} \\
+    \omega_{\Delta G,s'\!s} & = - \left(b_{s'\!s \leftarrow s} + c_{s'\!s}\right) \frac{n_{s}-1}{n_{s}} \left( 1 - r_{s|s} \right) \: .
   \end{split}
 \end{equation}
-We can see immediately that if mutant allele expression is costly to the individual itself, $c_{s's}>0$, helps others, $b_{s's \leftarrow \sigma}>0$ for all classes $\sigma$, and the benefits outweigh the costs within a class, $b_{s's \leftarrow s} - c_{s's}>0$, then the mutant allele expression is positively selected by between-group selection and negatively selected by within-group selection.
-Since both the kin selection and group selection partitions of invasion fitness (equations \eqref{eq:eig:IF} and \eqref{eq:eig:GS}, respectively) are expressed using the same benefit, cost, and relatedness functions, a single ESS analysis using invasion can answer questions derived both both perspectives.
+We can see immediately that if mutant allele expression is costly to the individual itself, $c_{s'\!s}>0$, helps others, $b_{s'\!s \leftarrow \sigma}>0$ for all classes $\sigma$, and the benefits outweigh the costs within a class, $b_{s'\!s \leftarrow s} - c_{s'\!s}>0$, then the mutant allele expression is positively selected by between-group selection and negatively selected by within-group selection.
+Since both the kin selection and group selection partitions of invasion fitness (equations \eqref{eq:eig:IF} and \eqref{eq:eig:GS}, respectively) are expressed using the same benefit, cost, and relatedness functions, a single ESS analysis using invasion can answer questions derived from both perspectives.
 
@@
 Given the connection between invasion fitness and ESS analysis, the reduction principle shows that rates of zero recombination, mutation, or migration are strict ESSs.
-In effect, the reduction principle says natural selection on populations at equilibrium in constant environments acts to reduce genetic and demographic processes that increase genetic variation; this is intuitive as such genetic variation must results in lower fitness genotypes since the population is already at an equilibrium and the environment is constant.
+In effect, the reduction principle says natural selection on populations at equilibrium in constant environments acts to reduce genetic and demographic processes that increase genetic variation; this is intuitive as such genetic variation must result in lower fitness genotypes since the population is already at an equilibrium and the environment is constant.
 Like Result \ref{res:twolocus}, the reduction principle holds independent of the recombination rate between the modifier locus and major loci that affect fitness.
@@
 Thus, natural selection may favor genetic and demographic mechanisms that generate genetic variation like positive rates of recombination, mutation, and migration.
-In fact, environmental variation, particularly over time, is one of the most well-studied and important factors supporting the evolution of genotypic and phenotypic variation and the evolution of the mechanisms that promote such variation like recombination \cite{Charlesworth:1976,Sasaki:Iwasa:1987,Otto:Michalakis:1998}, mutation \cite{Leigh:1970,Ishii:Matsuda:1989,Lachmann:Jablonka:1996}, migration \cite{Gillespie:1981,McPeek:Holt:1992,Blanquart:Gandon:2011}, phenotypic plasticity \cite{Caswell:1983,Via:Lande:1985,Gavrilets:Scheiner:1993,Jong:1995} and bet-hedging \cite{Slatkin:1974,Seger:Brockmann:1987,Kussell:Leibler:2005,Salathe:VanCleve:2009}.
+In fact, environmental variation, particularly over time, is one of the most well-studied and important factors supporting the evolution of genotypic and phenotypic variation through mechanisms that promote such variation like recombination \cite{Charlesworth:1976,Sasaki:Iwasa:1987,Otto:Michalakis:1998}, mutation \cite{Leigh:1970,Ishii:Matsuda:1989,Lachmann:Jablonka:1996}, migration \cite{Gillespie:1981,McPeek:Holt:1992,Blanquart:Gandon:2011}, phenotypic plasticity \cite{Caswell:1983,Via:Lande:1985,Gavrilets:Scheiner:1993,Jong:1995} and bet-hedging \cite{Slatkin:1974,Seger:Brockmann:1987,Kussell:Leibler:2005,Salathe:VanCleve:2009}.
 Recent work shows that not only are the ES recombination, mutation, and migration evolve rates positive in variable environments, but they also depends on genetic parameters like the recombination rate between the major and modifier loci.
-In the case of mutation rate evolution, \citeauthor{Liberman:VanCleve:2011} \cite{Liberman:VanCleve:2011} use an analysis of the external stability of the mutation rate to show that the ES mutation rate depends on the recombination rate between phenotypic locus and the modifier locus.
+In the case of mutation rate evolution, \citeauthor{Liberman:VanCleve:2011} \cite{Liberman:VanCleve:2011} used an analysis of the external stability of the mutation rate to show that the ES mutation rate depends on the recombination rate between phenotypic locus and the modifier locus.
 For example, Figure \ref{fig:pip:switching}, shows pairwise invasibility plots for the mutation rate with different recombination rates and given weakly asymmetric selection in two different environments that alternate every $T=50$ generations.
 White regions are where mutant alleles with a switching rate on the vertical axis can invade a population with a resident rate given by the horizontal axis; black regions are when such a mutant cannot invade.
-The plots show that a rate of $~1/50$ is ES when the recombination rate is zero, similar to older results in \cite{Leigh:1970,Lachmann:Jablonka:1996} that show the ES rate is $~1/T$.
+The plots show that a rate of $~1/50$ is ES when the recombination rate is zero, similar to older results \cite{Leigh:1970,Lachmann:Jablonka:1996} that show the ES rate is $~1/T$.
 The ES switching rate decreases as recombination increases until there is an abrupt shift to an ES rate of zero switching.
 \citeauthor{Carja:Liberman:2014} \cite{Carja:Liberman:2014} generalize this approach to the evolution of recombination, mutation, and migration in fluctuating environments.
-They first show that the rate of environmental fluctuation has the same effect on ES recombination and migration rates as it does on ES mutation rates, namely that slower rates of fluctuation lead to slower recombination, mutation, and migration rates.
+They first show that the rate of environmental fluctuation has the same effect on ES recombination and migration rates as it does on ES mutation rates, namely that slower rates of fluctuation lead to slower recombination, mutation, and migration rates (see Figure 1 in \cite{Carja:Liberman:2014}).
 In other words, as the rate of environmental change slows, so do the ES rates of genetic and demographic processes that generate genetic variation.
-\citeauthor{Carja:Liberman:2014} also show that these ES rates all decrease with increasing rates of recombination between the major and modifier loci.
+\citeauthor{Carja:Liberman:2014} also show that these ES rates all decrease with increasing rates of recombination between the major and modifier loci (see Figure S5 in \cite{Carja:Liberman:2014}).
 This can be understood as due to the fact that the strength of indirect selection at the modifier locus decreases as increased recombination erodes linkage disequilibrium between the major and modifier loci.
@@
 Put another way, how justifiable is the phenotypic gambit?
-Result \ref{res:twolocus} shows that for large and randomly mating populations, ESSs can be fairly independent of genetic parameters like mutation and recombination rates and even the number of loci (for mixed ESSs; see refs.
-\cite{Eshel:Feldman:1998,Eshel:Feldman:2001}).
+Result \ref{res:twolocus} shows that for large and randomly mating populations, ESSs can be fairly independent of genetic parameters like mutation and recombination rates and even the number of loci (for mixed ESSs; see \cite{Eshel:Feldman:1998,Eshel:Feldman:2001}).
 At the time, this independence was important for supporting the use of ESS analysis across a broad range of species with different genetics and demography.
 However, this independence may no longer be as useful given the growing abundance of genomic, transcriptomic, and long-term demographic data for natural populations.
-Instead, we might rather want to make use of these data to refine our evolutionary models, and ESS may be a less useful approach if it cannot take advantage of these data.
+Instead, we might rather want to make use of these data to refine our evolutionary models, and the ESS may be a less useful approach if it cannot take advantage of these data.
 However, long-term evolution and consequently ESSs are in fact sensitive to genetics and demography for many traits and populations more complex than Result \ref{res:twolocus} assumes.
@@
 All of the ESS analysis so far as concerned a single phenotypic trait.
-However, the principle of invasion fitness can also be applied to multiple evolving traits \cite{Leimar:2009,Mullon:Keller:2016,Mullon:Lehmann:2019}.
+However, the principle of invasion fitness can also be applied to multiple coevolving traits \cite{Leimar:2009,Mullon:Keller:2016,Mullon:Lehmann:2019}.
 When multiple phenotypes coevolve, issues of pleiotropy and genetic constraint immediately arise as mutations may not cause independent effects along the different phenotypic dimensions.

### 2023-01-14 manuscript.tex
edis plus cover letter add

@@
 For the purpose of explaining the theory more concretely, assume that individuals can choose one of two possible behaviors in the interaction, $C$ and $D$.
-The phenotypes $x$ and $y$ measure the probability that individuals in an interaction choose behavior $C$, and $1-x$ and $1-y$ are the probabilities those individuals choose behavior $D$ instead (probabilistic phenotypes are called mixed strategies in game theory \cite{Osborne:Rubinstein:1994}).
+The phenotypes $x$ and $y$ measure the probability that individuals in an interaction with those phenotypes choose behavior $C$, and $1-x$ and $1-y$ are the probabilities those individuals choose behavior $D$ instead (probabilistic phenotypes are called mixed strategies in game theory \cite{Osborne:Rubinstein:1994}).
 For example, the behavior $C$ could be cooperation in the interaction and $D$ could be a lack of cooperation.
@@
 
-Early mathematical analyses began by using the simplest population genetic model, which is one that assumes a \textit{haploid asexual} population with two genotypes, one with phenotype $x=1$ that chooses $C$ with certainty and one that with phentoype $x=0$ that chooses $D$ with certainty.
+Early mathematical analyses began by using the simplest population genetic model, which is one that assumes a \textit{haploid asexual} population with two genotypes, one with phenotype $x=1$ that chooses $C$ with certainty and one with phentoype $x=0$ that chooses $D$ with certainty.
 There are four possible interacting pairs of phenotypes in this scenario yielding four fitnesses that can be collected into a so-called payoff matrix $W$:
@@
 For this scenario, the phenotype is controlled by a single genetic locus with $a$ possible alleles $A_{1},\ldots,A_{a}$.
-Diploid genotype $A_{i} A_{j}$ cooperates with probability $x_{ij}$ and does not cooperate with probability $1-x_{ij}$.
+Diploid genotype $A_{i} A_{j}$ chooses behavior $C$ with probability $x_{ij}$ and behavior $D$ with probability $1-x_{ij}$.
 If there are no restrictions on the $x_{ij}$, then dominance is possible between the alleles.
 Let $\vec{p} = (p_{1}, \ldots, p_{a})$ be the vector of allele frequencies for alleles $A_{1},\ldots,A_{a}$.
-Given allele frequencies $\vec{p}$ in the parental generation and assuming random mating, the mean phenotype or probability of cooperation in the population after reproduction is
+Given allele frequencies $\vec{p}$ in the parental generation and assuming random mating, the mean phenotype or probability of choosing $C$ in the population after reproduction is
 \begin{equation*}
@@
 Eshel, Feldman, Lessard and others demonstrated that externally stable equilibria are those that generate phenotypes $\ess{x}$ that are ESS \cite{Eshel:Feldman:1984,Liberman:1988,Eshel:1996,Hammerstein:Selten:1994,Hammerstein:1996,Weissing:1996,Eshel:Feldman:1998}.
-Specifically, suppose there are two alternative phenotypes each determined by two loci $A$ and $B$ with alleles $A_{1},\ldots,A_{a}$ and $B_{1},\ldots,B_{b}$, respectively, and with recombination rate $0<\rec\le1/2$ between the loci.
+Specifically, suppose that the probability of selecting between behaviors $C$ and $D$ is determined by two loci $A$ and $B$ with alleles $A_{1},\ldots,A_{a}$ and $B_{1},\ldots,B_{b}$, respectively, and with recombination rate $0<\rec\le1/2$ between the loci.
 The vector $\vec{p} = (\ldots,p_{ij},\ldots)$ tracks the frequencies of the chromosomes (e.g., $A_{i} B_{j}$) after selection and recombination.
-Individuals with diploid genotype $A_{i}B_{j}/A_{k}B_{l}$ chooses phenotype $x$ with probability $x_{ijkl}$ and phenotype $y$ with probability $1 - x_{ijkl}$.
-The frequency of phenotype $x$ in the population after reproduction is
+Individuals with diploid genotype $A_{i}B_{j}/A_{k}B_{l}$ chooses behavior $C$ with probability $x_{ijkl}$ and behavior $D$ with probability $1 - x_{ijkl}$.
+The mean frequency of phenotype $x$ in the population after reproduction is
 \begin{equation*}
-  x = x(\vec{p}) = \sum_{i,j,k,l} p_{ij} p_{kl} x_{ijkl}
+  x = x(\vec{p}) = \sum_{i,j,k,l} p_{ij} p_{kl} x_{ijkl} \, .
 \end{equation*}
-and the mean or marginal fitness of an individual with genotype $A_{i}B_{j}/A_{k}B_{l}$ interacting with other individuals in the population according to their frequencies is $w(x_{ijkl}, x)$ (since the fitness function is linear).
+The mean or marginal fitness of an individual with genotype $A_{i}B_{j}/A_{k}B_{l}$ interacting with other individuals in the population according to their frequencies is $w(x_{ijkl}, x)$ (since the fitness function is linear).
 The mean fitness in the population is then

### 2023-01-14 rev1_cover_letter.tex
edis plus cover letter add

\documentclass[10pt,parskip=full,foldmarks=off,addrfield=off,backaddress=false,refline=dateleft,letterpaper]{scrlttr2}
\LoadLetterOption{kentucky}

\usepackage{xspace}

% Sender's information
\setkomavar{signature}{%
  \includegraphics[scale=0.3]{hand_sig.pdf}\\%
  Jeremy Van Cleve\\%
  Associate Professor\\%
  Department of Biology\\%
  University of Kentucky} % use PDF image of signature
\setkomavar{date}{\today}
\setkomavar{subject}{}

\usepackage[explicit]{titlesec}
\titleformat{\section}{\scshape \bfseries}{}{0ex}{#1}
\titlespacing{\section}{0pt}{1.0ex plus 0.25ex minus 0.25ex}{0.5ex plus 0.1ex minus 0.1ex}

\newenvironment{reviewerquote}{\begin{quote}\color{DarkBlue}\itshape}{\end{quote}}
\newenvironment{response}{}{}

\begin{document}

% Recipient's information
\begin{letter}

  \opening{Dear Phil Trans B}

  Please find enclosed my revision of the manuscript entitled ``Evolutionarily stable strategy analysis and its links to demography and genetics through invasion fitness''. The Reviewers made a number of helpful and probing comments, and I have attempted to address each one thoroughly below. Reviewer comments are quoted in blue italics and my responses immediately follow. I have also uploaded a ``tracked changes'' version of the PDF to enable easy reference for changes that have been made in response to the reviews.

  \closing{Best regards,}

\section{Editor comments}

\begin{reviewerquote}
  Reviewer 1 has made several useful suggestions for clarity and they should be straightforward to implement in the revision.

  While Reviewer 2 also finds the manuscript well written, they are more critical and make several comments which should be addressed and may require more extensive revision.
\end{reviewerquote}

I would like to thank the Editor and Reviewers for their very detailed in insightful comments. I have copied each Reviewer's comments below followed by a detailed response with changes I've made in response to the comments.

\section{Referee 1 comments}

\begin{reviewerquote}
  This paper reviews the history and current state of the field of ESS analysis. The author links ESS analysis to the concept of invasion fitness, and discusses the application of invasion fitness to kin selection and evolution in variable environments Although I learned some valuable and interesting things, and I enjoyed reading the first half of this paper, I also found it difficult to understand some of the latter parts of the paper. Since reviews usually aim to clarify and give a (relatively) simple overview of the literature, I think the paper would benefit from some rewrites and from more careful notation and explanation of notation.
\end{reviewerquote}

Thank you for the excellent feedback. I've endeavored to improve the latter parts of the paper and provide clearer explanations. Please see my responses below to the reviewer's specific comments.

\begin{reviewerquote}
Confusion/clarification/notational questions:
- You are using $x$ and $y$ to denote both the phenotypes, and their frequencies, and I’m not always sure which one you mean. E.g. up until line 138, x and y have been defined as denoting the phenotypes, so I assume that $w(x,y)$ in line 127 means that $x$ and $y$ are interacting, and it does not mean that w is a function of the frequencies of $x$ and $y$? It would help me a lot if you used different symbols for these two things, e.g. maybe $p_x$ and $p_y$ for the frequencies?
- Line 135: So now we have x denoting both the phenotype and the frequency of the phenotype, and then we have $x_{ij}$ denoting mating preference of genotype $A_i$ $A_j$. Is it possible to use a different symbol for $x_{ij}$?
\end{reviewerquote}

Thanks for mentioning this. I did start the section talking about two strategies, $x$ and $y$, and thus do mean as suggested above that $w(x,y)$  is a function of phenotypic values $x$ and $y$ and not of frequencies of those strategies in the population. However, I think what may be causing confusion is that then I moved to allowing individuals to choose strategies $x$ and $y$ probabilistically, so now mixed strategies are allowed and the strategy is technically the probability of selecting strategy $x$ (with complementary probability of selecting strategy $y$). Since the probabilities $x_{ij}$ are strategies, I'd like to keep the same notation $x$ for the probabilities.

However, a reworking of this section and some notation change I hope will make things simpler and clearer. I've started off assuming in section 2 that the strategies are probabilities of selecting behavior $C$ when individuals have two possible behaviors, $C$ and $D$. Now, the equation on line 138 (now eqn 2) is more clearly described as the mean strategy in the population or mean probability of selecting behavior $C$ and is no longer confusingly described as a frequency. I've added additional explanation in sections 2 and 3 to make the description of the phenotypes clearer.

\begin{reviewerquote}
- I’m not sure I understand what “the ESSs obtained by applying condition (1) to the payoff matrix $A$” means. Does it mean that we are considering some trait to evolve until the entries of the pay off matrix satisfies the ESS conditions, are x and y the evolving phenotypic trait? Does it means that at some frequencies of x and y, the entries of the pay off matrix satisfy the ESS condition (related to the previous question)?
\end{reviewerquote}

Yes, I agree that this is unclear. I have rewritten this paragraph to better describe both the population genetic model and the ESS analysis. The ESS analysis is simply to use the condition (1) to determine which of the two strategies is ESS or if there is a mixed strategy ESS. It should be clear now that the ESS value of the phenotype is the same value of the genotype frequencies that are stable in the population genetic model.

\begin{reviewerquote}
  - Line 184: Equation for mean fitness: what happened to $y$? Shouldn’t the mean fitness in the population also be a function of phenotype $y$?
\end{reviewerquote}

The equation is correct as written and the confusion here stems from the earlier issue the reviewer highlighted. In this scenario, the phenotype $x$ is the probability of selecting behavior $C$ (which I was calling ``$x$'' before confusingly) and thus the probability of selecting behavior $D$ is simply $1-x$. Thus, only one phenotypic variable, $x$, is needed to calculate the mean fitness. I have added additional clarification around this equation in addition to the earlier text clarifying the phenotypes.

\begin{reviewerquote}
  - Line 182: “the fitness function is linear”, linear in what/a linear function of what?
\end{reviewerquote}

I've had added that the fitness $w(x,y)$ is a linear function of its arguments, which are the phenotypes of the two interacting individuals. Linearity here is reasonable given the phenotype is a probability of selecting among two pure strategies.

\begin{reviewerquote}
  - Equation 3: $u_j$ is not yet defined I think? $u_{xy}$ has been defined as an entry of the pay off matrix, but this $u$ only comes with one index?
\end{reviewerquote}

\begin{reviewerquote}
- Line 212-213: “under the right assumptions”, and earlier in that paragraph “given the general genetic and demographic assumptions above”. I read through the above again to try and find what these assumptions are but had trouble locating them, except for maybe “the fitness function is linear” and “mutant allele is rare”? Can you repeat/list the assumptions necessary for Result 1 to hold?
- I guess $u_k(y,x)$ on line 305 is an element in the vector $u(y,x)$ from line 292?
- What are the dimensions of the matrix $A(y,x)$ in line 287? I’ve never encountered vectors used as matrix indices before (as in $a_{k’k}$, usually matrix indices are scalars), and I’m not sure how to interpret the notation? If $A$ contains all possible $k$ and $k’$ vectors, its dimensions must quickly get very large?
- Really enjoyed the section on variable environments.
Minor comments:
- Line 53: I’m not sure what it means for a method to be orthogonal to issues
Typos:
- Line 87: exits -> exists
- Line 97: in that in can in
- Line 155 an genetic -> a genetic
- Line 347 “One of the important features of Result 1 is that shows how”
\end{reviewerquote}

\section{Referee 2 comments}

\begin{reviewerquote}

This manuscript attempts to review the 50-year history of the ESS concept. In view of the more than 50,000 “ESS articles” published since 1973, this is a highly ambitious task. Although I agree with some of the take-home messages of the manuscript, I am not entirely convinced that Van Cleve has succeeded in his mission.

From the start, the ESS history was multi-faceted. The 1973 paper of Maynard Smith and Price was based on an earlier manuscript of Price (which I could inspect when visiting Maynard Smith many years ago). Here, the main novel idea was not the ESS concept but rather “strategy thinking”, adopted from game theory (strategy = recipe for condition-dependent behaviour). Price argued that a “mouse” strategy (nowadays called Dove) could be rapidly outcompeted by a more aggressive strategy, but that this would not be the case for a ”retaliator” strategy, which, in the absence of aggressive individuals, leads to precisely the same behaviour as the “mouse” strategy. From this, Price drew the (for him) most important conclusion that the “organisation of behaviour” (that is, the “strategy” underlying behaviour) is crucial for understanding behavioural evolution. In contrast, Maynard Smith had one main goal: to publish his ESS concept before Hamilton could work out his ideas on “unbeatable strategies” in more detail.

I mention all this because Von Cleve starts talking about evolutionary “game theory”, while game aspects (like strategy thinking) play a minor role in his review. He is not really interested in behavioural interactions, but in generalizing the ESS concept to more complicated settings. At first, I found this quite confusing, as, from the historical perspective, it is crucial that the ESS concept was first developed in a game context. In fact, Maynard Smith’s derivation of the ESS conditions (1a) and (1b) only makes sense for two-person matrix games, as in this case the fitness function w(x,y) is linear in both components. Various authors later “corrected” this “mistake” and derived more general ESS concepts for non-linear fitness functions. It is a pity that all this is not mentioned by Van Cleve. Indeed, non-linearities arise, for example, in “games between relatives”. How to deal with this (and, hence, how to capture kin selection in the ESS formalism) led to a vivid exchange of arguments between Grafen and Maynard Smith in the late 1970s and the early 1980s. Grafen pointed out that his approach (corresponding to an “actor-modulated approach”) leads to quite different outcomes than Maynard Smith’s inclusive fitness approach. As kin selection is a central theme of Van Cleve’s manuscript, I would have expected to hear about these early debates.

Anyway, it is now quite obvious that Maynard Smith would have designed the ESS conditions quite differently if he had considered the possibility of a non-linear fitness function. It soon became clear that such functions can have quite different evolutionary properties, like “Garden of Eden” scenarios where an ESS is not attainable, or “evolutionary branching” scenarios where the population converges to an evolutionarily unstable strategy. Later work on this resulted in Adaptive Dynamics theory, which is not mentioned at all in Van Cleve’s manuscript, although the invasion fitness concept of Adaptive Dynamics is more sophisticated than the variant propagated by Van Cleve. Interestingly, from the Adaptive Dynamics perspective, the second ESS condition of Maynard Smith is not a condition for “evolutionary stability” at all, but rather a condition for “convergence stability”. This has been noticed before (in writing) by various colleagues, including Eshel, Lessard, Metz, and Dieckmann. It would have been very interesting to hear Van Cleve’s position regarding this reinterpretation of “ESS history”.

The middle part of Van Cleve’s review (Sections 3 and 4) contains some interesting observations, but the results are often stated in a relatively sloppy manner (i.e., without mentioning the conditions under which they hold). In fact, the references referred to in the review all make quite limiting assumptions, such as a linear fitness function (in both components) or a system with only two phenotypes. Take, for example, the discussion of internal versus internal stability in Section 4, two very interesting concepts indeed. These concepts are mainly useful in systems with many different strategies. In contrast, the central Result 1 of that Section only applies to systems with two pure strategies (or phenotypes), and it does not hold anymore in three-strategy games like Rock-Scissors-Paper (this is, for example, explicitly mentioned in the key reference 103). Obviously, this limitation of Result 1 should have been mentioned. In fact, various counter-examples to both parts of Result 1 were constructed in the evolutionary branch of economic game theory (most notably by Friedman and Nachbar). Like my colleagues from economics, I also spent several years on attempts to extend Result 1 to more than two phenotypes. Like them, I had to conclude that long-term evolution does not “save” the ESS concept when more than two strategies are around.

But also the case of only two phenotypes is less clear than presented by Van Cleve. A good example can be found in the excellent book of Karlin and Lessard on sex ratio evolution (which, historically, was very important for the idea of long-term evolution). Karlin and Lessard realized that genetic constraints often prevent the evolution of an “evolutionarily stable” sex ratio (sensu Maynard Smith). Around this time, they collaborated with Eshel, and together the three considered long-term evolution. Karlin and Lessard decided to conduct a long-term simulation study by once in a while adding new genetic variants, in order to remove genetic constraints. This study resulted, however, in deception, as it did not converge at all. Sometimes the trajectory converged to a state close to the ESS sex ratio, but consecutively, it always moved away again. Later, I found the same in my own simulations, both in the context of sex ratio evolution and in the context of two-strategy games. Accordingly, I am much less convinced than Van Cleve that “long-term evolution” has been shown to “save” the ESS concept.

In contrast to the earlier sections, which are broadly accessible, Sections 6 and 7 are technically more sophisticated and will most likely only be understandable to experts. This holds, in particular, for Section 7. I wonder whether it is really useful to present Fig. 1 without discussing the underlying assumptions in sufficient depth and detail. The “reduction principle” is a wonderful tool but I would have liked to see a more critical discussion of its applicability.

Generally, the manuscript is well-written. But as mentioned above, there is a certain imbalance in this manuscript: no (restrictive) assumptions are stated in the first sections (although the experts would have liked to see them spelt out explicitly), while the treatment in the later sections is only understandable to experts in the field.

\end{reviewerquote}

\end{letter}

\end{document}

### 2023-01-16 manuscript.tex
more edits

@@
 \newcommand{\eig}{\lambda}
-\newcommand{\numc}{\mathcal{S}}
+\newcommand{\numc}{c}
 \newcommand{\cost}{}
@@
  Suppose that $p$ measures the frequency of the phenotype $x=1$ that always chooses $C$.
-The equilibria of this model are either fixation of one of the two genotype resulting a population of individuals that always chooses $C$, $p=1$, or always choose $D$, $p=0$ or a polymorphism of both genotypes, $0 < p < 1$, resulting in a population where some choose $C$ and some choose $D$.
+Pairs are assumed to form randomly according to the frequencies $p$ and $1-p$ of the two types.
+The equilibria of this model are either fixation of one of the two genotypes resulting in a population of individuals that always chooses $C$, $p=1$, or always choose $D$, $p=0$ or a polymorphism of both genotypes, $0 < p < 1$, resulting in a population where some choose $C$ and some choose $D$.
 The analogous ESS model of this scenario applies condition \eqref{eq:ess:def} to the four possible fitness values in the payoff matrix $W$ and determines which phenotype, $x=1$ or $x=0$, is ESS or whether a mixed strategy (i.e., $0 < x < 1$) is ESS.
@@
 If there are no restrictions on the $x_{ij}$, then dominance is possible between the alleles.
-Let $\vec{p} = (p_{1}, \ldots, p_{a})$ be the vector of allele frequencies for alleles $A_{1},\ldots,A_{a}$.
-Given allele frequencies $\vec{p}$ in the parental generation and assuming random mating, the mean phenotype or probability of choosing $C$ in the population after reproduction is
+Like the asexual haploid model, diploid individuals interact in pairs according to their probability in the population and mating is random.
+Let $\vec{p} = (p_{1}, \ldots, p_{a})$ be the vector of allele frequencies for alleles $A_{1},\ldots,A_{a}$ in the parental generation.
+The mean phenotype or probability of choosing $C$ in the population after reproduction is then
 \begin{equation*}
@@
 Given a fixed set of genotypes, the evolution of those genotypes to an internally stable equilibrium is called ``short-term evolution'' whereas the process of new mutant invasions that move the population from one internally stable equilibrium to another is called ``long-term evolution'' \cite{Eshel:1991,Eshel:1996,Hammerstein:1996,Weissing:1996,Eshel:Feldman:1998,VanCleve:2015,VanCleve:2020} and eventually leads to an externally stable equilibrium.
+The process of long-term evolution assumes that mutations are infrequent enough so that a new mutation has sufficient time to either go extinct or reach a new short-term equilibrium before additional mutations occurs.
+Exactly how infrequent this is depends on additional parameters such as the strength of selection and the population size and requires analyzing the stochastic interaction between mutation, drift, and selection \cite{Gillespie:1991,Champagnat:2006,Champagnat:Ferriere:2006,Desai:Fisher:2007}.
+Long-term evolution involving sequential substitution of genotypes is sometimes called the ``streetcar process'' \cite{Hammerstein:1996} or the ``trait substitution sequence'' \cite{Metz:Nisbet:1992,Dieckmann:Law:1996,Champagnat:Ferriere:2001} and is the fundamental evolutionary model of the \textit{adaptive dynamics} approach \cite{Metz:Geritz:1996,Geritz:Kisdi:1998,Dercole:Rinaldi:2008}.
 
@@
 \end{equation*}
-The mean or marginal fitness of an individual with genotype $A_{i}B_{j}/A_{k}B_{l}$ interacting with other individuals in the population according to their frequencies is $w(x_{ijkl}, x)$ (since the fitness function is linear).
-The mean fitness in the population is then
+If we assume that the fitness function $\w(x,y)$ is linear in the phenotypes $x$ and $y$ (which is a reasonable assumption given the phenotypes are probabilities of selection among behaviors $C$ and $D$), then the mean or marginal fitness of an individual with genotype $A_{i}B_{j}/A_{k}B_{l}$ interacting with other individuals in the population according to their frequencies is $w(x_{ijkl}, x)$.
+Likewise, the mean fitness in the population is then
 \begin{equation*}
@@
 A linear stability analysis shows that this allele will either initially increase or decrease in frequency at a geometric rate $\eig$ that is the magnitude of the largest eigenvalue of a Jacobian matrix \cite{Edelstein-Keshet:2005}.
-The eigenvalue $\eig$ has an associated normalized eigenvector $\vec{u}$ that measures the frequency of mutant chromosomes (i.e., chromosomes $A_{\mut} B_{j}$) once the invasion process has stabilized (i.e., has reached stationarity \cite{Caswell:2006}).
+The eigenvalue $\eig$ has an associated normalized eigenvector $\vec{u}$ whose elements $u_{j}$ measure the frequency of mutant chromosomes (i.e., chromosomes $A_{\mut} B_{j}$) once the invasion process has stabilized (i.e., has reached stationarity \cite{Caswell:2006}).
 Eshel and Feldman \cite{Eshel:Feldman:1984,Eshel:Feldman:1998} showed that this geometric growth rate is
@@
 \noindent
-Result \ref{res:twolocus} likely extends to more than two loci given work by Liberman \cite{Liberman:1988}.
+Work by Liberman \cite{Liberman:1988} on more than two loci likely extends Result \ref{res:twolocus} to any finite number of loci.
 
-One important conclusion from Result \ref{res:twolocus} is that the ESS condition of Maynard Smith and Price \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} is an appropriate stability condition for the long-term evolution process.
-More generally, given the the general genetic and demographic assumptions above, the fact that the ESS condition alone can describe the end points of long-term evolution is a mathematical demonstration of the importance of natural selection on phenotypes in determining long-term evolutionary trajectories.
-Thus, under the right assumptions, Result \ref{res:twolocus} provides a robust mathematical justification for the use of the ESS method in building evolutionary models.
+An important conclusion from Result \ref{res:twolocus} is that it shows that ESSs determined by the condition \eqref{eq:ess:def} of Maynard Smith and Price \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} are stable states of the long-term evolutionary process.
+Assuming that long-term evolution is a reasonable model of phenotypic evolution, Result \ref{res:twolocus} demonstrates the importance of natural selection on phenotypes in determining the course of phenotypic evolution and provides a robust mathematical justification for the use of the ESS method in building phenotypic evolution models.
+How appropriate the long-term evolution process is as a model of phenotypic evolution depends on its assumptions, which are summarized below from the references cited above in developing Result \ref{res:twolocus}:
+\begin{itemize}[nosep]
+  \item Phenotypes are probabilities of selecting among \textit{two} behaviors
+  \item Individuals interact in pairs that form randomly.
+  \item Populations are large enough to ignore stochastic forces including genetic drift.
+  \item Mutation is infrequent enough so that at most two genotypes occur at a time.
+  \item Phenotypes are determined by two (or more) haploid or diploid loci.
+  \item Fitness (i.e., payoff) is a linear function of phenotype.
+\end{itemize}
+While many natural populations deviate significantly from many if not all of the above assumptions, they may not be as restrictive as they seem for a few reasons.
+First, it is an open theoretical question as to whether the long-term evolution process occurs under more relaxed assumptions.
+Second, additional assumptions may relax some of the above assumptions in the long-term process.
+For example, weak selection on phenotypes allows for the analysis of nonlinear function functions and stochastic forces due to finite population effects and reveals that the ESS condition still describes the endpoints of long-term evolution \cite{Rousset:Billiard:2000,Rousset:2004,VanCleve:2015}.
+Finally, small enough deviations from the above assumptions may only induce small deviations from long-term evolution predictions from ESS analysis.
+Thus, ESS analysis might still be useful as a first approximation in more complex scenarios outside the above assumptions.
+The precise limits of the long-term evolution process and ESS analysis as predictive tools for phenotypic evolution remains a fruit area for future  mathematical, computational, and empirical research.
+
+%% do I put the convergence issue here? maybe in discussion!
 
@@
 
-While Result \ref{res:twolocus} emphasizes the role of the ESS natural selection on phenotypes in the long-term evolutionary process, the external stability approach used to prove \ref{res:twolocus} reveals a more general viewpoint where ESSs can depend in important ways on the genetic and demographic parameters of a population.
+While Result \ref{res:twolocus} emphasizes the role of the ESS and natural selection on phenotypes in phenotypic evolution, the external stability approach used to prove \ref{res:twolocus} reveals a more general viewpoint where ESSs can depend in important ways on the genetic and demographic parameters of a population.
 External stability analysis assumes that the population is at some genetic equilibrium where new mutant alleles can invade at low frequency and looks for a genetic equilibrium and its resulting phenotype that resists invasion by any mutant allele (within a set of possible phenotypic effects).
@@
 Since each individual belongs to a specific class, we actually track how many mutant individuals $k_{s}$ are in each class $s$ in a group or $\vec{k} = (k_{1},\ldots,k_{\numc})$ where $k = \sum_{s} k_{s}$.
+The number of possible states for $\vec{k}$ increases quickly with the group size $n$ and the number of classes $\numc$ but is finite.
+For simplicity of notation, we assume there is an ordering of these states so that we can use use $\vec{k}$ to index vectors and matrices below.
 When the mutant phenotype is rare, growth of the mutant population can be modeled using a matrix population model \cite{Caswell:2006} or multitype branching process \cite{Kimmel:Axelrod:2015} with a projection matrix $\vec{A}(\vec{y}, \vec{x}) = [a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x})]$ where the $a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x})$ element measures the expected number of groups with $\vec{k}'$ mutants with phenotype $\vec{y}$ that result from the reproduction occurring in a group with $\vec{k}$ mutants in a population resident for phenotype $\vec{x}$.
@@
 \end{equation}
-where the associated eigenvector $\vec{u}(\vec{y}, \vec{x})$ measures the relative frequency of groups with $\vec{k}$ mutants when the invasion process has stabilized.
+where the associated eigenvector $\vec{u}(\vec{y}, \vec{x})$ has elements $u_{\vec{k}}$ that measure the relative frequency of groups with $\vec{k}$ mutants when the invasion process has stabilized.
 The elements $a_{\vec{k}'\vec{k}}(\vec{y}, \vec{x})$ of the projection matrix can be rewritten in terms of an individual (relative) fitness function $\omega_{s'\!s\vec{k}}(\vec{y}, \vec{x})$, which measures the number of class $s'$ mutant offspring in the population in the next generation produced by a class $s$ mutant individual in a group with $\vec{k}$ mutants.

### 2023-01-16 rev1_cover_letter.tex
more edits

@@
 
+Ah, apologies.
+The $u_{j}$ are elements of the eigenvector mentioned above equation 2 and I now spell that out more clearly.
+I have also changed notation for the payoff matrix to $W$ to make things clearer.
+
 \begin{reviewerquote}
@@
 Can you repeat/list the assumptions necessary for Result 1 to hold?
+\end{reviewerquote}
+
+I have reworked those sentences and have added a list of the assumptions after Result 1 with some discussion of how they might be relaxed.
+
+\begin{reviewerquote}
   - I guess $u_k(y,x)$ on line 305 is an element in the vector $u(y,x)$ from line 292?
+\end{reviewerquote}
+
+Yes, I now specify that in the text.
+
+\begin{reviewerquote}
   - What are the dimensions of the matrix $A(y,x)$ in line 287?
@@
 If $A$ contains all possible $k$ and $k’$ vectors, its dimensions must quickly get very large?
+\end{reviewerquote}
+
+Yes, the dimensions of the $\mathbf{A}$ matrix are all possible $\vec{k}$ and $\vec{k}$ vectors and the size quickly gets very.
+It may not be the most computationally efficient formalism but works well enough as a conceptual approach.
+The notation works by assuming that one creates an ordering for the possible states in $\vec{k}$ and then indexes the matrices and vectors by that ordering.
+I now specify this in the third paragraph of section 6.
+
+\begin{reviewerquote}
   - Really enjoyed the section on variable environments.
+\end{reviewerquote}
+
+Excellent, I'm glad!
+
+\begin{reviewerquote}
 Minor comments:
 - Line 53: I’m not sure what it means for a method to be orthogonal to issues
+\end{reviewerquote}
+
+\begin{reviewerquote}
 Typos:

### 2023-01-28 manuscript.tex
omg plz finish soon

@@
 It wasn't until 1973 when John Maynard Smith and George Price introduced the evolutionarily stable strategy (ESS) and applied it to the study of animal behavior \cite{Maynard-Smith:Price:1973} that biologists more widely came to appreciate the relevance and utility of game theoretic concepts and tools for questions in evolution biology and ecology.
-Specifically, Maynard Smith and Price posited that individual fitness could be viewed as an analog of the game-theoretic notion of ``utility'', which is the quantity that measures what agents optimize in pursuit of their objectives \cite{Myerson:1991}.
+Specifically, Maynard Smith and Price posited that individual fitness could be viewed as an analog of the game-theoretic notion of ``utility'' or ``payoff'', which is the quantity that measures what agents optimize in pursuit of their objectives \cite{Myerson:1991}.
 Viewed this way, the ESS is a refinement of the famous Nash equilibrium \cite{Nash:1950}.
@@
 As I will describe below, evolutionary theorists have since realized (including Maynard Smith himself \cite[p.
-33]{MaynardSmith:1978}) that the ESS method is really orthogonal to issues of individual vs.
+33]{MaynardSmith:1978}) that the issue of the applicability of the ESS method is really orthogonal to issues of individual vs.
 group vs.
@@
 Fisher's ``fundamental theorem of natural selection'' (FTNS) \cite{Fisher:1930} established a mathematical expression of the importance of natural selection.
-The FTNS states that the increase in the mean fitness of a population is equal to the genetic variance in fitness and thus seems to imply that populations always become better adapted to their environments (since genetic variance is always non-negative) and even that fitness is maximized over the long term.
+The FTNS states that the increase in the mean fitness of a population is equal to the genetic variance in fitness and thus seems to imply that populations always become better adapted to their environments (since genetic variance is always nonnegative) and even that fitness is maximized over the long term.
 It is hard to overstate the importance of the FTNS in shaping the direction of evolutionary theory.
@@
 He proposed the ESS method developed by Price and himself as the appropriate tool for predicting phenotypic evolution and particularly for social traits.
-While Maynard Smith acknowledged that ESS models make a number of biological assumptions including that traits have a simple genetic basis and that appropriate genetic variation exits \cite{MaynardSmith:1978}, he believed that the limitations imposed by these assumptions are biologically reasonable.
+While Maynard Smith acknowledged that ESS models make a number of biological assumptions including that traits have a simple genetic basis and that appropriate genetic variation exists \cite{MaynardSmith:1978}, he believed that the limitations imposed by these assumptions are biologically reasonable.
 This view that simplifying away genetic complexities or constraints is generally reasonable was termed the ``phenotypic gambit'' by Alan Grafen \cite{Grafen:1984} and came to dominate evolutionary theory in behavioral and evolutionary ecology \cite{Houston:McNamara:1999}.
@@
 Question \ref{q:II} arises here and asks whether the ESS method can be modified to accommodate other evolutionary forces such as mutation and recombination.
-I will argue below that the ESS method is more general than originally imagined by Maynard Smith and Price \cite{Maynard-Smith:Price:1973} in that in can in fact be incorporated into an evolutionary framework where natural selection and other evolutionary forces combine to shape the short and long-term evolution of traits.
+I will argue below that the ESS method is more general than originally imagined by Maynard Smith and Price \cite{Maynard-Smith:Price:1973} in that it can in fact be incorporated into an evolutionary framework where natural selection and other evolutionary forces combine to shape the short and long-term evolution of traits.
 
@@
 
-The definition of an ESS is relatively simple and only involves a set of phenotypes, a measure of fitness, and a stability condition.
+The definition of an ESS is relatively simple and only involves a set of phenotypes, a measure of fitness (the evolutionary measure of payoff), and a stability condition.
 Suppose that two individuals interact and the phenotypes that characterize their behavior in the interactions are measured with the variables $x$ and $y$.
@@
 For the purpose of explaining the theory more concretely, assume that individuals can choose one of two possible behaviors in the interaction, $C$ and $D$.
-The phenotypes $x$ and $y$ measure the probability that individuals in an interaction with those phenotypes choose behavior $C$, and $1-x$ and $1-y$ are the probabilities those individuals choose behavior $D$ instead (probabilistic phenotypes are called mixed strategies in game theory \cite{Osborne:Rubinstein:1994}).
+The phenotypes $x$ and $y$ measure the probability that individuals in an interaction with those phenotypes choose behavior $C$, and $1-x$ and $1-y$ are the probabilities those individuals choose behavior $D$ instead (probabilistic phenotypes are called \textit{mixed strategies} in game theory \cite{Osborne:Rubinstein:1994}).
 For example, the behavior $C$ could be cooperation in the interaction and $D$ could be a lack of cooperation.
@@
 
-\section{How is ESS related to population-genetic dynamics?}
-
-Notably, the conditions in \eqref{eq:ess:def} do not reference the genetic basis of the trait $x$ nor do they describe how the frequency or mean value of the trait changes over time.
-Population genetics models, on the other hand, are built to measure the dynamical process of evolution via the combined effects of natural selection and segregation, recombination, mutation, genetic drift, and other evolutionary forces \cite{Crow:Kimura:1970,Ewens:2004}.
-A natural question then is how the long-run behavior of a population genetic model (e.g., its equilibrium genotypes and induced phenotypes) compares to the equilibrium phenotypes obtained from an analogous ESS model.
-This question was tackled by mathematical population biologists beginning in the late 1970s and through 1980s and 1990s \cite[e..g,][]{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Eshel:1982,Hofbauer:Schuster:1982,Cressman:Hines:1984,Eshel:Feldman:1984,Lessard:1984,Cressman:1988,Cressman:Hofbauer:1996,Eshel:1996,Hammerstein:1996,Weissing:1996}.
-
-Early mathematical analyses began by using the simplest population genetic model, which is one that assumes a \textit{haploid asexual} population with two genotypes, one with phenotype $x=1$ that chooses $C$ with certainty and one with phentoype $x=0$ that chooses $D$ with certainty.
-There are four possible interacting pairs of phenotypes in this scenario yielding four fitnesses that can be collected into a so-called payoff matrix $W$:
-\begin{equation*}
+To make the ESS method more concrete, consider a scenario where $C$ and $D$ correspond to cooperating and doing nothing, respectively.
+We might start with the two deterministic or \textit{pure strategies}: $x=1$ individuals always cooperate and $x=0$ individuals never cooperate.
+There are four possible interaction pairs and their payoffs can be collected into the payoff matrix
+\begin{equation}
+  \label{eq:payoff:matrix}
   W =
@@
   \end{pmatrix} \, .
-\end{equation*}
+\end{equation}
+Individuals benefit from having a partner who cooperates, which translates into $\w(1,1) > \w(1,0)$ and $\w(0,1) > \w(0,0)$.
+Since cooperation is personally costly, an individual has less payoff when it cooperates than when it doesn't given its partner also doesn't cooperate, or $\w(1,0) < \w(0,0)$.
+Moreover, that cost is pervasive enough so that cooperation is more costly than noncooperation even when the partner does cooperate, or $\w(1,1) < \w(0,1)$.
+Finally, two individuals that cooperate do better than two individuals that do not, or $\w(1,1) > \w(0,0)$.
+These payoff inequalities can be summarized as $\w(0,1) > \w(1,1) > \w(0,0) > \w(1,0)$ and they characterize the famous prisoner's dilemma game \cite{Rapoport:Chammah:1965}, which often involves public goods.
+Applying the ESS conditions \eqref{eq:ess:def} shows that noncooperation is the ESS or $\ess{x}=0$ for the prisoner's dilemma.
+For an example of a mixed-strategy ESS where $0 < \ess{x} < 1$, assume that the payoffs obey the conditions $\w(0,1) > \w(1,1) > \w(1,0) > \w(0,0)$, which characterize a hawk-dove \cite{Maynard-Smith:Price:1973,Maynard-Smith:Parker:1976} or snowdrift game \cite{Sugden:1986}.
+The only change from the prisoner's dilemma conditions is the reversal of the last inequality so that now $\w(1,0) > \w(0,0)$.
+This inequality indicates that cooperation in the hawk-dove game yields enough payoff to the cooperator so that both individuals prefer that at least one of them cooperates.
+A hawk-dove game might describe two individuals foraging for resources where its beneficial for at least one individual to expend the effort to search.
+Applying the ESS conditions \eqref{eq:ess:def} reveals that neither pure strategy $x=1$ nor $x=0$ is an ESS.
+Rather, the hawk-dove game has a mixed strategy ESS where individuals cooperate with probability $\ess{x} = (\w(1,0) - \w(0,0))/(\w(1,0) - \w(0,0) + w(0,1) - \w(1,1))$.
+
+\section{How is ESS related to population-genetic dynamics?}
+
+The ESS conditions in \eqref{eq:ess:def} do not reference the genetic basis of the trait $x$ nor do they describe how the frequency or mean value of the trait changes over time.
+Population genetics models, on the other hand, are built to measure the dynamical process of evolution via the combined effects of natural selection and segregation, recombination, mutation, genetic drift, and other evolutionary forces \cite{Crow:Kimura:1970,Ewens:2004}.
+A natural question then is how the long-run behavior of a population genetic model (e.g., its equilibrium genotypes and induced phenotypes) compares to the equilibrium phenotypes obtained from an analogous ESS model.
+This question was tackled by mathematical population biologists beginning in the late 1970s and through 1980s and 1990s \cite[e..g,][]{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Eshel:1982,Hofbauer:Schuster:1982,Cressman:Hines:1984,Eshel:Feldman:1984,Lessard:1984,Cressman:1988,Cressman:Hofbauer:1996,Eshel:1996,Hammerstein:1996,Weissing:1996}.
+
+Early mathematical analyses began by using the simplest population genetic model, which is one that assumes a \textit{haploid asexual} population with two genotypes.
+Assume the two genotypes produce the phenotypes $x=1$ that chooses $C$ with certainty and $x=0$ that chooses $D$ with certainty, respectively.
+The fitness of each of the four pairs can is given in he payoff matrix $W$ in \eqref{eq:payoff:matrix}.
 Suppose that $p$ measures the frequency of the phenotype $x=1$ that always chooses $C$.
@@
 The equilibria of this model are either fixation of one of the two genotypes resulting in a population of individuals that always chooses $C$, $p=1$, or always choose $D$, $p=0$ or a polymorphism of both genotypes, $0 < p < 1$, resulting in a population where some choose $C$ and some choose $D$.
-The analogous ESS model of this scenario applies condition \eqref{eq:ess:def} to the four possible fitness values in the payoff matrix $W$ and determines which phenotype, $x=1$ or $x=0$, is ESS or whether a mixed strategy (i.e., $0 < x < 1$) is ESS.
+The analogous ESS model of this scenario applies condition \eqref{eq:ess:def} to the four possible fitness values in the payoff matrix $W$ and determines which phenotype, $x=1$ or $x=0$, is ESS or whether a mixed strategy ($0 < x < 1$) is ESS.
 These early analyses showed that the values of $x$ that are an ESS are also values of $p$ that produce stable equilibria in the haploid asexual population genetic model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Hofbauer:Sigmund:1998}.
@@
 Work by Liberman \cite{Liberman:1988} on more than two loci likely extends Result \ref{res:twolocus} to any finite number of loci.
-
 An important conclusion from Result \ref{res:twolocus} is that it shows that ESSs determined by the condition \eqref{eq:ess:def} of Maynard Smith and Price \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} are stable states of the long-term evolutionary process.
 Assuming that long-term evolution is a reasonable model of phenotypic evolution, Result \ref{res:twolocus} demonstrates the importance of natural selection on phenotypes in determining the course of phenotypic evolution and provides a robust mathematical justification for the use of the ESS method in building phenotypic evolution models.
-How appropriate the long-term evolution process is as a model of phenotypic evolution depends on its assumptions, which are summarized below from the references cited above in developing Result \ref{res:twolocus}:
-\begin{itemize}[nosep]
-  \item Phenotypes are probabilities of selecting among \textit{two} behaviors
-  \item Individuals interact in pairs that form randomly.
-  \item Populations are large enough to ignore stochastic forces including genetic drift.
-  \item Mutation is infrequent enough so that at most two genotypes occur at a time.
-  \item Phenotypes are determined by two (or more) haploid or diploid loci.
-  \item Fitness (i.e., payoff) is a linear function of phenotype.
-\end{itemize}
+
+\subsection{Assumptions of long-term evolution}
+
+The degree to which the ESS is a reasonable prediction for long-term phenotypic evolution depends on assumptions made above in developing the long-term evolution process.
+These assumptions are\footnote{For additional discussion of these assumptions, see \cite{Eshel:1996,Hammerstein:1996,Weissing:1996}}:
+\begin{enumerate}[nosep]
+  \item Phenotypes are probabilities of selecting among \textit{two} behaviors or strategies. \label{assume:twostrat}
+  \item Fitness (i.e., payoff) is a linear function of phenotype. \label{assume:linear}
+  \item Individuals interact in pairs that form randomly. \label{assume:pairs}
+  \item Populations are large enough to ignore stochastic forces including genetic drift. \label{assume:large-pop}
+  \item Mutations are infrequent enough so that a new mutation reaches a short-term equilibrium before additional mutations occur. \label{assume:weak-mut}
+  \item Phenotypes are determined by two (or more) haploid or diploid loci. \label{assume:ploidy}
+\end{enumerate}
+Assumption \ref{assume:twostrat} that there are two behaviors or strategies is particularly important.
+In practice, games like rock paper scissors \cite{Taylor:Jonker:1978,Zeeman:1980,Weissing:1991,Hofbauer:Sigmund:1998} involving three or more strategies may have more complex long-term evolutionary dynamics resulting in equilibria that may or may not be ESSs.
+Support for this possibility comes from examples of three strategy interactions like rock paper scissors where the ESS strategy is actually an internally \textit{unstable} equilibrium \cites{Friedman:1991,Weissing:1991}.
+Weissing sees this issue as a potentially serious limitation for using ESS to describe long-term phenotypic evolution \cite{Weissing:1996}.
+Nevertheless, much of the mathematical machinery developed for analyzing external stability can be generalized to three or more strategies \cite[e.g.,][]{Eshel:Feldman:1984,Weissing:1996,Eshel:Feldman:1998,Hofbauer:Sigmund:1998}, and ESS is an appropriate determinant for stability for some three strategy population genetic models \cite[e.g., single-locus three-allele models in continuous time][]{Cressman:Hofbauer:1996}.
+Thus, future theoretical work might yield conditions under which the two strategy assumption can be relaxed.
+
+The linear fitness function assumption \ref{assume:linear} is also important.
+It arises naturally from assuming that the fitness of an individual using a mixed strategy phenotype is the expected fitness that individual would obtain using pure strategies with a probability given by the mixed strategy.
+However, phenotypes do not have to be probabilities of using pure strategies and can instead involve continuous quantities like body size or metabolic rate; fitness can easily a nonlinear function of such phenotypes.
+Studying nonlinear fitness functions, Eshel, Taylor, and others \cite{Eshel:Motro:1981,Eshel:1983,Taylor:1989,Christiansen:1991} showed that evolutionary stability is more complex than simply the ability of a population with mean phenotype $\ess{x}$ to resist invasion by mutant genotypes with phenotypes not equal to $\ess{x}$.
+
+Some of these of these assumptions may be less restrictive than others.
+For example, assumption \ref{assume:ploidy} may not be very limiting as many organisms are haploid or diploid and ESS models of other genetic systems like haplodiploidy can be studied by adding demographic structure using matrix population models \cite{Taylor:1990,Taylor:Frank:1996,Frank:1998} like the one described below in section \ref{sec:IF:group-structure} on ESS and inclusive fitness.
+Assumptions \ref{assume:large-pop} and \ref{assume:weak-mut}
+
 While many natural populations deviate significantly from many if not all of the above assumptions, they may not be as restrictive as they seem for a few reasons.
@@
 The precise limits of the long-term evolution process and ESS analysis as predictive tools for phenotypic evolution remains a fruit area for future  mathematical, computational, and empirical research.
+
+\subsection{Convergence of long-term evolution}
 
@@
 \section{Inclusive fitness, group-structured populations, and ESS}
+\label{sec:IF:group-structure}
 
@@
 
-One of the important features of Result \ref{res:twolocus} is that shows how under some genetic and demographic scenarios (i.e., a large, randomly mating population in a constant environment) an ESS phenotype maybe independent of some genetic parameters like the recombination rate between the loci affecting the phenotype.
+One of the important features of Result \ref{res:twolocus} is that it shows how under some genetic and demographic scenarios (i.e., a large, randomly mating population in a constant environment) an ESS phenotype maybe independent of some genetic parameters like the recombination rate between the loci affecting the phenotype.
 Another example of this kind of result is the ``reduction principle'' of Feldman, Liberman, and colleagues \cite{Feldman:Liberman:1986,Liberman:Feldman:1986,Liberman:Feldman:1986a,Liberman:Feldman:1989,Altenberg:Liberman:2017}, which says that for large, randomly mating populations with viability selection in a constant environment, the only uninvadable or unbeatable rates of recombination, mutation, and migration are the lowest possible rates (typically zero).
@@
 \end{document}
+
+# Definitions
+
+Externally-stable set
+- Eshel: stable short-term equilibria invariant under long-term evolution
+
+Phenotypically stable
+- Eshel/Hammerstein: phenotype is generated by externally-stable set
+
+Long-term stable
+- Eshel (1998): nearby phenotypes converge under long-term evolution to this phenotype
+- Weissing (1996): same as phenotypically stable
+- Hofbauer/Sigmund: same as phenotypically stable
+
+Strategically stable
+- Hofbauer/Sigmund: same as phenotypically stable (but focus on one-locus model)
+
+# Theorems using Eshel's terminology:
+
+Weissing (1996, Result 6, linear fitness in first component) / Hofbauer Sigmund (1998, Theorem 22.4.3, linear fitness):
+An internally stable phenotypic monomorphism is phenotypically stable if it is ESS
+- not clear in the proof why it must be phenotypic monomorphism
+
+Weissing (1996, Result 6+7, linear fitness) / Hofbauer Sigmund (1998, Theorem 22.4.4, linear fitness):
+An internally stable genetic monomorphism is phenotypically stable if and only if it generates ESS phenotype
+
+Eshel et al. (1998, Theorem 1, linear fitness):
+If internally stable two-locus equilibrium generates phenotypically stable phenotype, then it is ESS
+
+Eshel et al. (1998, Theorem 4, linear fitness):
+If internally stable two-locus equilibrium generates a phenotype that is ESS, it is phenotypically stable
+- requires equilibrium to have positive frequency of double homozygote that codes for ESS (hence some sense of genetic monomorphism)
+
+Eshel et al. (1998, Theorem 3, linear fitness):
+Two strategy ESS that is not best response is also long-term stable
+
+# Nachbar and Friedman papers to look at
+
+Nachbar 1990. “Evolutionary” selection dynamics in games: Convergence and limit properties
+Nachbar 1991. A comment on ‘evolution in economic games’
+Friedman 1991. Evolutionary games in economics
+Friedman 1998. On economic applications of evolutionary game theory

### 2023-01-28 rev1_cover_letter.tex
omg plz finish soon

@@
 
-\newenvironment{reviewerquote}{\begin{quote}\color{DarkBlue}\itshape}{\end{quote}}
+\newenvironment{reviewerquote}{\begin{quote}\color{DarkBlue}\bfseries}{\end{quote}}
 \newenvironment{response}{}{}
@@
 \begin{reviewerquote}
-Confusion/clarification/notational questions:
+Confusion/clarification/notational questions:\\
 - You are using $x$ and $y$ to denote both the phenotypes, and their frequencies, and I’m not always sure which one you mean.
@@
 It would help me a lot if you used different symbols for these two things, e.g.
-maybe $p_x$ and $p_y$ for the frequencies?
+maybe $p_x$ and $p_y$ for the frequencies?\\
 - Line 135: So now we have x denoting both the phenotype and the frequency of the phenotype, and then we have $x_{ij}$ denoting mating preference of genotype $A_i$ $A_j$.
@@
 \begin{reviewerquote}
-Minor comments:
+Minor comments:\\
 - Line 53: I’m not sure what it means for a method to be orthogonal to issues
@@
 
+Corrected.
+
 \begin{reviewerquote}
-Typos:
+Typos:\\
 - Line 87: exits -> exists
-- Line 97: in that in can in
-- Line 155 an genetic -> a genetic
+- Line 97: in that in can in\\
+- Line 155 an genetic -> a genetic\\
 - Line 347 “One of the important features of Result 1 is that shows how”
 \end{reviewerquote}
+
+Fixed.
 
@@
 \begin{reviewerquote}
-
 This manuscript attempts to review the 50-year history of the ESS concept.
@@
 Although I agree with some of the take-home messages of the manuscript, I am not entirely convinced that Van Cleve has succeeded in his mission.
+\end{reviewerquote}
 
+Thanks for the critical comments!
+I think indeed the task is a big one but I hope the revisions have improved the manuscript.
+The review certainly can't cover the whole history but I hope its captured a reasonable slice.
+
+\begin{reviewerquote}
 From the start, the ESS history was multi-faceted.
@@
 
-I mention all this because Von Cleve starts talking about evolutionary “game theory”, while game aspects (like strategy thinking) play a minor role in his review.
+I mention all this because Van Cleve starts talking about evolutionary “game theory”, while game aspects (like strategy thinking) play a minor role in his review.
 He is not really interested in behavioural interactions, but in generalizing the ESS concept to more complicated settings.
@@
 In fact, Maynard Smith’s derivation of the ESS conditions (1a) and (1b) only makes sense for two-person matrix games, as in this case the fitness function w(x,y) is linear in both components.
+\end{reviewerquote}
+
+There is some very interesting history here that I was not aware of. The reviewer has a unique perspective that I hope they write up soon!
+
+With respect to the two-person matrix game aspect, I now make the introduction of conditions (1a) and (1b) refer specifically to a pairwise game with to discrete strategies where the phenotype is the probability of selection among the discrete strategies (i.e., a a mixed strategy).
+I also introduce both the prisoner's dilemma and the snowdrift/hawk-dove game as examples to give the reader more context going into the more theoretical sections.
+
+I realize I don't focus on the ``strategy thinking'' aspect too much, though there is a little more of that aspect with the prisoner's dilemma and hawk-dove examples.
+I am writing from my background in population genetics where I learned about ESS as is related to general questions related to the dynamics of genetic and phenotypic evolution.
+As the reviewer says, there are many facets to the development of ESS, theory and I think this review is aiming for the evolutionary dynamic aspects and leaving the strategic thinking perspective to other efforts.
+I think this isn't an unprecedented perspective either.
+For example, Hofbauer and Sigmund in their 1998 book frame ESS in general around a population dynamic perspective.
+
+\begin{reviewerquote}
   Various authors later “corrected” this “mistake” and derived more general ESS concepts for non-linear fitness functions.
-It is a pity that all this is not mentioned by Van Cleve.
-Indeed, non-linearities arise, for example, in “games between relatives”.
-How to deal with this (and, hence, how to capture kin selection in the ESS formalism) led to a vivid exchange of arguments between Grafen and Maynard Smith in the late 1970s and the early 1980s.
-Grafen pointed out that his approach (corresponding to an “actor-modulated approach”) leads to quite different outcomes than Maynard Smith’s inclusive fitness approach.
-As kin selection is a central theme of Van Cleve’s manuscript, I would have expected to hear about these early debates.
-
-Anyway, it is now quite obvious that Maynard Smith would have designed the ESS conditions quite differently if he had considered the possibility of a non-linear fitness function.
+It is a pity that all this is not mentioned by Van Cleve...Anyway, it is now quite obvious that Maynard Smith would have designed the ESS conditions quite differently if he had considered the possibility of a non-linear fitness function.
 It soon became clear that such functions can have quite different evolutionary properties, like “Garden of Eden” scenarios where an ESS is not attainable, or “evolutionary branching” scenarios where the population converges to an evolutionarily unstable strategy.
@@
 It would have been very interesting to hear Van Cleve’s position regarding this reinterpretation of “ESS history”.
+\end{reviewerquote}
 
+Yes, I have largely left out reviewing adaptive dynamics, which is not directly intentional per se but more because I wanted to build around the ESS condition in the original 1973 paper as the Phil Trans issue commemorates 50 years since that paper.
+However, adaptive dynamics is of course extremely important to evolutionary game theory and to the concept of invasion fitness.
+I've added XXX...
+%%% Add something about convergence stability to discussion.
+
+As to whether invasion fitness in adaptive dynamics is more general than what I discuss here, I would guess the reviewer is thinking of cases where short-term evolution reaches a non-equilibrium attractor like a cycle or chaotic attractor and invasion fitness must then really be measured in the sense of a Lyapunov exponent.
+I agree those cases are important but I do not mention them so as to keep the scope of the review from growing too much.
+It should be noted too that the assumptions and tools employed in typical deterministic and stochastic adaptive dynamics models ignore the non-equilibrium cases too as they are difficult to handle in general.
+One of those assumptions is often, as I discuss in reference to long-term evolution, infrequent mutation, which when combined with finite population size, leads to at most two genotypes existing at a time (see the Champagnat 2006 ref) and one of them either going extinct or fixing before the arrival of the next mutation.
+
+Finally, I totally agree that the second ESS condition from the 1973 paper is really the convergence stability condition.
+My review wasn't structured to synthesize the original ESS conditions with the stability and convergence conditions from adaptive dynamics but this would be the basis from a great review or synthesis article.
+%% Add something here maybe if I write convergence stability paragraph?
+
+\begin{reviewerquote}
+  Indeed, non-linearities arise, for example, in “games between relatives”.
+How to deal with this (and, hence, how to capture kin selection in the ESS formalism) led to a vivid exchange of arguments between Grafen and Maynard Smith in the late 1970s and the early 1980s.
+Grafen pointed out that his approach (corresponding to an “actor-modulated approach”) leads to quite different outcomes than Maynard Smith’s inclusive fitness approach.
+As kin selection is a central theme of Van Cleve’s manuscript, I would have expected to hear about these early debates.
+\end{reviewerquote}
+
+I think the reviewer may be referring to Maynard Smith (1978, Ann Rev Ecol Syst), Grafen (1979, Anim Behav), and Hines and Maynard Smith (1979, JTB).
+
+\begin{reviewerquote}
   The middle part of Van Cleve’s review (Sections 3 and 4) contains some interesting observations, but the results are often stated in a relatively sloppy manner (i.e., without mentioning the conditions under which they hold).
+\end{reviewerquote}
+
+I agree that I need to be clearer about the conditions necessary the results in sections 3 and 4 to hold.
+I've now added a bulleted list at the end of section 4 that lists all of the conditions and also discuss how restrictive they may or may not be.
+
+\begin{reviewerquote}
   In fact, the references referred to in the review all make quite limiting assumptions, such as a linear fitness function (in both components) or a system with only two phenotypes.
@@
 Obviously, this limitation of Result 1 should have been mentioned.
+\end{reviewerquote}
+
+Yes, this limitation is important. I now specify it clearly as the first assumption in the bulleted list at the end of section 4.
+
+\begin{reviewerquote}
   In fact, various counter-examples to both parts of Result 1 were constructed in the evolutionary branch of economic game theory (most notably by Friedman and Nachbar).
@@
 Like them, I had to conclude that long-term evolution does not “save” the ESS concept when more than two strategies are around.
+\end{reviewerquote}
 
+%% mention this more clearly at the end of section 4 and/or in the discussion
+
+\begin{reviewerquote}
   But also the case of only two phenotypes is less clear than presented by Van Cleve.
@@
 Accordingly, I am much less convinced than Van Cleve that “long-term evolution” has been shown to “save” the ESS concept.
+\end{reviewerquote}
 
+%% ugh. ok take care of this
+
+\begin{reviewerquote}
   In contrast to the earlier sections, which are broadly accessible, Sections 6 and 7 are technically more sophisticated and will most likely only be understandable to experts.
@@
 1 without discussing the underlying assumptions in sufficient depth and detail.
+\end{reviewerquote}
+
+I agree that section 6 is more technical. I have added XXX..
+%% add here
+
+I have tried to keep Section 7 to a minimum of technical depth.
+I do not agree that its understandable to only technically sophisticated readers.
+However, it is relevant to add some of the assumptions for the modifier models discussed and I now do that in XXXX
+%% add here
+
+\begin{reviewerquote}
   The “reduction principle” is a wonderful tool but I would have liked to see a more critical discussion of its applicability.
+\end{reviewerquote}
 
+I mention the reduction principle since it uses external stability to show the general evolutionary principle that selection in a constant generates purifying selection for the highest fitness genotype and against genetic variation created by other genotypes.
+Varying environments don't exhibit the reduction principle since genetic or phenotypic variation can be adaptive in these cases.
+I don't see how discussing technical criticisms of the reduction principle advances the purpose of the section, which is to show how the ESS concept as it relates to external stability can study the evolution of genetic and demographic parameters in the case of varying environments.
+
+\begin{reviewerquote}
 Generally, the manuscript is well-written.
 But as mentioned above, there is a certain imbalance in this manuscript: no (restrictive) assumptions are stated in the first sections (although the experts would have liked to see them spelt out explicitly), while the treatment in the later sections is only understandable to experts in the field.
+\end{reviewerquote}
 
-\end{reviewerquote}
+I hope I have addressed these criticisms as described above and hope these changes improves the manuscript!
 

### 2023-01-29 manuscript.tex
almost done with CS section

@@
 Fitness $\w$ for the moment is simply the survival rate (i.e., viability selection) but as we'll see we can define much more general fitness measures.
-The stability condition \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} says that a phenotype $\ess{x}$ is an ESS, or is evolutionarily stable (ES), when no alternative strategy $y$ can improve upon the fitness that $\ess{x}$ receives when interacting with itself:
+The stability condition \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} says that a phenotype $\ess{x}$ is an ESS, or is evolutionarily stable (ES), when one of two scenarios occurs.
+For any alternative strategy $y$, either $y$ receives less fitness when interacting with $\ess{x}$ than $\ess{x}$ receives when interacting with itself,
 \begin{subequations}
@@
   \label{eq:ess:def:a}
-  \w(\ess{x}, \ess{x}) \ge \w(y, \ess{x})
+  \w(\ess{x}, \ess{x}) > \w(y, \ess{x}) \, ,
 \end{equation}
-for all phenotypes $y$.
-In the case that the strategy $y$ gets the same fitness interacting with $\ess{x}$ as $\ess{x}$ does with itself (equation \eqref{eq:ess:def:a} holds as an equality) then $\ess{x}$ must do better interacting with $y$ than $y$ does with itself:
+or $y$ gets the same fitness interacting with $\ess{x}$ as $\ess{x}$ does with itself and $\ess{x}$ receives more fitness interacting with $y$ than $y$ does with itself,
 \begin{equation}
   \label{eq:ess:def:b}
-  \w(\ess{x}, y) > \w(y, y)
+  \w(\ess{x}, y) > \w(y, y) \quad \text{when} \quad  \w(\ess{x}, \ess{x}) = \w(y, \ess{x}) \, .
 \end{equation}
 \end{subequations}
-The two conditions in \eqref{eq:ess:def} are equivalent to requiring that phenotype $\ess{x}$ receives higher average fitness than any alternative phenotype $y$ when $y$ is rare in the population \cite{Maynard-Smith:1974,Bishop:Cannings:1976}.
-If condition \eqref{eq:ess:def} holds as a strict inequality when $y \ne \ess{x}$, then $\ess{x}$ is a ``strict'' ESS.
+The two conditions in \eqref{eq:ess:def} are together equivalent to requiring that phenotype $\ess{x}$ receives higher average fitness than any alternative phenotype $y$ when $y$ is rare in the population \cite{Maynard-Smith:1974,Bishop:Cannings:1976}.
+The phenotype $\ess{x}$ is called a ``strict'' ESS when condition \eqref{eq:ess:def:a} holds and a ``weak ESS'' when condition \eqref{eq:ess:def:b} holds.
 
@@
 for all possible mutant $A_{\mut}$ that generate mutant phenotype $x_{\mut}$.
-This condition is in fact the condition for a strict ESS from \eqref{eq:ess:def} and can be used to prove that the following result for large, randomly mating, diploid populations with two-locus genetic systems \cite{Eshel:Feldman:1984,Eshel:1996,Eshel:Feldman:1998}:
+This condition is in fact the condition for a strict ESS from \eqref{eq:ess:def:a} and can be used to prove that the following result for large, randomly mating, diploid populations with two-locus genetic systems \cite{Eshel:Feldman:1984,Eshel:1996,Eshel:Feldman:1998}:
 \begin{result}
@@
 Assuming that long-term evolution is a reasonable model of phenotypic evolution, Result \ref{res:twolocus} demonstrates the importance of natural selection on phenotypes in determining the course of phenotypic evolution and provides a robust mathematical justification for the use of the ESS method in building phenotypic evolution models.
-
-\subsection{Assumptions of long-term evolution}
-
-The degree to which the ESS is a reasonable prediction for long-term phenotypic evolution depends on assumptions made above in developing the long-term evolution process.
-These assumptions are\footnote{For additional discussion of these assumptions, see \cite{Eshel:1996,Hammerstein:1996,Weissing:1996}}:
-\begin{enumerate}[nosep]
-  \item Phenotypes are probabilities of selecting among \textit{two} behaviors or strategies. \label{assume:twostrat}
-  \item Fitness (i.e., payoff) is a linear function of phenotype. \label{assume:linear}
-  \item Individuals interact in pairs that form randomly. \label{assume:pairs}
-  \item Populations are large enough to ignore stochastic forces including genetic drift. \label{assume:large-pop}
-  \item Mutations are infrequent enough so that a new mutation reaches a short-term equilibrium before additional mutations occur. \label{assume:weak-mut}
-  \item Phenotypes are determined by two (or more) haploid or diploid loci. \label{assume:ploidy}
-\end{enumerate}
-Assumption \ref{assume:twostrat} that there are two behaviors or strategies is particularly important.
-In practice, games like rock paper scissors \cite{Taylor:Jonker:1978,Zeeman:1980,Weissing:1991,Hofbauer:Sigmund:1998} involving three or more strategies may have more complex long-term evolutionary dynamics resulting in equilibria that may or may not be ESSs.
-Support for this possibility comes from examples of three strategy interactions like rock paper scissors where the ESS strategy is actually an internally \textit{unstable} equilibrium \cites{Friedman:1991,Weissing:1991}.
-Weissing sees this issue as a potentially serious limitation for using ESS to describe long-term phenotypic evolution \cite{Weissing:1996}.
-Nevertheless, much of the mathematical machinery developed for analyzing external stability can be generalized to three or more strategies \cite[e.g.,][]{Eshel:Feldman:1984,Weissing:1996,Eshel:Feldman:1998,Hofbauer:Sigmund:1998}, and ESS is an appropriate determinant for stability for some three strategy population genetic models \cite[e.g., single-locus three-allele models in continuous time][]{Cressman:Hofbauer:1996}.
-Thus, future theoretical work might yield conditions under which the two strategy assumption can be relaxed.
-
-The linear fitness function assumption \ref{assume:linear} is also important.
-It arises naturally from assuming that the fitness of an individual using a mixed strategy phenotype is the expected fitness that individual would obtain using pure strategies with a probability given by the mixed strategy.
-However, phenotypes do not have to be probabilities of using pure strategies and can instead involve continuous quantities like body size or metabolic rate; fitness can easily a nonlinear function of such phenotypes.
-Studying nonlinear fitness functions, Eshel, Taylor, and others \cite{Eshel:Motro:1981,Eshel:1983,Taylor:1989,Christiansen:1991} showed that evolutionary stability is more complex than simply the ability of a population with mean phenotype $\ess{x}$ to resist invasion by mutant genotypes with phenotypes not equal to $\ess{x}$.
-
-Some of these of these assumptions may be less restrictive than others.
-For example, assumption \ref{assume:ploidy} may not be very limiting as many organisms are haploid or diploid and ESS models of other genetic systems like haplodiploidy can be studied by adding demographic structure using matrix population models \cite{Taylor:1990,Taylor:Frank:1996,Frank:1998} like the one described below in section \ref{sec:IF:group-structure} on ESS and inclusive fitness.
-Assumptions \ref{assume:large-pop} and \ref{assume:weak-mut}
-
-While many natural populations deviate significantly from many if not all of the above assumptions, they may not be as restrictive as they seem for a few reasons.
-First, it is an open theoretical question as to whether the long-term evolution process occurs under more relaxed assumptions.
-Second, additional assumptions may relax some of the above assumptions in the long-term process.
-For example, weak selection on phenotypes allows for the analysis of nonlinear function functions and stochastic forces due to finite population effects and reveals that the ESS condition still describes the endpoints of long-term evolution \cite{Rousset:Billiard:2000,Rousset:2004,VanCleve:2015}.
-Finally, small enough deviations from the above assumptions may only induce small deviations from long-term evolution predictions from ESS analysis.
-Thus, ESS analysis might still be useful as a first approximation in more complex scenarios outside the above assumptions.
-The precise limits of the long-term evolution process and ESS analysis as predictive tools for phenotypic evolution remains a fruit area for future  mathematical, computational, and empirical research.
-
-\subsection{Convergence of long-term evolution}
-
-%% do I put the convergence issue here? maybe in discussion!
 
@@
 where $\omega(y, \ess{x})$ is the relative fitness of a rare mutant population with mean phenotype $y$ in a population with mean phenotype $\ess{x}$.
-The strict ESS condition in \eqref{eq:strict:relative} using relative fitness is identical to the condition for uninvadability in condition \eqref{eq:unbeatable} using invasion fitness, which means that unbeatable or uninvadable phentoypes are strict ESSs\footnote{The relationship between invasion fitness and non-strict ESSs is more complex and depends on specific details of the genetics of the population.
+The strict ESS condition in \eqref{eq:strict:relative} using relative fitness is identical to the condition for uninvadability in condition \eqref{eq:unbeatable} using invasion fitness, which means that unbeatable or uninvadable phenotypes are strict ESSs\footnote{The relationship between invasion fitness and non-strict ESSs is more complex and depends on specific details of the genetics of the population.
 For example, see the section entitled ``Long-term Stability of an ESS'' in \cite{Eshel:Feldman:1998}}.
@@
 
+\section{Assumptions of long-term evolution and ESS}
+
+The previous sections show how ESS analysis using invasion fitness can be used to predict the long-term outcomes of phenotypic evolution in scenarios with both genetic and demographic structure.
+However, the generality of these results ultimately depends on the assumptions made in developing Result \ref{res:twolocus} in section \ref{sec:long-term}, which relates the ESS condition to equilibria of the long-term evolution process.
+These assumptions are\footnote{For additional discussion of these assumptions, see \cite{Eshel:1996,Hammerstein:1996,Weissing:1996}}:
+\begin{enumerate}[nosep]
+  \item Phenotypes are probabilities of selecting among \textit{two} behaviors or strategies. \label{assume:twostrat}
+  \item Fitness (i.e., payoff) is a linear function of phenotype. \label{assume:linear}
+  \item Individuals interact in pairs that form randomly. \label{assume:pairwise}
+  \item Populations are large enough to ignore stochastic forces including genetic drift. \label{assume:large-pop}
+  \item Mutations are infrequent enough so that a new mutation reaches a short-term equilibrium before additional mutations occur. \label{assume:weak-mut}
+  \item Phenotypes are determined by two (or more) haploid or diploid loci. \label{assume:ploidy}
+\end{enumerate}
+Some of these of these assumptions are likely less restrictive than others.
+For example, assumption \ref{assume:ploidy} may not be very limiting as many organisms are haploid or diploid and ESS models of other genetic systems like haplodiploidy can be studied by adding demographic structure using matrix population models \cite{Taylor:1990,Taylor:Frank:1996,Frank:1998} like the one described below in section \ref{sec:IF:group-structure} on ESS and inclusive fitness.
+The large population size assumption \ref{assume:large-pop} also can be relaxed by using stochastic population models that allow for genetic drift and and use fixation probability as a fitness measure instead of invasion fitness \cite{Rousset:Billiard:2000,Rousset:2003,Nowak:Sasaki:2004,Lessard:2005,VanCleve:2015}.
+Assumption \ref{assume:weak-mut} is more difficult to relax since doing so allows multiple mutants to occur at a time.
+Analysis of such multi-mutant invasions is beyond the scope of the mathematical tools typically used in evolutionary analysis.
+One potential route forward for such analyses might be to follow Luo and Cooney who used non-local integro-differential partial differential equations to study the evolution of cooperation via multilevel selection \cite{Luo:2014,Luo:Mattingly:2017,Cooney:2019,Cooney:2020,Cooney:Mori:2022}.
+
+Assumption \ref{assume:twostrat} that there are two behaviors or strategies is particularly important.
+In practice, games like rock paper scissors \cite{Taylor:Jonker:1978,Zeeman:1980,Weissing:1991,Hofbauer:Sigmund:1998} involving three or more strategies may have more complex long-term evolutionary dynamics resulting in equilibria that may or may not be ESSs.
+Support for this possibility comes from examples of three strategy interactions like rock paper scissors where the ESS strategy is actually an internally \textit{unstable} equilibrium \cites{Friedman:1991,Weissing:1991}.
+Weissing sees this issue as a potentially serious limitation for using ESS to describe long-term phenotypic evolution \cite{Weissing:1996}.
+Interestingly, much of the mathematical machinery developed for analyzing external stability can be generalized to three or more strategies \cite[e.g.,][]{Eshel:Feldman:1984,Weissing:1996,Eshel:Feldman:1998,Hofbauer:Sigmund:1998}, and ESS is an appropriate determinant for stability for some three strategy population genetic models \cite[e.g., single-locus three-allele models in continuous time][]{Cressman:Hofbauer:1996}.
+Thus, future theoretical work might yield conditions under which the two strategy assumption can be relaxed.
+
+The linear fitness function assumption \ref{assume:linear} is also important restriction.
+It arises naturally from assuming that the fitness of an individual using a mixed strategy phenotype is the expected fitness that individual would obtain using pure strategies with probabilities given by the mixed strategy.
+Linear fitness functions are convenient in part because they simplify interactions involving more than two individuals: fitness in such interactions can be expressed as a linear combination of the fitness of pairwise interactions.
+This means the pairwise interaction assumption \ref{assume:pairwise} is unnecessary for linear fitness functions.
+However, phenotypes often involve continuous quantities like body size or metabolic rate instead of pure strategy probabilities and nonlinear fitness functions are the norm in these cases.
+Studying nonlinear fitness functions, Eshel, Taylor, and others \cite{Eshel:Motro:1981,Eshel:1983,Taylor:1989,Christiansen:1991,Metz:Geritz:1996,Eshel:Motro:1997,Geritz:Kisdi:1998} showed that evolutionary stability involves two qualitatively different kinds of stability.
+The first kind is the ability of a population with mean phenotype $\ess{x}$ to resist invasion by any mutant genotypes with phenotype not equal to $\ess{x}$, which is the condition captured by the strict ESS condition \eqref{eq:ess:def:a}.
+This is the classic notion of evolutionary stability evoked by verbal descriptions of the ESS concept.
+The second kind of stability is called ``convergence stability'' \cite{Christiansen:1991,Eshel:1996,Geritz:Kisdi:1998}.
+
+\section{Convergence of long-term evolution and ESS}
+
+Convergence stability refers to populations with mean phenotype $x$ near an ESS $\ess{x}$ that are invaded by mutants that bring the population closer to $\ess{x}$ and resist invasion by mutants that move the population farther from $\ess{x}$ \cite{Eshel:1983,Taylor:1989}.
+Thus, populations near a convergence stable phenotype $\ess{x}$ will evolve towards $\ess{x}$.
+If an ESS $\ess{x}$ is to really represent the outcome of a long-term evolution process, then it should be convergence stable as well as an ESS.
+The evolutionary and convergence stability conditions for continuous phenotypes and nonlinear fitness functions are expressed in terms of partial derivatives of the fitness function with respect to the mutant and resident phenotypes \cite{Eshel:1983,Taylor:1989,Geritz:Kisdi:1998}.
+However, we can still get a sense of a formal statement of the convergence stability condition without the partial derivative expressions by examining the the first inequality in the weak ESS condition \eqref{eq:ess:def:b}, $\w(\ess{x}, y) > \w(y, y)$, which says that $\ess{x}$ should invade a population with an alternative phenotype $y$.
+In a sense then, the original Maynard Smith and Price definition of an ESS \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} contained the notion of convergence stability but did not distinguish it from evolutionary stability more generally.
+Only by examining continuous phenotypes and nonlinear fitness functions did convergence stability emerge clearly as an essential and distinct condition.
+
+Given there are two kinds of stability to asses, evolutionary stability \textit{sensu} condition \eqref{eq:ess:def:a} and convergence stability, a phenotype $\ess{x}$ that is not long-term stable might fail either one of them.
+If $\ess{x}$ is not convergence stable, then a population with mean phenotype $\ess{x}$ will resist invasion by mutants with alternative phenotypes but populations with mean phenotypes near $\ess{x}$ will eventually evolve away from $\ess{x}$.
+Such ESSs are unattainable under long-term evolution and are sometimes called ``Garden of Eden'' phenotypes \cite{Hofbauer:Sigmund:1990}.
+Even more interesting are the cases where the phenotype $\ess{x}$ is convergence stable but not an ESS.
+This possibility was was not explored much before
+
 \section{Conclusions and the future of ESS analysis}
@@
 
-J.V.C. would like to acknowledge the hard work and extraordinary patience of the editors and staff who brought this special issue together.
+J.V.C.
+would like to acknowledge the hard work and extraordinary patience of the editors and staff who brought this special issue together.
+Two reviewers provided thoughtful and expert comments that have improved the manuscript.
+Erol Akçay provided intellectual and emotional support.
 

### 2023-01-30 manuscript.tex
just need final edits

@@
   In this paper, I review the history of the ESS concept and show how early uncertainty about the method lead to important mathematical theory linking ESS analysis to general population-genetic models.
-I use this theory to emphasize the link between ESS analysis and the concept of ``invasion fitness''.
+I use this theory to emphasize the link between ESS analysis and the concept of \textit{invasion fitness}.
 I give examples of how invasion fitness can measure kin selection, group selection, and the evolution of linked modifier genes in response to variable environments.
@@
 It wasn't until 1973 when John Maynard Smith and George Price introduced the evolutionarily stable strategy (ESS) and applied it to the study of animal behavior \cite{Maynard-Smith:Price:1973} that biologists more widely came to appreciate the relevance and utility of game theoretic concepts and tools for questions in evolution biology and ecology.
-Specifically, Maynard Smith and Price posited that individual fitness could be viewed as an analog of the game-theoretic notion of ``utility'' or ``payoff'', which is the quantity that measures what agents optimize in pursuit of their objectives \cite{Myerson:1991}.
+Specifically, Maynard Smith and Price posited that individual fitness could be viewed as an analog of the game-theoretic notion of \textit{utility} or \textit{payoff}, which is the quantity that measures what agents optimize in pursuit of their objectives \cite{Myerson:1991}.
 Viewed this way, the ESS is a refinement of the famous Nash equilibrium \cite{Nash:1950}.
-Maynard Smith explained the ESS concept in more detail in an important paper in 1974 \cite{Maynard-Smith:1974}, introduced the famous ``Hawk-Dove'' game in his study of asymmetric games in 1976 \cite{Maynard-Smith:Parker:1976}, and summarized the nascent field of evolutionary game theory in his now classic 1982 book \cite{MaynardSmith:1982}.
+Maynard Smith explained the ESS concept in more detail in an important paper in 1974 \cite{Maynard-Smith:1974}, introduced the famous \textit{Hawk-Dove} game in his study of asymmetric games in 1976 \cite{Maynard-Smith:Parker:1976}, and summarized the nascent field of evolutionary game theory in his now classic 1982 book \cite{MaynardSmith:1982}.
 In the 50 years since the 1973 paper, evolutionary game theory has become an essential tool in evolutionary and behavioral ecology and includes rich theoretical work that delves into foundational evolutionary and game-theoretic concepts and mathematical and simulation models that predict ESSs for many specific biological systems.
@@
 group vs.
-kin selection are about the ``units'' or ``levels'' at which selection acts and how fitness should be measured to account for selection at those levels.
+kin selection are about the \textit{units} or \textit{levels} at which selection acts and how fitness should be measured to account for selection at those levels.
 Lewontin also hints at the levels of selection question in his 1961 game theory paper when he discusses the relative merits of individual-level measures of fitness like intrinsic or Malthusian growth rate $r$ versus population-level measures like mean fitness $\mean{w}$ \cite[pp.
@@
 The levels of selection question became a major topic of study in evolutionary theory \cite[e.g.,][]{Lewontin:1970,Dawkins:1982,Wilson:Sober:1989,Maynard-Smith:Szathmary:1995,Wilson:1997,Michod:1999,Michod:2006,Szathmary:2015} and philosophy of biology \cite{Hull:1980,Brandon:1982,Damuth:Heisler:1988,Lloyd:1992,Lloyd:1994,Sober:Wilson:1994,Okasha:2006,Okasha:2016} and continues to generate substantial research \cite[e.g.,][]{Black:Bourrat:2020,Cooney:Mori:2022,Veit:2022}.
-For the purpose of explaining how an ESS can capture selection at multiple levels and among relatives or kin, I will argue below that the right measure of utility is the ``invasion fitness'' \cite{Metz:Nisbet:1992,Heino:Metz:1998} or ```lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016} of a mutant allele at a single genetic locus, which can be shown to be functionally equivalent to a measure of inclusive fitness \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020}, which is a concept originally proposed by W.D.
+For the purpose of explaining how an ESS can capture selection at multiple levels and among relatives or kin, I will argue below that the right measure of utility is the \textit{invasion fitness} \cite{Metz:Nisbet:1992,Heino:Metz:1998} or \textit{lineage fitness} \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016} of a mutant allele at a single genetic locus, which can be shown to be functionally equivalent to a measure of inclusive fitness \cite{Lehmann:Mullon:2016,Lehmann:Rousset:2020}, which is a concept originally proposed by W.D.
 Hamilton in relation to kin selection \cite{Hamilton:1964}.
@@
 The two conditions in \eqref{eq:ess:def} are together equivalent to requiring that phenotype $\ess{x}$ receives higher average fitness than any alternative phenotype $y$ when $y$ is rare in the population \cite{Maynard-Smith:1974,Bishop:Cannings:1976}.
-The phenotype $\ess{x}$ is called a ``strict'' ESS when condition \eqref{eq:ess:def:a} holds and a ``weak ESS'' when condition \eqref{eq:ess:def:b} holds.
+The phenotype $\ess{x}$ is called a \textit{strict} ESS when condition \eqref{eq:ess:def:a} holds and a \textit{weak ESS} when condition \eqref{eq:ess:def:b} holds.
 
@@
 The important idea is that the conditions under which a genetic equilibrium is stable given a fixed set of alleles or genotypes does not determine whether a new and rare allele or genotype outside of that fixed set might invade the population.
-When an equilibrium is stable given a fixed set of alleles or genotypes, it is called ``internally stable'' \cite{Eshel:1996}.
-If that equilibrium also resists invasion by new alleles or genotypes outside the fixed set, it is called ``externally stable'' \cite{Liberman:1988,Eshel:1996}.
+When an equilibrium is stable given a fixed set of alleles or genotypes, it is called \textit{internally stable} \cite{Eshel:1996}.
+If that equilibrium also resists invasion by new alleles or genotypes outside the fixed set, it is called \textit{externally stable} \cite{Liberman:1988,Eshel:1996}.
 Equilibria that are internally stable may not be externally stable.
@@
 The invasion of this new allele will lead to its fixation, which is a new internally stable equilibrium that may or may not be externally stable to further invasions depending on the the genotype-phenotype map (i.e., how the genotypes $A_{i} A_{j}$ generate fitness values $w_{ij}$).
-Given a fixed set of genotypes, the evolution of those genotypes to an internally stable equilibrium is called ``short-term evolution'' whereas the process of new mutant invasions that move the population from one internally stable equilibrium to another is called ``long-term evolution'' \cite{Eshel:1991,Eshel:1996,Hammerstein:1996,Weissing:1996,Eshel:Feldman:1998,VanCleve:2015,VanCleve:2020} and eventually leads to an externally stable equilibrium.
+Given a fixed set of genotypes, the evolution of those genotypes to an internally stable equilibrium is called \textit{short-term evolution} whereas the process of new mutant invasions that move the population from one internally stable equilibrium to another is called \textit{long-term evolution} \cite{Eshel:1991,Eshel:1996,Hammerstein:1996,Weissing:1996,Eshel:Feldman:1998,VanCleve:2015,VanCleve:2020} and eventually leads to an externally stable equilibrium.
 The process of long-term evolution assumes that mutations are infrequent enough so that a new mutation has sufficient time to either go extinct or reach a new short-term equilibrium before additional mutations occurs.
@@
 \end{equation*}
-Now, assume that a new mutant allele $A_{\mut}$ (increasing the total number of alleles at locus $A$ to $a+1$) arrives at low frequency in a population at an internally stable equilibrium $\fixp{\vec{p}}$ with ``resident'' or wild-type mean phenotype $\fixp{x} = x(\fixp{\vec{p}})$.
+Now, assume that a new mutant allele $A_{\mut}$ (increasing the total number of alleles at locus $A$ to $a+1$) arrives at low frequency in a population at an internally stable equilibrium $\fixp{\vec{p}}$ with \textit{resident} or wild-type mean phenotype $\fixp{x} = x(\fixp{\vec{p}})$.
 An individual with this new allele chooses phenotype $x$ with probability $x_{\mut jkl}$ if its genotype is $A_{\mut}B_{j}/A_{k}B_{l}$ and phenotype $y$ with probability $1 - x_{\mut jkl}$.
@@
 External stability analysis assumes that the population is at some genetic equilibrium where new mutant alleles can invade at low frequency and looks for a genetic equilibrium and its resulting phenotype that resists invasion by any mutant allele (within a set of possible phenotypic effects).
-Before the mathematical machinery of external stability was developed, Hamilton appreciated the conceptual importance of uninvadable phenotypes in his study of the sex ratio, and he called sex ratios that resisted invasion ``unbeatable'' phenotypes \cite{Hamilton:1967}.
+Before the mathematical machinery of external stability was developed, Hamilton appreciated the conceptual importance of uninvadable phenotypes in his study of the sex ratio, and he called sex ratios that resisted invasion \textit{unbeatable} phenotypes \cite{Hamilton:1967}.
 As we saw above, determining whether a phenotype can be unbeatable and its genetic equilibria externally stable involves calculating the magnitude of the eigenvalue $\eig$ associated with the linear dynamical system that approximates the dynamics of the mutant allele frequency when rare.
@@
 Thus, $\eig$ really measures the fitness of the whole mutant genetic lineage.
-The growth rate $\eig$ is often called the ``invasion fitness'' \cite{Metz:Nisbet:1992,Heino:Metz:1998}\footnote{When invasion occurs at more complex attractors like cycles, the ``invasion exponent'' is used, which is $\log |\eig|$ \cite{Rand:Wilson:1994}} and sometimes called the ``lineage fitness'' \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016,Lehmann:Rousset:2020}.
+The growth rate $\eig$ is often called the \textit{invasion fitness} \cite{Metz:Nisbet:1992,Heino:Metz:1998}\footnote{When invasion occurs at more complex attractors like cycles, the \textit{invasion exponent} is used, which is $\log |\eig|$ \cite{Rand:Wilson:1994}} and sometimes called the \textit{lineage fitness} \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016,Lehmann:Rousset:2020}.
 In effect, the invasion or lineage fitness measures the average number of offspring copies produced by a mutant allele sampled randomly from among all the genotypes in the mutant lineage \cite{Lehmann:Mullon:2016}.
@@
 \end{equation}
-Regardless of the complexity of the genetics, demography, or ecology of a population, we can in principle perform an external stability analysis\footnote{This is also referred to as an ``evolutionary invasion analysis''.} and use invasion fitness to determine whether there are unbeatable phenotypes.
+Regardless of the complexity of the genetics, demography, or ecology of a population, we can in principle perform an external stability analysis\footnote{This is also referred to as an \textit{evolutionary invasion analysis}.} and use invasion fitness to determine whether there are unbeatable phenotypes.
 
@@
 In the general case for an arbitrary genetic and demographic scenario, we assume only that we have the invasion fitness $\eig(y, \ess{x})$ of a rare mutant $y$ in the population that otherwise has mean phenotype $\ess{x}$.
-Starting with the condition for a strict ESS, $w(\ess{x}, \ess{x}) > w(y, \ess{x})$ for $y \ne \ess{x}$ from \eqref{eq:ess:def:a}, we can convert that condition to one in terms of ``relative fitness'' by dividing by the mean population fitness $w(\ess{x}, \ess{x})$ to obtain
+Starting with the condition for a strict ESS, $w(\ess{x}, \ess{x}) > w(y, \ess{x})$ for $y \ne \ess{x}$ from \eqref{eq:ess:def:a}, we can convert that condition to one in terms of \textit{relative fitness} by dividing by the mean population fitness $w(\ess{x}, \ess{x})$ to obtain
 \begin{equation}
@@
 The benefit term $b_{s'\!\sigma \leftarrow s}(\vec{y}, \vec{x})$ measures the marginal increase in the number of class $s'$ offspring produced by individuals of class $\sigma$ due to the mutant allele expression by a groupmate of class $s$.
-The invasion fitness can then be expressed as (using an ``actor-modulated'' approach) \cite{Lehmann:Rousset:2020}
+The invasion fitness can then be expressed as (using an \textit{actor-modulated} approach) \cite{Lehmann:Rousset:2020}
 \begin{equation}
@@
 Given that previous work using the Price equation has shown that kin selection and group selection partitions of fitness are simply alternative ways of expressing genotype frequency change in a population \cite{Queller:1992,Bijma:Wade:2008,Gardner:West:2007,Marshall:2011,Akcay:VanCleve:2012}, it should be possible to rewrite invasion fitness so as to group terms that constitute between-group selection, $\omega_{G,s'\!s}(\vec{y}, \vec{x})$, and those that constitute within-group selection, $\omega_{\Delta G,s'\!s}(\vec{y}, \vec{x})$.
-In fact, if we apply a ``neighbor-modulated'' approach from \cite{Queller:1992,Akcay:VanCleve:2012} to the results from \cite{Lehmann:Rousset:2020}, we obtain
+In fact, if we apply a \textit{neighbor-modulated} approach from \cite{Queller:1992,Akcay:VanCleve:2012} to the results from \cite{Lehmann:Rousset:2020}, we obtain
 \begin{equation}
@@
 This is the classic notion of evolutionary stability evoked by verbal descriptions of the ESS concept.
-The second kind of stability is called ``convergence stability'' \cite{Christiansen:1991,Eshel:1996,Geritz:Kisdi:1998}.
+The second kind of stability is called \textit{convergence stability} \cite{Christiansen:1991,Eshel:1996,Geritz:Kisdi:1998}.
 
@@
 If $\ess{x}$ is not convergence stable, then a population with mean phenotype $\ess{x}$ will resist invasion by mutants with alternative phenotypes but populations with mean phenotypes near $\ess{x}$ will eventually evolve away from $\ess{x}$.
-Such ESSs are unattainable under long-term evolution and are sometimes called ``Garden of Eden'' phenotypes \cite{Hofbauer:Sigmund:1990}.
+Such ESSs are unattainable under long-term evolution and are sometimes called ``Garden of Eden'' phenotypes because no predecessor population could evolve towards them \cite{Hofbauer:Sigmund:1990}.
 Even more interesting are the cases where the phenotype $\ess{x}$ is convergence stable but not an ESS.
-This possibility was was not explored much before
+This possibility was was not explored much before the Metz, Geritz, Kisdi, and colleagues laid out in detail how such cases should lead to the evolution of phenotypic polymorphisms as the population mean approaches $\ess{x}$ \cite{Metz:Geritz:1996,Geritz:Kisdi:1998}.
+This work spawned the field of \textit{adaptive dynamics} \cite{Dercole:Rinaldi:2008} where the emergence of phnenotypic polymorphism at convergence stable points $\ess{x}$ that aren't ESSs is called ``evolutionary branching'' since the polymorphism evolves as a dimorphism of phenotypes on either side of $\ess{x}$ in phenotype space \cite{Metz:Geritz:1996,Geritz:Kisdi:1998}.
+Studies of evolutionary branching have been important for theories of adaptive speciation \cite{Dieckmann:Doebeli:2004,Weissing:Edelaar:2011}, evolutionary rescue \cite{Ferriere:Legendre:2013}, and other areas of evolutionary ecology \cite{Dercole:Rinaldi:2008}.
+While evolutionary branching reveals the limits of long-term evolution since it can only predict long-term stable monomorphic phenotypes, the long-term process is still an important tool as it identifies where these branching points occur so that further analysis can study the evolution of the resulting phenotypic diversity.
 

### 2023-01-30 rev1_cover_letter.tex
just need final edits

@@
 I would like to thank the Editor and Reviewers for their very detailed in insightful comments.
-I have copied each Reviewer's comments below followed by a detailed response with changes I've made in response to the comments.
+I have copied each Reviewer's comments below (in blue) followed by a detailed response with changes I've made in response to the comments.
 
@@
 Linearity here is reasonable given the phenotype is a probability of selecting among two pure strategies.
+I provide more discussion of this assumption in the new section 8.
 
@@
 
-I have reworked those sentences and have added a list of the assumptions after Result 1 with some discussion of how they might be relaxed.
+I have reworked those sentences.
+I have also added a whole new section 8 that details all of the assumptions for Result 1 and discusses them in turn.
+See some of my additional responses below to Reviewer 2 for more detail.
 
@@
 
-Yes, I have largely left out reviewing adaptive dynamics, which is not directly intentional per se but more because I wanted to build around the ESS condition in the original 1973 paper as the Phil Trans issue commemorates 50 years since that paper.
+Yes, I had largely left out reviewing adaptive dynamics, which is not directly intentional per se but more because I wanted to build around the ESS condition in the original 1973 paper as the Phil Trans issue commemorates 50 years since that paper.
 However, adaptive dynamics is of course extremely important to evolutionary game theory and to the concept of invasion fitness.
-I've added XXX...
-%%% Add something about convergence stability to discussion.
+I've have added a new section 9 with a short discussion of adaptive dynamics and evolutionary branching.
 
 As to whether invasion fitness in adaptive dynamics is more general than what I discuss here, I would guess the reviewer is thinking of cases where short-term evolution reaches a non-equilibrium attractor like a cycle or chaotic attractor and invasion fitness must then really be measured in the sense of a Lyapunov exponent.
-I agree those cases are important but I do not mention them so as to keep the scope of the review from growing too much.
+I agree those cases are important but I do not mention them to limit the length of the review.
 It should be noted too that the assumptions and tools employed in typical deterministic and stochastic adaptive dynamics models ignore the non-equilibrium cases too as they are difficult to handle in general.
@@
 My review wasn't structured to synthesize the original ESS conditions with the stability and convergence conditions from adaptive dynamics but this would be the basis from a great review or synthesis article.
-%% Add something here maybe if I write convergence stability paragraph?
+However, I now briefly mention in section 9 how the second condition connects to convergence stability.
 
@@
 I think the reviewer may be referring to Maynard Smith (1978, Ann Rev Ecol Syst), Grafen (1979, Anim Behav), and Hines and Maynard Smith (1979, JTB).
+While this history is interesting, I haven't added a discussion of those papers since there isn't room in this paper for a proper analysis of the relevant issues.
 
@@
 I agree that I need to be clearer about the conditions necessary the results in sections 3 and 4 to hold.
-I've now added a bulleted list at the end of section 4 that lists all of the conditions and also discuss how restrictive they may or may not be.
+I've now added a new section 8 with a bulleted list that lists all of the conditions.
+The new section then discusses these assumptions in some detail.
 
@@
 
-Yes, this limitation is important. I now specify it clearly as the first assumption in the bulleted list at the end of section 4.
+Yes, this limitation is important. I now specify it clearly as the first assumption in the bulleted list at in the new section 8. I also mention the rock paper scissors example.
 
@@
 
-%% mention this more clearly at the end of section 4 and/or in the discussion
+I now cite Friedman in addition to the rock paper scissors model as examples where ESS may not imply stability.
 
@@
 
-%% ugh. ok take care of this
+This is very interesting and I would like to learn more but I could not locate where in the Karlin and Lessard book they clearly described these simulations and the reviewer does not note whether they have published their own simulations.
+Based on what I think is a general lack of published clear simulation tests of long-term evolution, I think that what the reviewer is describing is a fruitful area for future research that isn't mature enough for me to cover here.
 
@@
 
-I agree that section 6 is more technical. I have added XXX..
-%% add here
-
-I have tried to keep Section 7 to a minimum of technical depth.
+I agree that section 6 is more technical.
+Nevertheless, I hope that it still gives the reader a sense for how invasion fitness can be used in these situations.
+On the other hand, I have tried to keep Section 7 to a minimum of technical depth.
 I do not agree that its understandable to only technically sophisticated readers.
-However, it is relevant to add some of the assumptions for the modifier models discussed and I now do that in XXXX
+With respect to the assumptions that of the modifier models, they share a number in common with the broader long-term evolution approach and those assumptions are now discussed in detail in section 8.
 %% add here
@@
 
-I mention the reduction principle since it uses external stability to show the general evolutionary principle that selection in a constant generates purifying selection for the highest fitness genotype and against genetic variation created by other genotypes.
+I mention the reduction principle since it uses external stability to show the general evolutionary principle that selection in a constant environment generates purifying selection for the highest fitness genotype and against genetic variation created by other genotypes.
 Varying environments don't exhibit the reduction principle since genetic or phenotypic variation can be adaptive in these cases.

### 2023-01-30 manuscript.tex
submitted revision!

@@
 I give examples of how invasion fitness can measure kin selection, group selection, and the evolution of linked modifier genes in response to variable environments.
-The ESSs in these examples depend crucially on demographic and genetic parameters, which highlights how ESS analysis will continue to be an important tool in understanding evolutionary patterns even as models must address the increasing abundance of genetic and long-term demographic data in natural populations.
+The ESSs in these examples depend crucially on demographic and genetic parameters, which highlights how ESS analysis will continue to be an important tool in understanding evolutionary patterns as new models address the increasing abundance of genetic and long-term demographic data in natural populations.
 
@@
 Suppose that two individuals interact and the phenotypes that characterize their behavior in the interactions are measured with the variables $x$ and $y$.
-The phenotypes can be drawn from a set of discrete or continuous set of values that represents a single trait like body size or propensity to cooperate.
+The phenotypes can be drawn from a set of discrete or continuous values that represents a single trait like body size or propensity to cooperate.
 For the purpose of explaining the theory more concretely, assume that individuals can choose one of two possible behaviors in the interaction, $C$ and $D$.
@@
 Early mathematical analyses began by using the simplest population genetic model, which is one that assumes a \textit{haploid asexual} population with two genotypes.
-Assume the two genotypes produce the phenotypes $x=1$ that chooses $C$ with certainty and $x=0$ that chooses $D$ with certainty, respectively.
-The fitness of each of the four pairs can is given in he payoff matrix $W$ in \eqref{eq:payoff:matrix}.
+Assume the two genotypes correspond to the phenotypes $x=1$ that chooses $C$ with certainty and $x=0$ that chooses $D$ with certainty, respectively.
+The fitness of each of the four interaction pairs is given in he payoff matrix $W$ in \eqref{eq:payoff:matrix}.
 Suppose that $p$ measures the frequency of the phenotype $x=1$ that always chooses $C$.
 Pairs are assumed to form randomly according to the frequencies $p$ and $1-p$ of the two types.
-The equilibria of this model are either fixation of one of the two genotypes resulting in a population of individuals that always chooses $C$, $p=1$, or always choose $D$, $p=0$ or a polymorphism of both genotypes, $0 < p < 1$, resulting in a population where some choose $C$ and some choose $D$.
-The analogous ESS model of this scenario applies condition \eqref{eq:ess:def} to the four possible fitness values in the payoff matrix $W$ and determines which phenotype, $x=1$ or $x=0$, is ESS or whether a mixed strategy ($0 < x < 1$) is ESS.
+The equilibria of this model are either fixation of one of the two genotypes resulting in a population of individuals that always chooses $C$, $p=1$, or always chooses $D$, $p=0$ or a polymorphism of both genotypes, $0 < p < 1$, resulting in a population where some choose $C$ and some choose $D$.
+The analogous ESS model of this scenario applies condition \eqref{eq:ess:def} to the four possible fitness values in the payoff matrix $W$ in \eqref{eq:payoff:matrix} and determines which phenotype, $x=1$ or $x=0$, is ESS or whether a mixed strategy ($0 < x < 1$) is ESS.
 These early analyses showed that the values of $x$ that are an ESS are also values of $p$ that produce stable equilibria in the haploid asexual population genetic model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Hofbauer:Sigmund:1998}.
@@
 \end{equation*}
-If we assume that the fitness function $\w(x,y)$ is linear in the phenotypes $x$ and $y$ (which is a reasonable assumption given the phenotypes are probabilities of selection among behaviors $C$ and $D$), then the mean or marginal fitness of an individual with genotype $A_{i}B_{j}/A_{k}B_{l}$ interacting with other individuals in the population according to their frequencies is $w(x_{ijkl}, x)$.
+If we assume that the fitness function $\w(x,y)$ is linear in the phenotypes $x$ and $y$ (which is a reasonable assumption given the phenotypes are probabilities of selecting among behaviors $C$ and $D$), then the mean or marginal fitness of an individual with genotype $A_{i}B_{j}/A_{k}B_{l}$ interacting with other individuals in the population according to their frequencies is $w(x_{ijkl}, x)$.
 Likewise, the mean fitness in the population is then
@@
 Now, assume that a new mutant allele $A_{\mut}$ (increasing the total number of alleles at locus $A$ to $a+1$) arrives at low frequency in a population at an internally stable equilibrium $\fixp{\vec{p}}$ with \textit{resident} or wild-type mean phenotype $\fixp{x} = x(\fixp{\vec{p}})$.
-An individual with this new allele chooses phenotype $x$ with probability $x_{\mut jkl}$ if its genotype is $A_{\mut}B_{j}/A_{k}B_{l}$ and phenotype $y$ with probability $1 - x_{\mut jkl}$.
+An individual with this new allele chooses behavior $C$ with probability $x_{\mut jkl}$ if its genotype is $A_{\mut}B_{j}/A_{k}B_{l}$ and behavior $D$ with probability $1 - x_{\mut jkl}$.
 A linear stability analysis shows that this allele will either initially increase or decrease in frequency at a geometric rate $\eig$ that is the magnitude of the largest eigenvalue of a Jacobian matrix \cite{Edelstein-Keshet:2005}.
@@
 \end{equation}
-is the frequency of phenotype $x$ among mutant individuals once the invasion process has stabilized.
+is the average probability of choosing behavior $C$ among mutant individuals once the invasion process has stabilized.
 Note that since the mutant allele is rare, mutant individuals can have only one mutant chromosome; the other chromosome has no mutant alleles at locus $A$ and remains at the internal equilibrium frequency $\fixp{p}_{kl}$.
@@
 \end{equation*}
-for all possible mutant $A_{\mut}$ that generate mutant phenotype $x_{\mut}$.
+for all possible mutants $A_{\mut}$ that generate mutant phenotype $x_{\mut}$.
 This condition is in fact the condition for a strict ESS from \eqref{eq:ess:def:a} and can be used to prove that the following result for large, randomly mating, diploid populations with two-locus genetic systems \cite{Eshel:Feldman:1984,Eshel:1996,Eshel:Feldman:1998}:
@@
 Work by Liberman \cite{Liberman:1988} on more than two loci likely extends Result \ref{res:twolocus} to any finite number of loci.
-An important conclusion from Result \ref{res:twolocus} is that it shows that ESSs determined by the condition \eqref{eq:ess:def} of Maynard Smith and Price \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} are stable states of the long-term evolutionary process.
+Result \ref{res:twolocus} shows that ESSs determined by the condition \eqref{eq:ess:def} of Maynard Smith and Price \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} are stable states of the long-term evolutionary process.
 Assuming that long-term evolution is a reasonable model of phenotypic evolution, Result \ref{res:twolocus} demonstrates the importance of natural selection on phenotypes in determining the course of phenotypic evolution and provides a robust mathematical justification for the use of the ESS method in building phenotypic evolution models.
@@
 The magnitude of the eigenvalue $\eig$ measures the geometric growth rate (e.g., equation \eqref{eq:eig:twolocus}) of all mutant genotypes in a population of resident genotypes.
-If there are multiple possible genotypes with the mutant allele, as is the case for the two-locus model in section \ref{sec:long-term} where there may be many possible alleles at the non-mutant locus, then mutant genotype fractions during the invasion process are given by the normalized right eigenvector of the linear system $\vec{u}$.
-Thus, $\eig$ really measures the fitness of the whole mutant genetic lineage.
+If there are multiple possible genotypes with the mutant allele, as is the case for the two-locus model in section \ref{sec:long-term} where there may be many possible alleles at the non-mutant locus, then mutant genotype frequencies during the invasion process are given by the normalized right eigenvector of the linear system $\vec{u}$.
+Thus, $\eig$ really measures the fitness of the whole mutant genetic lineage that originates from the original invading mutants.
 The growth rate $\eig$ is often called the \textit{invasion fitness} \cite{Metz:Nisbet:1992,Heino:Metz:1998}\footnote{When invasion occurs at more complex attractors like cycles, the \textit{invasion exponent} is used, which is $\log |\eig|$ \cite{Rand:Wilson:1994}} and sometimes called the \textit{lineage fitness} \cite{Lehmann:Alger:2015,Akcay:VanCleve:2016,Lehmann:Mullon:2016,Lehmann:Rousset:2020}.
@@
   \label{eq:strict:relative}
-  \frac{w(y, \ess{x})}{w(\ess{x}, \ess{x})} = \omega(y, \ess{x}) < 1 \qquad \text{for all } y \ne x
+  \frac{w(y, \ess{x})}{w(\ess{x}, \ess{x})} = \omega(y, \ess{x}) < 1 \qquad \text{for all } y \ne \ess{x}
 \end{equation}
@@
     \cmidrule(lr){3-3}  \cmidrule(lr){4-4}  \cmidrule(lr){5-5} \cmidrule(lr){6-6}
-    \raisebox{-0.2cm}{\rotatebox{90}{Mutant switching rate}} & \raisebox{-0.2cm}{\yaxisandtickspip} &
+    \raisebox{-0.2cm}{\rotatebox{90}{Mutant mutation rate}} & \raisebox{-0.2cm}{\yaxisandtickspip} &
     \includegraphics[width=1.25in]{pip_n50_s1_0-01_s2_0-015_r_0-0.pdf} &
@@
     & & \xaxisandtickspip & \xaxisandtickspip & \xaxisandtickspip & \xaxisandtickspip \\
-    & & \multicolumn{4}{c}{Resident switching rate}
+    & & \multicolumn{4}{c}{Resident mutation rate}
   \end{tabular}
@@
 The fitness costs of being in the wrong environments are $s_{1} = 0.01$ and $s_{2} = 0.015$ and the recombination rate between the phenotypic locus and the modifier is given above each plot.
-White regions are where a mutant modifier allele with a switching rate given on the horizontal axis can invade a population fixed for a modifier allele with switching rate given on the vertical axis, and black regions are where the mutant cannot invade; in other words, the leading eigenvalue of external stability matrix $\mathcal{L}_{\mathrm{ex}}$ in equation (17) of Liberman et al.
+White regions are where a mutant modifier allele with a mutation rate given on the horizontal axis can invade a population fixed for a modifier allele with mutation rate given on the vertical axis, and black regions are where the mutant cannot invade; in other words, the leading eigenvalue of external stability matrix $\mathcal{L}_{\mathrm{ex}}$ in equation (17) of Liberman et al.
 \cite{Liberman:VanCleve:2011} is greater than one in the white regions and less than one in the black regions.}
@@
 In fact, environmental variation, particularly over time, is one of the most well-studied and important factors supporting the evolution of genotypic and phenotypic variation through mechanisms that promote such variation like recombination \cite{Charlesworth:1976,Sasaki:Iwasa:1987,Otto:Michalakis:1998}, mutation \cite{Leigh:1970,Ishii:Matsuda:1989,Lachmann:Jablonka:1996}, migration \cite{Gillespie:1981,McPeek:Holt:1992,Blanquart:Gandon:2011}, phenotypic plasticity \cite{Caswell:1983,Via:Lande:1985,Gavrilets:Scheiner:1993,Jong:1995} and bet-hedging \cite{Slatkin:1974,Seger:Brockmann:1987,Kussell:Leibler:2005,Salathe:VanCleve:2009}.
-Recent work shows that not only are the ES recombination, mutation, and migration evolve rates positive in variable environments, but they also depends on genetic parameters like the recombination rate between the major and modifier loci.
+Recent work shows that not only are the ES recombination, mutation, and migration rates positive in variable environments, but they also depends on genetic parameters like the recombination rate between the major and modifier loci.
 In the case of mutation rate evolution, \citeauthor{Liberman:VanCleve:2011} \cite{Liberman:VanCleve:2011} used an analysis of the external stability of the mutation rate to show that the ES mutation rate depends on the recombination rate between phenotypic locus and the modifier locus.
 For example, Figure \ref{fig:pip:switching}, shows pairwise invasibility plots for the mutation rate with different recombination rates and given weakly asymmetric selection in two different environments that alternate every $T=50$ generations.
-White regions are where mutant alleles with a switching rate on the vertical axis can invade a population with a resident rate given by the horizontal axis; black regions are when such a mutant cannot invade.
+White regions are where mutant alleles with a mutation rate on the vertical axis can invade a population with a resident rate given by the horizontal axis; black regions are when such a mutant cannot invade.
 The plots show that a rate of $~1/50$ is ES when the recombination rate is zero, similar to older results \cite{Leigh:1970,Lachmann:Jablonka:1996} that show the ES rate is $~1/T$.
-The ES switching rate decreases as recombination increases until there is an abrupt shift to an ES rate of zero switching.
+The ES mutation rate decreases as recombination increases until there is an abrupt shift to an ES rate of zero switching.
 \citeauthor{Carja:Liberman:2014} \cite{Carja:Liberman:2014} generalize this approach to the evolution of recombination, mutation, and migration in fluctuating environments.
@@
 Some of these of these assumptions are likely less restrictive than others.
-For example, assumption \ref{assume:ploidy} may not be very limiting as many organisms are haploid or diploid and ESS models of other genetic systems like haplodiploidy can be studied by adding demographic structure using matrix population models \cite{Taylor:1990,Taylor:Frank:1996,Frank:1998} like the one described below in section \ref{sec:IF:group-structure} on ESS and inclusive fitness.
+For example, assumption \ref{assume:ploidy} may not be very limiting as many organisms are haploid or diploid and ESS models of other genetic systems like haplodiploidy can be studied by adding demographic structure using matrix population models \cite{Taylor:1990,Taylor:Frank:1996,Frank:1998} like the one described in section \ref{sec:IF:group-structure} on ESS and inclusive fitness.
 The large population size assumption \ref{assume:large-pop} also can be relaxed by using stochastic population models that allow for genetic drift and and use fixation probability as a fitness measure instead of invasion fitness \cite{Rousset:Billiard:2000,Rousset:2003,Nowak:Sasaki:2004,Lessard:2005,VanCleve:2015}.
@@
 Assumption \ref{assume:twostrat} that there are two behaviors or strategies is particularly important.
-In practice, games like rock paper scissors \cite{Taylor:Jonker:1978,Zeeman:1980,Weissing:1991,Hofbauer:Sigmund:1998} involving three or more strategies may have more complex long-term evolutionary dynamics resulting in equilibria that may or may not be ESSs.
+In practice, games like rock paper scissors \cite{Taylor:Jonker:1978,Zeeman:1980,Weissing:1991,Hofbauer:Sigmund:1998} involving three or more strategies may have more complex evolutionary dynamics resulting in equilibria that may or may not be ESSs.
 Support for this possibility comes from examples of three strategy interactions like rock paper scissors where the ESS strategy is actually an internally \textit{unstable} equilibrium \cites{Friedman:1991,Weissing:1991}.
@@
 
-The linear fitness function assumption \ref{assume:linear} is also important restriction.
+The linear fitness function assumption \ref{assume:linear} is also an important restriction.
 It arises naturally from assuming that the fitness of an individual using a mixed strategy phenotype is the expected fitness that individual would obtain using pure strategies with probabilities given by the mixed strategy.
-Linear fitness functions are convenient in part because they simplify interactions involving more than two individuals: fitness in such interactions can be expressed as a linear combination of the fitness of pairwise interactions.
+Linear fitness functions are convenient in part because they simplify interactions involving more than two individuals: fitness in such interactions can be expressed as a linear combination of the fitnesses of pairwise interactions.
 This means the pairwise interaction assumption \ref{assume:pairwise} is unnecessary for linear fitness functions.
-However, phenotypes often involve continuous quantities like body size or metabolic rate instead of pure strategy probabilities and nonlinear fitness functions are the norm in these cases.
+However, phenotypes often involve continuous quantities like body size or metabolic rate and nonlinear fitness functions are the norm in these cases.
 Studying nonlinear fitness functions, Eshel, Taylor, and others \cite{Eshel:Motro:1981,Eshel:1983,Taylor:1989,Christiansen:1991,Metz:Geritz:1996,Eshel:Motro:1997,Geritz:Kisdi:1998} showed that evolutionary stability involves two qualitatively different kinds of stability.
@@
 
-Convergence stability refers to populations with mean phenotype $x$ near an ESS $\ess{x}$ that are invaded by mutants that bring the population closer to $\ess{x}$ and resist invasion by mutants that move the population farther from $\ess{x}$ \cite{Eshel:1983,Taylor:1989}.
+Convergence stability refers to populations with mean phenotype $x$ near an ESS $\ess{x}$ that are invaded by mutants that bring the population closer to $\ess{x}$ and that resist invasion by mutants that move the population farther from $\ess{x}$ \cite{Eshel:1983,Taylor:1989}.
 Thus, populations near a convergence stable phenotype $\ess{x}$ will evolve towards $\ess{x}$.
@@
 
-Given there are two kinds of stability to asses, evolutionary stability \textit{sensu} condition \eqref{eq:ess:def:a} and convergence stability, a phenotype $\ess{x}$ that is not long-term stable might fail either one of them.
+Given there are two kinds of stability to assess, evolutionary stability \textit{sensu} condition \eqref{eq:ess:def:a} and convergence stability, a phenotype $\ess{x}$ that is not long-term stable might fail either one of them.
 If $\ess{x}$ is not convergence stable, then a population with mean phenotype $\ess{x}$ will resist invasion by mutants with alternative phenotypes but populations with mean phenotypes near $\ess{x}$ will eventually evolve away from $\ess{x}$.
@@
 Studies of evolutionary branching have been important for theories of adaptive speciation \cite{Dieckmann:Doebeli:2004,Weissing:Edelaar:2011}, evolutionary rescue \cite{Ferriere:Legendre:2013}, and other areas of evolutionary ecology \cite{Dercole:Rinaldi:2008}.
-While evolutionary branching reveals the limits of long-term evolution since it can only predict long-term stable monomorphic phenotypes, the long-term process is still an important tool as it identifies where these branching points occur so that further analysis can study the evolution of the resulting phenotypic diversity.
+While evolutionary branching reveals the limits of long-term evolution since long-term evolution can only predict stable monomorphic phenotypes, the long-term process is still an important tool as it identifies where these branching points occur so that further analysis can study the evolution of the resulting phenotypic diversity.
 
@@
 Put another way, how justifiable is the phenotypic gambit?
-Result \ref{res:twolocus} shows that for large and randomly mating populations, ESSs can be fairly independent of genetic parameters like mutation and recombination rates and even the number of loci (for mixed ESSs; see \cite{Eshel:Feldman:1998,Eshel:Feldman:2001}).
+Result \ref{res:twolocus} shows that for large and randomly mating populations, ESSs can be fairly independent of genetic parameters like mutation and recombination rates and even the number of loci.
 At the time, this independence was important for supporting the use of ESS analysis across a broad range of species with different genetics and demography.
